---
title: "03-BirdNET-Analysis"
author: "Kelly Faller"
date: "2025-06-02"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    number_sections: true
    highlight: tango
    df_print: paged
    code_folding: hide
params:
  output_dir: "Outputs/BirdNET/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# BirdNET Analysis Appendix
This appendix contains all analyses and visualizations related to BirdNET—a tool for automated bird sound detection and classification. The BirdNET workflow was separated from the main `02-Acoustics-Analysis.Rmd` file for clarity and reproducibility.

**Why is BirdNET separated?**
- BirdNET is used only occasionally in this project and doesn’t form the core of the daily acoustic analysis.
- Its results and plots are valuable, but they can clutter the main analysis if included directly.
- Having a separate appendix allows for a modular, reproducible workflow—users who don’t need BirdNET analysis can skip this file entirely.

**How to use this appendix:**
1. Open this file (`03-BirdNET-Analysis.Rmd`) in RStudio or an R environment.
2. Adjust any file paths or data-specific parameters as needed.
3. Knit the document to generate a standalone HTML report of BirdNET results. Will be saved in \Outputs\BirdNET

**Important:** This appendix is fully independent of the main `02-Acoustics-Analysis.Rmd` file. If you update your dataset or processing steps, ensure that any BirdNET-specific inputs are updated here as well.

---
# Methods
INCLUDE METHODS HERE!!!!!!!!

## Data Handling

Clear Environment

```{r, warning=FALSE, message=FALSE}
rm(list=ls())
```

Load Packages

```{r, warning=FALSE, message=FALSE}
# Data manipulation and cleaning
library(tidyverse)
library(dplyr)
library(tidyr)
library(lubridate)

# BirdNET-specific tools
# library(birdnetR)

# Visualization
library(ggplot2)
library(ggpubr)
library(viridis)
```

Data Loading
```{r Loading BirdNet, include=FALSE}
species_data_birdnet<- read.csv("../Species Id Data.csv", header=TRUE, sep = ",", stringsAsFactors=FALSE)
species_list_birdnet<- read.csv("../Data/Species List.csv", header=TRUE, sep = ",", stringsAsFactors=FALSE)

# lubridate formatting
species_data_birdnet$Date <- mdy(species_data_birdnet$Date)
species_data_birdnet$DateTime <- mdy_hm(species_data_birdnet$DateTime)

#connecting species data and species list
species_data_birdnet <- species_data_birdnet %>%
   left_join(dplyr::select(species_list_birdnet, c(Scientific_Name,Lower_Guild, Common_Birds, Guild)), by = 'Scientific_Name')
```
# BirdNet Species Summaries

These data are coming from Cornell's BirdNET Analyzer. It is essentially a GUI version of their popular Merlin app. It requires a lot of decisions on how to analyze your recordings. The user selects confidence levels, sensitiviy, and overlap. The data crunched below is the first deployment (n=10) and is using very lax confidence levels. It will require manual verification (especially for rare species) and the accuracy should be tested. However, we are currently in the stage of seeing how we can best visualize the data. In the near future, we will be testing difference methods to get the most accurate representation of species presence and community composition. 

The summary below counts the amount of unique species identified by BirdNET at each site. These data will be verified to eliminate false detections. The software does a good job of identifying common birds but will occasionally include a few ecologically unreasonable detections. These generally account for very few detections (i.e. <0.01% of all detections), so the numbers appear more inflated in the summaries.
```{r Species Summary, echo=TRUE}

n_distinct(species_data_birdnet$Scientific_Name)
setDT(species_data_birdnet)[, .(count = uniqueN(Scientific_Name)), by = Station]
setDT(species_data_birdnet)[, .(count = uniqueN(Scientific_Name)), by = Site]
guild_count<- setDT(species_data_birdnet)[, .(count = uniqueN(Scientific_Name)), by = Guild]
species_breakdown <- species_data_birdnet %>%
    group_by(Station, Scientific_Name) %>%
    summarize(Count = n()) %>%
    mutate(Percent = round(Count / sum(Count) * 100, 2))

species_over_1p<- species_breakdown %>% 
  mutate(Percent_1p = case_when(
    Percent > 0.01~ "Above",
    Percent  < 0.01 ~ "Below",
    Percent == 0.01 ~ "At")
    %>% as.factor())

species_data_birdnet2 <- species_data_birdnet %>% 
  mutate(Site_Type = case_when(
    Station == "RF_01" ~ "Reference",
    Station == "RF_02" ~ "Reference",
    Station == "GAR_01" ~ "Wild Card",
    Station == "GAR_02" ~ "Restoration",
    Station == "GAR_03" ~ "Wild Card",
    Station == "GAR_04" ~ "Restoration")
    %>% as.factor())

RF_01_spps_breakdown <- species_data_birdnet %>%
    filter(Station == "RF_01") %>%
    group_by(Scientific_Name) %>%
    summarize(Count = n()) %>%
    mutate(Percent = round(Count / sum(Count) * 100, 2))
RF_02_spps_breakdown <- species_data_birdnet %>%
    filter(Station == "RF_02") %>%
    group_by(Scientific_Name) %>%
    summarize(Count = n()) %>%
    mutate(Percent = round(Count / sum(Count) * 100, 2))
GAR_01_spps_breakdown <- species_data_birdnet %>%
    filter(Station == "GAR_01") %>%
    group_by(Scientific_Name) %>%
    summarize(Count = n()) %>%
    mutate(Percent = round(Count / sum(Count) * 100, 2))
GAR_02_spps_breakdown <- species_data_birdnet %>%
    filter(Station == "GAR_02") %>%
    group_by(Scientific_Name) %>%
    summarize(Count = n()) %>%
    mutate(Percent = round(Count / sum(Count) * 100, 2))
```

# BirdNet Species Outputs

These data are coming from Cornell's BirdNET Analyzer. It is essentially a GUI version of their popular Merlin app. It requires a lot of decisions on how to analyze your recordings. The user selects confidence levels, sensitiviy, and overlap. The data crunched below is the first deployment (n=10) and is using very lax confidence levels. It will require manual verification (especially for rare species) and the accuracy should be tested. However, we are currently in the stage of seeing how we can best visualize the data. In the near future, we will be testing difference methods to get the most accurate representation of species presence and community composition. 

Data is visualized in two ways, common species and guilds. Guilds gives a good overview of how sites vary (i.e. more shorebirds and ducks in a mudflat). Below the graphics will be referencing the same dataset, but either calling out specific common bird species or guild. 

## Species Diversity

*Currently the species graphics are only for the first deployments of 2023. Running BirdNET requires a lot of time and processing power, so we have only completed one round of processing.Will need to update this if it is ever worth the effort*

```{r Species Diversity1, echo=FALSE,fig.height=12, fig.width=18}

#| fig.cap= "Appendix Figure 1: Guild breakdown for all sites."

ggplot(data= species_data_birdnet, aes(x= DateTime, fill= Lower_Guild), na.rm=T)+ 
         geom_histogram()+ coord_polar()+
         theme_bw(base_size=15)+
         ggtitle("All Station Guild Breakdown")+
         labs(x= "DateTime", y= "Guild")
```

```{r Species Diversity2, echo=FALSE,fig.height=12, fig.width=18}

#| fig.cap= "Appendix Figure 2: Breakdown of common bird species by site during the first deployment (4/26/23 to 5/7/23). Four-letter alpha codes represent common name of bird species. Seaside sparrow (SESP) is by far the most dominant species with 380,311 total detections."

ggplot(species_data_birdnet, aes(x= Common_Birds,  group= Station)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
    geom_text(aes( label = scales::percent(..prop..,accuracy=0.01),
                   y= ..prop.. ), stat= "count", vjust =- 1.5, size= 3) +
    labs(y = "Percent", fill="Common_Birds") + 
  guides(fill= FALSE)+
         theme_bw(base_size=15)+
    facet_grid(~Station) +
    scale_y_continuous(labels = scales::percent)+ 
         theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r Species Diversity3, echo=FALSE,fig.height=12, fig.width=18}

#| fig.cap= "Appendix Figure 3: Guild breakdown by site during the first deployment (4/26/23 to 5/7/23). Passerines were the most common guild represented in the data due to the high level of SESP detections."

ggplot(species_data_birdnet, aes(x= Lower_Guild,  group= Station)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
    geom_text(aes( label = scales::percent(..prop..,accuracy=0.1),
                   y= ..prop.. ), stat= "count", vjust =- 1.5, size= 2.5) +
    labs(x= "Guild", y = "Percent", fill="Lower_Guild") + 
  guides(fill= FALSE)+
         theme_bw(base_size=18)+
    facet_grid(~Station) +
    scale_y_continuous(labels = scales::percent)+ 
         theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```


```{r Species Diversity4, echo=TRUE,fig.height=20, fig.width=15}

# Set up for graphics

Restore_common <- ggplot(species_data_birdnet2 %>% filter(Site_Type == "Restoration"), aes(x= Common_Birds,  group= Station)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
    geom_text(aes( label = scales::percent(..prop..,accuracy=0.01),
                   y= ..prop.. ), stat= "count", vjust =- 1.5, size= 3) +
    labs(x= " ", y = "Percent", fill="Common_Birds") + 
  guides(fill= FALSE)+
         theme_bw(base_size=15)+
    facet_grid(~Station) +
  labs(subtitle = "Restoration Sites") +
    scale_y_continuous(labels = scales::percent)+ 
         theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

Refer_common <- ggplot(species_data_birdnet2 %>% filter(Site_Type == "Reference"), aes(x= Common_Birds,  group= Station)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
    geom_text(aes( label = scales::percent(..prop..,accuracy=0.1),
                   y= ..prop.. ), stat= "count", vjust =- 1.5, size= 2.5) +
    labs(x= " ", y = "Percent", fill="Lower_Guild") + 
  guides(fill= FALSE)+
         theme_bw(base_size=18)+
    facet_grid(~Station) +
  labs(subtitle = "Reference Sites") +
    scale_y_continuous(labels = scales::percent)+ 
         theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

Wild_common <- ggplot(species_data_birdnet2 %>% filter(Site_Type == "Wild Card"), aes(x= Common_Birds,  group= Station)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
    geom_text(aes( label = scales::percent(..prop..,accuracy=0.1),
                   y= ..prop.. ), stat= "count", vjust =- 1.5, size= 2.5) +
    labs(x= "Common Bird Species Alpha Code", y = "Percent", fill="Lower_Guild") + 
  guides(fill= FALSE)+
         theme_bw(base_size=18)+
    facet_grid(~Station) +
  labs(subtitle = "Wild Card Sites") +
    scale_y_continuous(labels = scales::percent)+ 
         theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

# By Guild

Restore_guild <- ggplot(species_data_birdnet2 %>% filter(Site_Type == "Restoration"), aes(x= Lower_Guild,  group= Station)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
    geom_text(aes( label = scales::percent(..prop..,accuracy=0.01),
                   y= ..prop.. ), stat= "count", vjust =- 1.5, size= 3) +
    labs(x= " ", y = "Percent", fill="Common_Birds") + 
  guides(fill= FALSE)+
         theme_bw(base_size=15)+
    facet_grid(~Station) +
  labs(subtitle = "Restoration Sites") +
    scale_y_continuous(labels = scales::percent)+ 
         theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

Refer_guild <- ggplot(species_data_birdnet2 %>% filter(Site_Type == "Reference"), aes(x= Lower_Guild,  group= Station)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
    geom_text(aes( label = scales::percent(..prop..,accuracy=0.1),
                   y= ..prop.. ), stat= "count", vjust =- 1.5, size= 2.5) +
    labs(x= " ", y = "Percent", fill="Lower_Guild") + 
  guides(fill= FALSE)+
         theme_bw(base_size=18)+
    facet_grid(~Station) +
  labs(subtitle = "Reference Sites") +
    scale_y_continuous(labels = scales::percent)+ 
         theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

Wild_guild <- ggplot(species_data_birdnet2 %>% filter(Site_Type == "Wild Card"), aes(x= Lower_Guild,  group= Station)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
    geom_text(aes( label = scales::percent(..prop..,accuracy=0.1),
                   y= ..prop.. ), stat= "count", vjust =- 1.5, size= 2.5) +
    labs(x= "Guild", y = "Percent", fill="Lower_Guild") + 
  guides(fill= FALSE)+
         theme_bw(base_size=18)+
    facet_grid(~Station) +
  labs(subtitle = "Wild Card Sites") +
    scale_y_continuous(labels = scales::percent)+ 
         theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

title1  = "Common Bird Species By Site Type"
title2  = "Bird Guild By Site Type"

#| fig.cap= "Appendix Figure 4: Common bird species breakdown by site type."
plot_grid(Restore_common, 
         Refer_common,
         Wild_common,
         labels = title1, label_size = 14, hjust = -2,
          nrow= 3, align = "v", axis="l")
```

```{r Species Diversity5, echo=TRUE,fig.height=20, fig.width=15}
#| fig.cap= "Appendix Figure 5: Bird guild breakdown by site type (Restoration, Reference, Wild Card) during the first deployment period."

plot_grid(Restore_guild, 
         Refer_guild,
         Wild_guild,
         labels = title2, label_size = 14, hjust = -3,
          nrow= 3, align = "v", axis="l")
```

## Summary Tables
Gives further context to the data, but in a tabular form rather than graphically. 
Broken down by site, station, acoustic indices, site types, and condfidence levels.

```{r Summaries, echo=TRUE}

st(index_data %>% dplyr::select(c("SITE", "STATION", "NDSI", "ACI", "ADI" , "AEI", "BI")))
st(avg_index2 %>% dplyr::select(c("SITE", "STATION", "Site_Type", "avg_NDSI", "avg_ACI", "avg_ADI" , "avg_AEI", "avg_BI")))#,
                                #"sd_NDSI","sd_ACI", "sd_ADI" , "sd_AEI", "sd_BI")))
st(species_data_birdnet2 %>% dplyr::select(c("Site", "Station", "Confidence", "Site_Type")))
st(species_data_birdnet %>% dplyr::select(c("Site", "Station", "Confidence", "Common_Birds")), group= 'Common_Birds', group.long = TRUE)
st(species_breakdown)

```
<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>