---
title: "Acoustic Calculation Script"
author: "Kelly Faller (adapted)"
date: "2025-06-09"
output: 
  html_document:
    toc: true
    number_sections: true
    toc_float: true
    theme: flatly
    code_folding: hide
params:
  write_combined: true
---
# Purpose
This code is meant to generate acoustic index values and conduct bioacoustic calculations in the absence of Kaliedoscope. Kaliedoscope (Wildlife Acoustics) is fantastic and can run these calculations very quickly. However, the software comes at an annual rate of $399. 

This code is coming from Donovan Tye's "acoustic indices.R" script. I've amended the end of the script to create one dataset for queries.

```{r setup, include= FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# ==== USER SETTINGS - change these for each project ====
main_dir <- "D:/Acoustic Data/Masters - Cattus and LHC"  # top-level directory
output_dir <- "C:/Users/kelly/Documents/GitHub/Soundscape-Analysis-Rare-Bird-Detection/Outputs/Acoustic_Index" #where you want the outputs to go
cores <- 10 # how much computational power you need
min_freq   <- 0
max_freq   <- 22000
db_thresh  <- -35

# Create output directory if missing
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(glue)
  library(tuneR)
  library(seewave)
  library(soundecology)
})

# helper: only keep folders that actually contain WAVs
has_wavs <- function(path) any(grepl("(?i)\\.wav$", list.files(path, full.names = FALSE)))
subfolders <- list.dirs(main_dir, recursive = TRUE, full.names = TRUE)
subfolders <- subfolders[vapply(subfolders, has_wavs, logical(1))]
stopifnot(length(subfolders) > 0)

```

```{r Run Indices}
run_indices <- function(wav_dir, out_prefix) {
  message("Processing: ", out_prefix, " @ ", wav_dir, 
          " | band ", min_freq, "–", max_freq, " Hz")

  # ACI
  multiple_sounds(
    directory  = wav_dir,
    resultfile = file.path(output_dir, paste0(out_prefix, "_ACI_F.csv")),
    soundindex = "acoustic_complexity",
    min_freq   = min_freq, 
    max_freq   = max_freq, 
    no_cores   = cores
  )

  # ADI
  multiple_sounds(
    directory  = wav_dir,
    resultfile = file.path(output_dir, paste0(out_prefix, "_ADI_F_35db.csv")),
    soundindex = "acoustic_diversity",
    db_threshold = db_thresh,
    no_cores   = cores
  )

  # AEI
  multiple_sounds(
    directory  = wav_dir,
    resultfile = file.path(output_dir, paste0(out_prefix, "_AEI_F.csv")),
    soundindex = "acoustic_evenness",
    no_cores   = cores
  )

  # BI
  multiple_sounds(
    directory  = wav_dir,
    resultfile = file.path(output_dir, paste0(out_prefix, "_BI_F.csv")),
    soundindex = "bioacoustic_index",
    min_freq   = min_freq, 
    max_freq   = max_freq, 
    no_cores   = cores
  )

  # NDSI (defaults OK; leave band out unless you want to constrain)
  multiple_sounds(
    directory  = wav_dir,
    resultfile = file.path(output_dir, paste0(out_prefix, "_NDSI_F.csv")),
    soundindex = "ndsi",
    no_cores   = cores
  )

  # Acoustic Entropy (H)
  multiple_sounds(
    directory  = wav_dir,
    resultfile = file.path(output_dir, paste0(out_prefix, "_H_F.csv")),
    soundindex = "H",
    no_cores   = cores
  )
}

for (folder in subfolders) {
  run_indices(folder, basename(folder))
}
```

Create one large dataset for queries. 

```{r Combine, eval = params$write_combined}
`%||%` <- function(x, y) if (is.null(x)) y else x

extract_metadata <- function(df, csv_path) {
  site_pat <- "(?i)[A-Z]{2,5}_[A-Z]{0,3}\\d{1,3}"  # e.g., CAT_A1, LHC_01
  ts_str   <- stringr::str_extract(df$FILENAME %||% "", "\\d{8}_\\d{6}")
  site_fn  <- stringr::str_extract(df$FILENAME %||% "", site_pat)
  site_csv <- toupper(sub("_(ACI|ADI|AEI|BI|NDSI|H).*", "", basename(csv_path)))

  # normalize DB_THRESHOLD if present
  has_db <- names(df)[tolower(names(df)) == "db_threshold"]
  if (length(has_db) == 1 && !"DB_THRESHOLD" %in% names(df)) {
    df$DB_THRESHOLD <- as.character(df[[has_db]])
  } else if (!"DB_THRESHOLD" %in% names(df)) {
    df$DB_THRESHOLD <- NA_character_
  }

  df %>%
    dplyr::mutate(
      site_name = dplyr::coalesce(site_fn, site_csv),
      recording_datetime = suppressWarnings(
        lubridate::ymd_hms(stringr::str_replace(ts_str, "_", " "), tz = "UTC", quiet = TRUE)
      )
    )
}

all_csvs <- list.files(output_dir, pattern = "_(ACI|ADI|AEI|BI|NDSI|H)_F\\.csv$", full.names = TRUE)

if (length(all_csvs)) {
  ResultsData <- purrr::map_dfr(all_csvs, ~{
    df <- suppressMessages(readr::read_csv(.x, show_col_types = FALSE))
    extract_metadata(df, .x)
  })
  out_file <- file.path(output_dir, "all_indices_combined.csv")
  readr::write_csv(ResultsData, out_file)
  message("Combined file: ", out_file)
} else {
  warning("No per-index CSVs found to combine in: ", output_dir)
}
```
