---
title: "Acoustic Calculation Script"
author: "Kelly Faller (adapted)"
date: "2025-06-09"
output: 
  html_document:
    toc: true
    number_sections: true
    toc_float: true
    theme: flatly
    code_folding: hide
params:
  # Top-level folder of .wav files. Searched recursively; every folder that contains wavs is processed separately.
  # Use a year/station layout (e.g. 2024/CAT_A1) - outputs are named from the path so years don't overwrite each other.
  main_dir: "D:/Acoustic Data/Masters - Cattus and LHC"
  output_dir: "outputs/indices"          # relative = inside this repo (configs/project.yml outputs$indices_dir)
  overwrite: false                       # false = skip folders that already have all six index files (lets a long run resume)
  write_combined: true
---
# Purpose
This code is meant to generate acoustic index values and conduct bioacoustic calculations in the absence of Kaliedoscope. Kaliedoscope (Wildlife Acoustics) is fantastic and can run these calculations very quickly. However, the software comes at an annual rate of $399. 

This code is coming from Donovan Tye's "acoustic indices.R" script. I've amended the end of the script to create one dataset for queries.

```{r setup, include= FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# ==== USER SETTINGS ====
# Folders are set in the YAML header above; frequency band, dB threshold and cores come from configs/project.yml
# (params: min_freq_hz, max_freq_hz, db_threshold, cores) so every script uses the same settings.
main_dir   <- params$main_dir
output_dir <- if (fs::is_absolute_path(params$output_dir)) params$output_dir else here::here(params$output_dir)
proj_cfg   <- yaml::read_yaml(here::here("configs", "project.yml"))$params
min_freq   <- proj_cfg$min_freq_hz
max_freq   <- proj_cfg$max_freq_hz   # must not be above half the sample rate (16000 Hz for 32 kHz recordings)
db_thresh  <- proj_cfg$db_threshold
cores      <- min(proj_cfg$cores, max(1, parallel::detectCores() - 1))
tz         <- "America/New_York"     # recorder file names are local time

# Create output directory if missing
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(glue)
  library(tuneR)
  library(seewave)
  library(soundecology)
})

# helper: only keep folders that actually contain WAVs
has_wavs <- function(path) any(grepl("(?i)\\.wav$", list.files(path, full.names = FALSE)))
if (!dir.exists(main_dir)) stop("main_dir doesn't exist: ", main_dir, "\nChange main_dir in the YAML header.")
subfolders <- list.dirs(main_dir, recursive = TRUE, full.names = TRUE)
subfolders <- subfolders[vapply(subfolders, has_wavs, logical(1))]
stopifnot(length(subfolders) > 0)

# output prefix from the folder path below main_dir, e.g. "2024/CAT_A1" -> "2024_CAT_A1".
# (Using only the last folder name made 2025/CAT_A1 overwrite 2024/CAT_A1.)
folder_prefix <- function(folder) {
  base <- normalizePath(main_dir, winslash = "/")
  f    <- normalizePath(folder, winslash = "/")
  rel  <- if (startsWith(f, base) && nchar(f) > nchar(base)) substring(f, nchar(base) + 2) else basename(folder)
  gsub("[/ ]+", "_", rel)
}

# the sample rate of the first wav in each folder; the upper frequency can't be above half of it
check_band <- function(folder) {
  wav <- list.files(folder, pattern = "(?i)\\.wav$", full.names = TRUE)[1]
  sr <- tuneR::readWave(wav, header = TRUE)$sample.rate
  if (max_freq > sr / 2) stop("max_freq_hz (", max_freq, ") is above half the sample rate (", sr / 2, " Hz) in ", folder,
                              ". Lower max_freq_hz in configs/project.yml.")
}

```

```{r Run Indices}
run_indices <- function(wav_dir, out_prefix) {
  message("Processing: ", out_prefix, " @ ", wav_dir, 
          " | band ", min_freq, "–", max_freq, " Hz")

  # ACI
  multiple_sounds(
    directory  = wav_dir,
    resultfile = file.path(output_dir, paste0(out_prefix, "_ACI_F.csv")),
    soundindex = "acoustic_complexity",
    min_freq   = min_freq, 
    max_freq   = max_freq, 
    no_cores   = cores
  )

  # ADI
  multiple_sounds(
    directory  = wav_dir,
    resultfile = file.path(output_dir, paste0(out_prefix, "_ADI_F_35db.csv")),
    soundindex = "acoustic_diversity",
    db_threshold = db_thresh,
    no_cores   = cores
  )

  # AEI
  multiple_sounds(
    directory  = wav_dir,
    resultfile = file.path(output_dir, paste0(out_prefix, "_AEI_F.csv")),
    soundindex = "acoustic_evenness",
    no_cores   = cores
  )

  # BI
  multiple_sounds(
    directory  = wav_dir,
    resultfile = file.path(output_dir, paste0(out_prefix, "_BI_F.csv")),
    soundindex = "bioacoustic_index",
    min_freq   = min_freq, 
    max_freq   = max_freq, 
    no_cores   = cores
  )

  # NDSI (defaults OK; leave band out unless you want to constrain)
  multiple_sounds(
    directory  = wav_dir,
    resultfile = file.path(output_dir, paste0(out_prefix, "_NDSI_F.csv")),
    soundindex = "ndsi",
    no_cores   = cores
  )

  # Acoustic Entropy (H)
  multiple_sounds(
    directory  = wav_dir,
    resultfile = file.path(output_dir, paste0(out_prefix, "_H_F.csv")),
    soundindex = "H",
    no_cores   = cores
  )
}

index_suffixes <- c("_ACI_F.csv", "_ADI_F_35db.csv", "_AEI_F.csv", "_BI_F.csv", "_NDSI_F.csv", "_H_F.csv")
for (folder in subfolders) {
  prefix <- folder_prefix(folder)
  done <- all(file.exists(file.path(output_dir, paste0(prefix, index_suffixes))))
  if (done && !isTRUE(params$overwrite)) { message("Skipping ", prefix, " (already done)"); next }
  check_band(folder)
  run_indices(folder, prefix)
}
```

Create one large dataset for queries. 

```{r Combine, eval = params$write_combined}
`%||%` <- function(x, y) if (is.null(x)) y else x

extract_metadata <- function(df, csv_path) {
  site_pat <- "(?i)[A-Z]{2,5}_[A-Z]{0,3}\\d{1,3}"  # e.g., CAT_A1, LHC_01
  ts_str   <- stringr::str_extract(df$FILENAME %||% "", "\\d{8}_\\d{6}")
  site_fn  <- stringr::str_extract(df$FILENAME %||% "", site_pat)
  # output files are named <year>_<station>_<INDEX>...; station = everything after a leading year
  prefix   <- sub("_(ACI|ADI|AEI|BI|NDSI|H)_F.*", "", basename(csv_path))
  site_csv <- toupper(sub("^\\d{4}_", "", prefix))
  year_csv <- suppressWarnings(as.integer(stringr::str_extract(prefix, "^\\d{4}")))

  # normalize DB_THRESHOLD if present
  has_db <- names(df)[tolower(names(df)) == "db_threshold"]
  if (length(has_db) == 1 && !"DB_THRESHOLD" %in% names(df)) {
    df$DB_THRESHOLD <- as.character(df[[has_db]])
  } else if (!"DB_THRESHOLD" %in% names(df)) {
    df$DB_THRESHOLD <- NA_character_
  }

  df %>%
    dplyr::mutate(
      site_name = dplyr::coalesce(site_fn, site_csv),
      # local recorder time (the file names are local, not UTC)
      recording_datetime = suppressWarnings(
        lubridate::ymd_hms(stringr::str_replace(ts_str, "_", " "), tz = tz, quiet = TRUE)
      ),
      year = dplyr::coalesce(lubridate::year(recording_datetime), year_csv),
      source_csv = basename(csv_path)
    )
}

# includes the ADI files, which are named _ADI_F_35db.csv (the old pattern only matched _F.csv and left ADI out)
all_csvs <- list.files(output_dir, pattern = "_(ACI|ADI|AEI|BI|NDSI|H)_F(_\\d+db)?\\.csv$", full.names = TRUE)

if (length(all_csvs)) {
  ResultsData <- purrr::map_dfr(all_csvs, ~{
    # read as text: the index files disagree on column types (e.g. DB_THRESHOLD is a number in ADI, "-" elsewhere)
    df <- suppressMessages(readr::read_csv(.x, col_types = readr::cols(.default = "c")))
    extract_metadata(df, .x)
  }) %>% readr::type_convert(col_types = readr::cols(DB_THRESHOLD = "c", .default = readr::col_guess())) %>%
    # soundecology writes each recording twice in the AEI file (identical values); keep one row per file and index
    dplyr::distinct(source_csv, FILENAME, INDEX, .keep_all = TRUE)
  out_file <- file.path(output_dir, "all_indices_combined.csv")
  readr::write_csv(ResultsData %>% dplyr::mutate(recording_datetime = format(recording_datetime, "%Y-%m-%d %H:%M:%S")), out_file)
  message("Combined file: ", out_file)
} else {
  warning("No per-index CSVs found to combine in: ", output_dir)
}
```
