---
title: "Arbimon Data Prep Script"
author: "Kelly Faller"
date: "2024-10-11"
output: html_document
# >>> Edit only these params. These will change when you run this file, but the rest should be automated <<<
params:
  project_dir: "C:/Users/kelly/Documents/Master's Project Files/arbimon_backups/salt-marsh-restoration-soundscape-project_2025-08-12_export" #change this !!!
   # File name patterns (do not change unless your export uses different names)
  recordings_pattern: "^recordings\\."
  sites_pattern: "^sites\\."
  pm_pattern: "^pattern_matching_rois\\."
  species_pattern: "^species\\."
  validations_pattern: "^recording_validations\\."  # manual labels - not from pm analysis
  # Join and time parsing
  join_by_recording: ["site_id"]
  join_by_metadata:  ["site_id"]
  datetime_candidates: ["datetime", "timestamp", "date_time"]   # columns to try
  datetime_orders: ["Y-m-d H:M:S", "m/d/Y H:M:S", "Ymd HMS", "Ymd_HMS", "Y-m-d H:M:OS", "Ymd HMSOS"]
  tz: "UTC" 
  # Optional: keep only these columns from recordings after join (NULL = keep all). Also use validations or not.
  keep_columns: null
  use_validations: true                        # set to false to ignore manual labels
  # Outputs
  detections_file: "C:/Users/kelly/Documents/GitHub/Soundscape-Analysis-Rare-Bird-Detection/outputs/arbimon/arbimon_detections.csv"
  presence_file:    "C:/Users/kelly/Documents/GitHub/Soundscape-Analysis-Rare-Bird-Detection/outputs/arbimon/arbimon_presence_by_site_day.csv"
  recordings_out:  "C:/Users/kelly/Documents/GitHub/Soundscape-Analysis-Rare-Bird-Detection/outputs/arbimon/recording_metadata.csv"
---

This document provides a standardized workflow for preparing Arbimon project data for analysis.
To begin, log in to your Arbimon account and request a full project backup from the site. Arbimon will email you a download link to a .zip file containing all of your project data.
Unzip this file to your computer. Inside, you will find the Recording Files CSVs (often split into multiple files if your project is large, starting with recordings.0001.csv), pattern matching results (pattern_matchings.0001.csv), site information (sites.0001.csv), as well as other analytical results and  metadata files. These files are the inputs for this prep script, which will combine, clean, and format them into a single dataset ready for analysis. Please do not adjust the code below, as it is written to be repeatable by any user of this script. The only information you should be changing is above in the YAML header. The parameters is where you will input your own file paths and parameters used wihtin the document. After that, it should be ready to go!

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(vroom)
  library(fs)
  library(glue)
  library(lubridate)
  library(stringr)
  library(tidyr)
})

proj <- params$project_dir
stopifnot(fs::dir_exists(proj))

all_files <- fs::dir_ls(proj, recurse = FALSE, type = "file")
fname <- fs::path_file(all_files)

read_many <- function(pattern) {
  hits <- all_files[grepl(pattern, fname, perl = TRUE)]
  if (!length(hits)) return(NULL)
  vroom::vroom(hits, .name_repair = "unique", show_col_types = FALSE)
}

# choose first existing datetime column
pick_datetime_col <- function(df, candidates) {
  for (c in candidates) if (c %in% names(df)) return(c)
  NA_character_
}

# robust parser allowing multiple formats
parse_dt <- function(x, orders, tz = "UTC") {
  if (inherits(x, "POSIXct")) return(lubridate::with_tz(x, tz))
 
  xc <- trimws(as.character(x)) # coerce to character and normalize common ISO-8601 shapes
  xc <- sub("T", " ", xc, fixed = TRUE)  # replace 'T' separator with space
  had_z <- grepl("Z$", xc, perl = TRUE) # drop trailing 'Z' (treat as UTC)
  xc <- sub("Z$", "", xc, perl = TRUE)  # trim whitespace

  # first try fast parser for ymd_hms
  dt <- suppressWarnings(lubridate::ymd_hms(xc, tz = tz, quiet = TRUE))

  # if all NA, try general multi-order parser
  if (all(is.na(dt))) {
    dt <- suppressWarnings(lubridate::parse_date_time(xc, orders = orders, tz = tz, quiet = TRUE))
  }

  # if source had Z (UTC), ensure tz is UTC regardless of system default
  if (any(had_z, na.rm = TRUE)) {
    dt <- lubridate::force_tz(dt, tzone = "UTC")
    # convert to requested tz (usually still UTC)
    dt <- lubridate::with_tz(dt, tzone = tz)
  }

  dt
}

# Normalize any "validated" style column to 0/1 integers
normalize_validated <- function(df) {
  if (is.null(df) || !nrow(df)) return(df)

  # common names Arbimon exports use (add more if you spot them)
  cand <- intersect(
    c("validated","is_validated","validation","is_verified","verified",
      "approved","is_approved","is_correct","correct","present","label"),
    tolower(names(df))
  )
  if (!length(cand)) {
    df$validated <- NA_integer_
    return(df)
  }
  col <- cand[1]
  v <- df[[ which(tolower(names(df)) == col) ]]

  # map to 0/1
  v_norm <-
    if (is.logical(v)) as.integer(v) else
    if (is.numeric(v)) as.integer(v != 0) else
    if (is.character(v)) {
      s <- trimws(tolower(v))
      as.integer(s %in% c("1","true","t","yes","y","valid","validated","correct","approved","present","positive"))
    } else NA_integer_

  df$validated <- v_norm
  df
}

```

```{r clean, include=FALSE}
# --- Recordings (bind all parts) ---
recs <- read_many(params$recordings_pattern)
stopifnot(!is.null(recs), nrow(recs) > 0)

# --- Sites metadata (optional but recommended) ---
sites <- read_many(params$sites_pattern)

recordings <- recs
if (!is.null(sites) && nrow(sites)) {
  rec_cols  <- as.character(params$join_by_recording)
  meta_cols <- as.character(params$join_by_metadata)
  stopifnot(length(rec_cols) == length(meta_cols))

  # sanity check columns exist
  miss_r <- setdiff(rec_cols, names(recordings))
  miss_m <- setdiff(meta_cols, names(sites))
  if (length(miss_r)) stop("Recordings missing join column(s): ", paste(miss_r, collapse=", "))
  if (length(miss_m)) stop("Sites missing join column(s): ", paste(miss_m, collapse=", "))

  join_map <- stats::setNames(meta_cols, rec_cols)
  recordings <- left_join(recordings, sites, by = join_map)
}

# --- Datetime parsing + calendar fields ---
dt_col <- pick_datetime_col(recordings, params$datetime_candidates)
if (is.na(dt_col)) stop("Could not find a datetime column among: ", paste(params$datetime_candidates, collapse=", "))

recordings <- recordings %>%
  mutate(
    datetime_parsed = parse_dt(.data[[dt_col]], params$datetime_orders, params$tz),
    Year            = year(datetime_parsed),
    Julian_Day      = yday(datetime_parsed)
  )

parsed_rate <- mean(!is.na(recordings$datetime_parsed)) * 100

# Optional: select columns
if (!is.null(params$keep_columns)) {
  keep <- unique(c(params$keep_columns, "recording_id", "datetime_parsed", "Year", "Julian_Day"))
  keep <- intersect(keep, names(recordings))
  recordings <- select(recordings, all_of(keep))
}

# Write clean recordings table
fs::dir_create(fs::path_dir(params$recordings_out))
write_csv(recordings, params$recordings_out)

```

# Joining forces

```{r join, include= FALSE}

%||%` <- function(x, y) if (is.null(x)) y else x

choose_first_col <- function(df, candidates) {
  cands <- tolower(candidates)
  nms   <- tolower(names(df))
  hit   <- which(nms %in% cands)
  if (length(hit)) names(df)[hit[1]] else NULL
}

# --- Pattern matching ROIs (bind) ---
pm <- read_many(params$pm_pattern)
if (!is.null(pm) && nrow(pm)) {
  pm <- pm %>% rename_with(tolower)
  # keep typical PM columns if present
  keep_pm <- intersect(
    c("pattern_matching_id","recording_id","species_id","songtype_id","x1","x2","y1","y2","score","validated"),
    names(pm)
  )
  pm <- pm %>% select(all_of(keep_pm))
  # coerce ids
  pm <- pm %>% mutate(
    species_id  = suppressWarnings(as.integer(species_id)),
    songtype_id = suppressWarnings(as.integer(songtype_id))
  )
}

# --- Species dictionary (sci name, songtype labels) ---
sp <- read_many(params$species_pattern)

sp_names   <- NULL
sp_songmap <- NULL

if (!is.null(sp) && nrow(sp)) {
  sp <- sp %>% dplyr::rename_with(tolower)

  # Find likely columns across export variants
  sci_col       <- choose_first_col(sp, c("scientific_name","species","latin_name","sci_name","taxon_name","name"))
  songtype_id_c <- choose_first_col(sp, c("songtype_id","songtype_ix","songtype_ix1","songtype"))
  songtype_c    <- choose_first_col(sp, c("songtype","song_type","songname"))

  if (!"species_id" %in% names(sp)) {
    stop("In species file, required key 'species_id' is missing. Columns present: ",
         paste(names(sp), collapse = ", "))
  }
  
# Build standardized fields; tolerate missing columns
  sp_std <- sp %>%
    mutate(
      species_id   = suppressWarnings(as.integer(.data[["species_id"]])),
      scientific_name = if (!is.null(sci_col)) as.character(.data[[sci_col]]) else NA_character_,
      songtype_id  = if (!is.null(songtype_id_c)) suppressWarnings(as.integer(.data[[songtype_id_c]])) else NA_integer_,
      songtype     = if (!is.null(songtype_c)) as.character(.data[[songtype_c]]) else NA_character_
    ) %>%
    # keep only our standardized columns; don't reference originals anymore
    select(species_id, scientific_name, songtype_id, songtype, any_of(c("family","taxon"))) %>%
    distinct()

  sp_names   <- sp_std %>% select(species_id, scientific_name) %>% distinct()
  sp_songmap <- sp_std %>% select(species_id, songtype_id, songtype) %>% distinct()

  # Friendly messages so you can see what was detected
  message("Species mapping: using scientific name column: ",
          sci_col %||% "<none>")
  message("Species mapping: using songtype_id column: ",
          songtype_id_c %||% "<none>",
          " | songtype label column: ",
          songtype_c %||% "<none>")
} else {
  warning("No species.* file(s) found; species names will be NA.")
}

# --- Manual validations (optional) ---
normalize_validated <- function(df) {
  if (is.null(df) || !nrow(df)) return(df)
  cand <- intersect(
    c("validated","is_validated","validation","is_verified","verified",
      "approved","is_approved","is_correct","correct","present","label"),
    tolower(names(df))
  )
  if (!length(cand)) { df$validated <- NA_integer_; return(df) }
  col <- cand[1]
  v <- df[[ which(tolower(names(df)) == col) ]]
  v_norm <-
    if (is.logical(v)) as.integer(v) else
    if (is.numeric(v)) as.integer(v != 0) else
    if (is.character(v)) {
      s <- trimws(tolower(v))
      as.integer(s %in% c("1","true","t","yes","y","valid","validated","correct","approved","present","positive"))
    } else NA_integer_
  df$validated <- v_norm
  df
}

# Read/normalize validations (if you haven’t already)
if (!exists("val")) {
  val <- read_many(params$validations_pattern)
  if (!is.null(val) && nrow(val)) {
    val <- val %>% rename_with(tolower)
    score_col <- intersect(c("score","confidence","probability"), names(val))[1]
    keep_val  <- c(intersect(c("recording_id","species_id"), names(val)), score_col)
    keep_val  <- keep_val[!is.na(keep_val)]
    val <- val %>%
      select(all_of(keep_val), everything()) %>%
      normalize_validated() %>%
      mutate(species_id = suppressWarnings(as.integer(species_id)))
  } else {
    val <- NULL
    score_col <- NULL
  }
}


# --- Build detections from PM ---
det_pm <- NULL
if (!is.null(pm) && nrow(pm)) {
  det_pm <- pm %>%
    left_join(rec_core, by = "recording_id") %>%
    left_join(sp_names,   by = "species_id") %>%
    left_join(sp_songmap, by = c("species_id","songtype_id")) %>%
    mutate(
      common_name = NA_character_,          # not present in this export
      detection_source = "pattern_match"
    ) %>%
    select(recording_id, site_id, site_label, .dt, date, Year, Julian_Day,
           species_id, scientific_name, common_name,
           songtype_id, songtype,
           any_of(c("pattern_matching_id","x1","x2","y1","y2","score","validated")),
           detection_source)
}

# --- Build detections from validations ---
det_pm <- NULL
if (!is.null(pm) && nrow(pm)) {
  det_pm <- pm %>%
    left_join(rec_core,   by = "recording_id") %>%
    left_join(sp_names,   by = "species_id") %>%
    { if (!is.null(sp_songmap)) left_join(., sp_songmap, by = c("species_id","songtype_id")) else . } %>%
    mutate(
      common_name = NA_character_,
      detection_source = "pattern_match"
    ) %>%
    select(recording_id, site_id, site_label, .dt, date, Year, Julian_Day,
           species_id, scientific_name, common_name,
           any_of(c("songtype_id","songtype",
                    "pattern_matching_id","x1","x2","y1","y2","score","validated")),
           detection_source)
}

det_val <- NULL
if (!is.null(val) && nrow(val)) {
  det_val <- val %>%
    left_join(rec_core, by = "recording_id") %>%
    left_join(sp_names, by = "species_id") %>%
    mutate(
      common_name = NA_character_,
      detection_source = "manual_validation"
    ) %>%
    select(recording_id, site_id, site_label, .dt, date, Year, Julian_Day,
           species_id, scientific_name, common_name,
           any_of(c("validated", score_col)),
           detection_source)
}

detections <- dplyr::bind_rows(det_pm, det_val) %>% distinct()
```

# Presence

```{r presence, include = FALSE}
if (exists("detections") && nrow(detections)) {
  presence <- detections %>%
    group_by(site_id, site_label, date, species_id, scientific_name, common_name) %>%
    summarize(
      n_detections = n(),
      any_validated = any(!is.na(validated) & validated == 1),
      sources = paste(sort(unique(detection_source)), collapse = "+"),
      .groups = "drop"
    )
  fs::dir_create(fs::path_dir(params$presence_file))
  write_csv(presence, params$presence_file)
```


Locate and combine recording files (recording.0001.csv)
```{r CSV Binding, include=FALSE}
stopifnot(fs::dir_exists(params$project_dir))

all_files <- fs::dir_ls(path = params$project_dir, recurse = FALSE, type = "file")
rec_files <- all_files[grepl(params$recordings_pattern, fs::path_file(all_files), perl = TRUE)]

if (length(rec_files) == 0) {
  stop(glue("No files matching '{params$recordings_pattern}' found in '{params$project_dir}'."))
}

recordings_raw <- vroom::vroom(rec_files, .name_repair = "unique")

glimpse(recordings_raw, width = 80)
recordings <- recordings_raw

```
