---
title: "Arbimon Data Prep Script"
author: "Kelly Faller"
date: "2024-10-11"
output: html_document
# >>> Edit only these params. These will change when you run this file, but the rest should be automated <<<
params:
  project_dir: "C:/Users/kelly/Documents/Master's Project Files/arbimon_backups/salt-marsh-restoration-soundscape-project_2025-08-12_export" #change this !!!
   # File name patterns (do not change unless your export uses different names)
  recordings_pattern: "^recordings\\."
  sites_pattern: "^sites\\."
  pm_pattern: "^pattern_matching_rois\\."
  pm_jobs_pattern: "^pattern_matchings\\."
  species_pattern: "^species\\."
  validations_pattern: "^recording_validations\\."  # manual labels - not from pm analysis
  # Time zone of the recorders. Arbimon's `datetime` column is the local time on the recorder (it matches the
  # wav file name); `datetime_utc` is UTC. If the sites table has a timezone column, that is used instead.
  tz: "America/New_York"
  use_validations: true                        # set to false to ignore manual labels
  # Outputs. Relative paths are inside this repository; full paths (C:/...) also work.
  recordings_out:  "data/arbimon/recording_metadata.csv"
  detections_file: "data/arbimon/pm_detections.csv"
  validations_file: "data/arbimon/manual_detections.csv"
  presence_file:   "data/arbimon/pm_presence_by_site_day.csv"
---

This document provides a standardized workflow for preparing Arbimon project data for analysis.
To begin, log in to your Arbimon account and request a full project backup from the site. Arbimon will email you a download link to a .zip file containing all of your project data.
Unzip this file to your computer. Inside, you will find the Recording Files CSVs (often split into multiple files if your project is large, starting with recordings.0001.csv), pattern matching results (pattern_matchings.0001.csv), site information (sites.0001.csv), as well as other analytical results and  metadata files. These files are the inputs for this prep script, which will combine, clean, and format them into a single dataset ready for analysis. Please do not adjust the code below, as it is written to be repeatable by any user of this script. The only information you should be changing is above in the YAML header. The parameters is where you will input your own file paths and parameters used wihtin the document. After that, it should be ready to go!

The backup is set up like a relational database: recordings link to sites by `site_id`; pattern matching ROIs and manual validations link to recordings by `recording_id` and to species by `species_id` + `songtype_id`; ROIs link to their pattern matching job by `pattern_matching_id`.

How validation is coded in the backup:

- **Pattern matching ROIs** (`validated`): 1 = the match was confirmed as the species, 0 = the match was reviewed and rejected (the template matched something else), blank = not reviewed.
- **Manual recording validations** (`present_visualizer`, `present_pm`, `present_aed`): 1 = species present in that recording, 0 = reviewed and absent, for each place the validation was made in Arbimon.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(fs)
  library(glue)
  library(lubridate)
  library(stringr)
  library(tidyr)
})

proj <- params$project_dir
if (!fs::dir_exists(proj)) stop("project_dir doesn't exist: ", proj, "\nChange project_dir in the YAML header to your unzipped Arbimon backup folder.")

all_files <- fs::dir_ls(proj, recurse = FALSE, type = "file")
fname <- fs::path_file(all_files)

# reads (and stacks) every file matching a pattern, e.g. recordings.0001.csv ... recordings.0011.csv.
# Everything is read as text so the parts always stack (ids stay text, which keeps joins exact); numbers are converted where needed.
read_many <- function(pattern) {
  hits <- all_files[grepl(pattern, fname, perl = TRUE)]
  if (!length(hits)) return(NULL)
  dplyr::bind_rows(lapply(sort(hits), readr::read_csv, col_types = readr::cols(.default = readr::col_character()), progress = FALSE))
}

# output paths: relative paths are inside this repository
out_path <- function(p) if (fs::is_absolute_path(p)) p else here::here(p)
write_out <- function(df, p) {
  p <- out_path(p)
  fs::dir_create(fs::path_dir(p))
  readr::write_csv(df, p)
  message("Wrote ", nrow(df), " rows to ", p)
}

# Arbimon writes date-times as "06/30/2024 15:00:00"; older exports used ISO "2024-06-30 15:00:00"
parse_arbimon_dt <- function(x, tz) lubridate::parse_date_time(x, orders = c("mdy HMS", "Ymd HMS", "mdy HM", "Ymd HM"), tz = tz, quiet = TRUE)
```

```{r clean, include=FALSE}
# --- Recordings (bind all parts) ---
recs <- read_many(params$recordings_pattern)
if (is.null(recs) || !nrow(recs)) stop("No recordings.*.csv files found in ", proj)

# --- Sites metadata ---
sites <- read_many(params$sites_pattern)
if (is.null(sites)) stop("No sites.*.csv file found in ", proj, " - it's needed to label recordings with station names.")

recordings <- recs %>%
  dplyr::select(-dplyr::any_of("url")) %>%                       # signed download links; they expire and shouldn't be shared
  dplyr::left_join(sites, by = "site_id")

# --- Datetime parsing + calendar fields ---
# `datetime` is local recorder time; it is parsed in the site's own time zone (or params$tz if the sites table has none)
recordings <- recordings %>%
  dplyr::mutate(site_tz = if ("timezone" %in% names(.)) dplyr::coalesce(timezone, params$tz) else params$tz)

recordings$datetime_parsed <- lubridate::as_datetime(rep(NA_real_, nrow(recordings)), tz = params$tz)
for (z in unique(recordings$site_tz)) {
  i <- recordings$site_tz == z
  recordings$datetime_parsed[i] <- lubridate::with_tz(parse_arbimon_dt(recordings$datetime[i], z), params$tz)
}

recordings <- recordings %>%
  dplyr::mutate(
    datetime_utc_parsed = parse_arbimon_dt(datetime_utc, "UTC"),
    source_file         = stringr::str_match(meta, '"filename":"([^"]+)"')[, 2],
    Year                = lubridate::year(datetime_parsed),
    Julian_Day          = lubridate::yday(datetime_parsed))

parsed_rate <- mean(!is.na(recordings$datetime_parsed)) * 100
if (parsed_rate < 100) warning(sprintf("%.1f%% of recording date-times could not be parsed", 100 - parsed_rate))

# Write clean recordings table (date-times written as local time with offset, e.g. 2024-06-30T15:00:00-04:00)
write_out(recordings %>% dplyr::mutate(dplyr::across(c(datetime_parsed), ~ format(.x, "%Y-%m-%dT%H:%M:%S%z"))),
          params$recordings_out)

# the columns every detection table gets from its recording
rec_core <- recordings %>%
  dplyr::transmute(recording_id, site_id, site_name = name,
                   recording_datetime = datetime_parsed,
                   recording_date = as.Date(datetime_parsed, tz = params$tz),
                   Year, Julian_Day)
recordings <- recordings %>% dplyr::mutate(dplyr::across(dplyr::any_of(c("sample_rate", "duration", "samples", "latitude", "longitude")), as.numeric))
```

The recordings table has `r format(nrow(recordings), big.mark = ",")` one-minute recordings from `r dplyr::n_distinct(recordings$site_id)` sites; `r round(parsed_rate, 1)`% of date-times parsed.

# Joining forces

```{r join, include=FALSE}
# --- Species dictionary (scientific name + songtype for each species_id x songtype_id) ---
sp <- read_many(params$species_pattern)
if (is.null(sp)) stop("No species.*.csv file found in ", proj)
sp <- sp %>%
  dplyr::mutate(scientific_name = stringr::str_squish(scientific_name), family = stringr::str_squish(family)) %>%
  dplyr::distinct(species_id, songtype_id, .keep_all = TRUE)

# --- Pattern matching jobs (name, template, threshold) ---
pm_jobs <- read_many(params$pm_jobs_pattern)

# --- Pattern matching ROIs ---
pm <- read_many(params$pm_pattern)
det_pm <- NULL
if (!is.null(pm) && nrow(pm)) {
  det_pm <- pm %>%
    dplyr::mutate(dplyr::across(dplyr::any_of(c("x1", "x2", "y1", "y2", "score")), as.numeric),
                  validated = suppressWarnings(as.integer(validated))) %>%
    dplyr::left_join(rec_core, by = "recording_id") %>%
    dplyr::left_join(sp %>% dplyr::select(species_id, songtype_id, scientific_name, songtype), by = c("species_id", "songtype_id")) %>%
    { if (!is.null(pm_jobs)) dplyr::left_join(., pm_jobs %>% dplyr::select(pattern_matching_id, pm_name = name, dplyr::any_of(c("template_name", "playlist_name"))),
                                              by = "pattern_matching_id") else . } %>%
    dplyr::mutate(
      validation_status  = dplyr::case_when(validated == 1 ~ "confirmed", validated == 0 ~ "rejected", TRUE ~ "not reviewed"),
      detection_source   = "pattern_match",
      detection_datetime = recording_datetime + x1,           # start of the matched sound within the 1-minute recording
      detection_date     = as.Date(detection_datetime, tz = params$tz))
}

# --- Manual validations ---
det_val <- NULL
if (isTRUE(params$use_validations)) {
  val <- read_many(params$validations_pattern)
  if (!is.null(val) && nrow(val)) {
    present_cols <- grep("^present", names(val), value = TRUE)
    pres <- as.data.frame(lapply(val[present_cols], function(x) suppressWarnings(as.integer(x))))
    det_val <- val %>%
      dplyr::select(-dplyr::any_of(c("scientific_name", "songtype"))) %>%    # taken from the species table instead
      dplyr::mutate(present = ifelse(rowSums(!is.na(pres)) == 0, NA_integer_, as.integer(rowSums(pres == 1, na.rm = TRUE) > 0))) %>%
      dplyr::left_join(rec_core, by = "recording_id") %>%
      dplyr::left_join(sp %>% dplyr::select(species_id, songtype_id, scientific_name, songtype), by = c("species_id", "songtype_id")) %>%
      dplyr::mutate(detection_source = "manual_validation",
                    detection_datetime = recording_datetime,
                    detection_date = recording_date)
  }
}

# date-times are written as local time with the UTC offset so they can't be mistaken for UTC
fmt_dt <- function(df) df %>% dplyr::mutate(dplyr::across(c(recording_datetime, detection_datetime), ~ format(.x, "%Y-%m-%dT%H:%M:%S%z")))
if (!is.null(det_pm))  write_out(fmt_dt(det_pm), params$detections_file)
if (!is.null(det_val)) write_out(fmt_dt(det_val), params$validations_file)
```

```{r join checks, echo=FALSE}
if (!is.null(det_pm)) {
  det_pm %>%
    dplyr::group_by(pattern_matching_id, dplyr::across(dplyr::any_of("pm_name")), scientific_name) %>%
    dplyr::summarise(ROIs = dplyr::n(), Confirmed = sum(validation_status == "confirmed"),
                     Rejected = sum(validation_status == "rejected"), `Not reviewed` = sum(validation_status == "not reviewed"),
                     `Unmatched recordings` = sum(is.na(site_id)), .groups = "drop") %>%
    knitr::kable(caption = "Pattern matching ROIs by job")
}
```

# Presence

Pattern matching results summarised by site and day. `n_detections` counts every ROI (all template matches), `n_confirmed` and `n_rejected` count ROIs reviewed as the species or not, and `present` is TRUE only if at least one ROI that day was confirmed. Unreviewed or rejected matches are **not** evidence the species was there.

```{r presence, include=FALSE}
if (!is.null(det_pm) && nrow(det_pm)) {
  presence <- det_pm %>%
    dplyr::group_by(site_id, site_name, detection_date, species_id, scientific_name) %>%
    dplyr::summarise(
      n_detections = dplyr::n(),
      n_recordings = dplyr::n_distinct(recording_id),
      n_confirmed  = sum(validation_status == "confirmed"),
      n_rejected   = sum(validation_status == "rejected"),
      present      = n_confirmed > 0,
      .groups = "drop")
  write_out(presence, params$presence_file)
}
```

THIS IS FROM THE LAST ANALYSIS. PLEASE ONLY USE IF NECESSARY. I DID NOT WANT TO DELETE. (Turned off with `eval=FALSE` - it refers to a `recording_metadata` object from an older version of this script and stops the knit if run.)

```{r Setup Arbimon, include=FALSE, eval=FALSE}

# Robust datetime handling from prep output:
# Accept 'datetime_parsed' (preferred) or fallback to 'datetime'
parse_dt <- function(x, tz = "UTC") {
  if (inherits(x, "POSIXct")) return(with_tz(x, tz))
  x_norm <- if (is.character(x)) sub("Z$", "+00:00", x) else x  # handle trailing Z
  try_order <- function(fun, ...) suppressWarnings(fun(...))
  out <- try_order(ymd_hms, x_norm, tz = tz, quiet = TRUE)
  if (all(is.na(out))) out <- try_order(mdy_hms, x_norm, tz = tz, quiet = TRUE)
  if (all(is.na(out))) out <- try_order(ymd_hm,  x_norm, tz = tz, quiet = TRUE)
  if (all(is.na(out))) out <- try_order(mdy_hm,  x_norm, tz = tz, quiet = TRUE)
  if (all(is.na(out))) out <- try_order(ymd,     x_norm, tz = tz, quiet = TRUE)
  if (all(is.na(out))) out <- try_order(mdy,     x_norm, tz = tz, quiet = TRUE)
  out
}

dt_col <- dplyr::case_when(
  "datetime_parsed" %in% names(recording_metadata) ~ "datetime_parsed",
  "datetime"        %in% names(recording_metadata) ~ "datetime",
  TRUE ~ NA_character_
)
if (is.na(dt_col)) stop("recording_metadata needs a datetime column (datetime_parsed or datetime).")

recording_metadata <- recording_metadata %>%
  mutate(
    datetime_parsed = if (dt_col == "datetime_parsed") .data[[dt_col]] else parse_dt(.data[[dt_col]], tz = "UTC"),
    Year            = if (!"Year" %in% names(.)) year(datetime_parsed) else Year,
    Julian_Day      = if (!"Julian_Day" %in% names(.)) yday(datetime_parsed) else Julian_Day
  )

# Optional 'name' fallback
if (!"name" %in% names(recording_metadata)) {
  nm <- intersect(c("site","station","recorder","Site","Station","Recorder"), names(recording_metadata))[1]
  if (!is.na(nm)) recording_metadata$name <- recording_metadata[[nm]]
}

```
