---
title: "BirdNET Outputs Data Prep"
author: "Kelly Faller"
date: '2023-06-08'
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages


```{r, warning=FALSE, include=FALSE}
library(tidyverse)
library(lubridate)
library(bit64)
library(magrittr)
library(plyr)
library(readr)
library(data.table)
library(chron)
```

## Create one CSV

Taking the folder full of individual CSV files for each output file (in BirdNET, there is one for every recording) and creating one master CSV

Changing the working directory back to my personal wd that connects with GitHub

Working directory set up - location where the CSV's will be held. This will change every time you run the script. Currently it is referencing my working ARU external
harddrive. I have it broken down by station, deployment, and day

```{r CSV Binding, include=FALSE}
root <- "C:/Users/kelly/Documents/Master's Project Files/02_data_processed/birdNET/2024/LHC_01"

all_paths <- list.files(
  path = root,
  pattern = "BirdNET\\.results\\.csv$",
  full.names = TRUE,
  recursive = TRUE
)

all_content <- lapply(all_paths, function(p) {
  x <- tryCatch(fread(p, encoding = "UTF-8", data.table = FALSE),
               error = function(e) NULL)
  if (is.null(x) || nrow(x) == 0) return(NULL)
  x
})
all_content <- Filter(Negate(is.null), all_content)

all_result <- rbindlist(all_content, fill = TRUE)

# FIX: rename the first 6 columns by position (Start/End/Scientific/Common/Confidence/File)
all_result <- as.data.frame(all_result)
if (ncol(all_result) >= 6) {
  names(all_result)[1:6] <- c("Start_s","End_s","Scientific_Name","Common_Name","Confidence","File")
}

# drop any non-detection rows if present
all_result <- all_result %>% filter(!is.na(Start_s))

# parse Detector_ID/Date/Time from File path and create Year + DateTime
species_data <- all_result %>%
  mutate(
    Audio_File = basename(as.character(File)),
    Audio_Stem = sub("\\.wav$", "", Audio_File, ignore.case = TRUE)
  ) %>%
  separate(Audio_Stem, c("Detector_ID", "Date", "Time"), sep = "_", extra = "drop") %>%
  mutate(
    Time = substr(Time, 1, 6),
    Date = ymd(Date),
    Year = year(Date),
    DateTime = ymd_hms(paste(Date, Time), tz = "UTC")
  )

# ---- Metadata join (Year + Unit) ----
meta_path <- "C:/Users/kelly/Documents/GitHub/Soundscape-Analysis-Rare-Bird-Detection/data/metadata/LHC-Cattus_ARU_Site_Metadata.csv"
metadata <- fread(meta_path, encoding = "UTF-8", data.table = FALSE)

metadata <- metadata %>%
  mutate(
    Year = as.integer(Year),
    Unit = trimws(as.character(Unit))
  )

species_data <- species_data %>%
  mutate(Detector_ID = trimws(as.character(Detector_ID))) %>%
  left_join(
    metadata %>% select(Year, Unit, Site, Station, Lat, Long),
    by = c("Year" = "Year", "Detector_ID" = "Unit")
  )

write.csv(
  species_data,
  "C:/Users/kelly/Documents/GitHub/Soundscape-Analysis-Rare-Bird-Detection/data/processed/birdNET/individual/LHC_01_2024_BirdNET_processed.csv",
  row.names = FALSE
)

```

rebuild
```{r rebuilt, include=FALSE}
# Replace your entire script with this (robust + adds a stable filename column)
# Libraries


root <- "C:/Users/kelly/Documents/Master's Project Files/02_data_processed/birdNET/2024/LHC_01"

all_paths <- list.files(
  path = root,
  pattern = "BirdNET\\.results\\.csv$",
  full.names = TRUE,
  recursive = TRUE
)

all_content <- lapply(all_paths, function(p) {
  x <- tryCatch(
    fread(p, encoding = "UTF-8", data.table = FALSE),
    error = function(e) NULL
  )
  if (is.null(x) || nrow(x) == 0) return(NULL)
  x
})
all_content <- Filter(Negate(is.null), all_content)

all_result <- rbindlist(all_content, fill = TRUE)

# Rename the first 6 columns by position (Start/End/Scientific/Common/Confidence/FilePath)
all_result <- as.data.frame(all_result)
if (ncol(all_result) >= 6) {
  names(all_result)[1:6] <- c(
    "Start_s", "End_s", "Scientific_Name", "Common_Name", "Confidence", "File"
  )
}

# Drop any non-detection rows if present
all_result <- all_result %>% filter(!is.na(Start_s))

# ---- Parse metadata from filename (robust to 3/4/6-part stems) ----
parse_birdnet_stem <- function(stem) {
  parts <- unlist(strsplit(stem, "_", fixed = TRUE))
  n <- length(parts)

  # defaults
  out <- list(
    File_Site   = NA_character_,
    Detector_ID = NA_character_,  # e.g., LHC_01
    Unit        = NA_character_,  # e.g., Unit4 (often what metadata uses)
    Date_chr    = NA_character_,
    Time_chr    = NA_character_
  )

  if (n == 3) {
    # old: detector_date_time
    out$Detector_ID <- parts[1]
    out$Date_chr    <- parts[2]
    out$Time_chr    <- parts[3]
  } else if (n == 4) {
    # new simple: site_detector_date_time
    out$File_Site   <- parts[1]
    out$Detector_ID <- parts[2]
    out$Date_chr    <- parts[3]
    out$Time_chr    <- parts[4]
  } else if (n >= 6) {
    # current: LighthouseCenter_LHC_01_Unit4_YYYYMMDD_HHMMSS
    # Take last two as date/time, token before date as Unit, first token as site.
    out$File_Site <- parts[1]
    out$Unit      <- parts[n - 2]
    out$Date_chr  <- parts[n - 1]
    out$Time_chr  <- parts[n]

    # Detector_ID as the middle tokens excluding [site] and [unit/date/time]
    # For your example this becomes LHC_01
    middle <- parts[2:(n - 3)]
    out$Detector_ID <- paste(middle, collapse = "_")
  } else {
    # fallback: at least try last two as date/time if present
    if (n >= 2) {
      out$Date_chr <- parts[n - 1]
      out$Time_chr <- parts[n]
    }
    if (n >= 1) out$Detector_ID <- parts[1]
  }

  as.data.frame(out, stringsAsFactors = FALSE)
}

species_data <- all_result %>%
  mutate(
    Audio_File = basename(as.character(File)),
    Audio_Stem = sub("\\.wav$", "", Audio_File, ignore.case = TRUE)
  )

parsed_df <- do.call(
  rbind,
  lapply(species_data$Audio_Stem, parse_birdnet_stem)
)

species_data <- bind_cols(species_data, parsed_df) %>%
  mutate(
    Time_chr = substr(Time_chr, 1, 6),
    Date = ymd(Date_chr),
    Year = year(Date),
    DateTime = ymd_hms(paste(Date_chr, Time_chr), tz = "UTC")
  )

# ---- Metadata join (Year + Unit) ----
meta_path <- "C:/Users/kelly/Documents/GitHub/Soundscape-Analysis-Rare-Bird-Detection/data/metadata/LHC-Cattus_ARU_Site_Metadata.csv"
metadata <- fread(meta_path, encoding = "UTF-8", data.table = FALSE)

metadata <- metadata %>%
  mutate(
    Year = as.integer(Year),
    Unit = trimws(as.character(Unit))
  ) %>%
  dplyr::rename(Meta_Site = Site)   # prevent Site.x / Site.y collisions

species_data <- species_data %>%
  mutate(
    Year = as.integer(Year),
    Unit = trimws(as.character(Unit)),
    Detector_ID = trimws(as.character(Detector_ID))
  ) %>%
  left_join(
    metadata %>% select(Year, Unit, Meta_Site, Station, Lat, Long),
    by = c("Year" = "Year", "Unit" = "Unit")
  )

write.csv(
  species_data,
  "C:/Users/kelly/Documents/GitHub/Soundscape-Analysis-Rare-Bird-Detection/data/processed/birdNET/individual/LHC_01_2024_BirdNET_processed.csv",
  row.names = FALSE
)


```

## Create final CSV

Combining the CSVs created from above chunk and create a final working CSV for graphics

```{r CSV Binding2, include=FALSE}

# read file path
all_paths <-
  list.files(path = "C:/Users/kelly/Documents/GitHub/Soundscape-Analysis-Rare-Bird-Detection/data/processed/birdNET/individual/",
             pattern = "*.csv",
             full.names = TRUE)

# read file content
all_content <-
  all_paths %>%
  lapply(read.table,
         header = TRUE,
         sep = ",",
         encoding = "UTF-8")

# read file name
all_filenames <- all_paths %>%
  basename() %>%
  as.list()

# combine file content list and file name list
all_lists <- mapply(c, all_content, all_filenames, SIMPLIFY = FALSE)

# unlist all lists and change column name
all_result <- rbindlist(all_lists, fill = T)


write.csv(all_result,"C:/Users/kelly/Documents/GitHub/Soundscape-Analysis-Rare-Bird-Detection/data/processed/birdNET/birdNET_results.csv", row.names = FALSE)  

```

rebuilding

```{r CSV Binding rebuilt, include=FALSE}

# Replace your entire script with this (robust + adds a stable filename column)

library(data.table)

# 1) List CSV files (use a proper regex for .csv)
all_paths <- list.files(
  path = "C:/Users/kelly/Documents/GitHub/Soundscape-Analysis-Rare-Bird-Detection/data/processed/birdNET/individual/",
  pattern = "\\.csv$",
  full.names = TRUE
)

# 2) Read each CSV robustly, and add a consistent source_file column
all_result <- rbindlist(
  lapply(all_paths, function(p) {
    dt <- fread(
      p,
      encoding = "UTF-8"
      # If you suspect messy quotes, you can try either of these:
      # , quote = "\""
      # , fill = TRUE
    )
    dt[, source_file := basename(p)]
    dt
  }),
  fill = TRUE
)

# 3) Write combined output
fwrite(
  all_result,
  file = "C:/Users/kelly/Documents/GitHub/Soundscape-Analysis-Rare-Bird-Detection/data/processed/birdNET/birdNET_results.csv"
)
```



