---
title: "BirdNET Outputs Data Prep"
author: "Kelly Faller"
date: '2023-06-08'
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages


```{r, warning=FALSE, include=FALSE}
library(tidyverse)
library(lubridate)
library(bit64)
library(magrittr)
library(plyr)
library(readr)
library(data.table)
library(chron)
```

## Create one CSV

Taking the folder full of individual CSV files for each output file (in BirdNET, there is one for every recording) and creating one master CSV

Changing the working directory back to my personal wd that connects with GitHub

Working directory set up - location where the CSV's will be held. This will change every time you run the script. Currently it is referencing my working ARU external
harddrive. I have it broken down by station, deployment, and day

```{r CSV Binding, include=FALSE}
root <- "C:/Users/kelly/Documents/Master's Project Files/02_data_processed/birdNET/2025/CAT_A1"

all_paths <- list.files(
  path = root,
  pattern = "BirdNET\\.results\\.csv$",
  full.names = TRUE,
  recursive = TRUE
)

all_content <- lapply(all_paths, function(p) {
  x <- tryCatch(fread(p, encoding = "UTF-8", data.table = FALSE),
               error = function(e) NULL)
  if (is.null(x) || nrow(x) == 0) return(NULL)
  x
})
all_content <- Filter(Negate(is.null), all_content)

all_result <- rbindlist(all_content, fill = TRUE)

# FIX: rename the first 6 columns by position (Start/End/Scientific/Common/Confidence/File)
all_result <- as.data.frame(all_result)
if (ncol(all_result) >= 6) {
  names(all_result)[1:6] <- c("Start_s","End_s","Scientific_Name","Common_Name","Confidence","File")
}

# drop any non-detection rows if present
all_result <- all_result %>% filter(!is.na(Start_s))

# parse Detector_ID/Date/Time from File path and create Year + DateTime
species_data <- all_result %>%
  mutate(
    Audio_File = basename(as.character(File)),
    Audio_Stem = sub("\\.wav$", "", Audio_File, ignore.case = TRUE)
  ) %>%
  separate(Audio_Stem, c("Detector_ID", "Date", "Time"), sep = "_", extra = "drop") %>%
  mutate(
    Time = substr(Time, 1, 6),
    Date = ymd(Date),
    Year = year(Date),
    DateTime = ymd_hms(paste(Date, Time), tz = "UTC")
  )

# ---- Metadata join (Year + Unit) ----
meta_path <- "C:/Users/kelly/Documents/GitHub/Soundscape-Analysis-Rare-Bird-Detection/data/metadata/LHC-Cattus_ARU_Site_Metadata.csv"
metadata <- fread(meta_path, encoding = "UTF-8", data.table = FALSE)

metadata <- metadata %>%
  mutate(
    Year = as.integer(Year),
    Unit = trimws(as.character(Unit))
  )

species_data <- species_data %>%
  mutate(Detector_ID = trimws(as.character(Detector_ID))) %>%
  left_join(
    metadata %>% select(Year, Unit, Site, Station, Lat, Long),
    by = c("Year" = "Year", "Detector_ID" = "Unit")
  )

write.csv(
  species_data,
  "C:/Users/kelly/Documents/GitHub/Soundscape-Analysis-Rare-Bird-Detection/data/processed/birdNET/2025 individual/CAT_A1_BirdNET_processed.csv",
  row.names = FALSE
)

```

## Create final CSV

Combining the CSVs created from above chunk and create a final working CSV for graphics

```{r CSV Binding, include=FALSE}

# read file path
all_paths <-
  list.files(path = "C:/Users/kelly/Documents/GitHub/Soundscape-Analysis-Rare-Bird-Detection/data/processed/birdNET/2025 individual/",
             pattern = "*.csv",
             full.names = TRUE)

# read file content
all_content <-
  all_paths %>%
  lapply(read.table,
         header = TRUE,
         sep = ",",
         encoding = "UTF-8")

# read file name
all_filenames <- all_paths %>%
  basename() %>%
  as.list()

# combine file content list and file name list
all_lists <- mapply(c, all_content, all_filenames, SIMPLIFY = FALSE)

# unlist all lists and change column name
all_result <- rbindlist(all_lists, fill = T)
# change column name
names(all_result)[13] <- "File_Name"

write.csv(all_result,"C:/Users/kelly/Documents/GitHub/Soundscape-Analysis-Rare-Bird-Detection/data/processed/birdNET/birdNET_results_2025.csv", row.names = FALSE)  

```


