---
title: "Data Prep Script"
author: "Kelly Faller"
date: '2023-05-20'
output: html_document
# >>> Edit these params; the rest should run without changes <<<
params:
  # Kaleidoscope output folder, laid out as <root>/<year>/<station>/acousticindex.csv. Relative = inside this repo.
  kaleidoscope_root: "data/processed/kaleidoscope_outputs"
  # Combined file read by 02-Acoustics-Analysis.Rmd (configs/project.yml -> paths$index)
  combined_out: "data/processed/kaleidoscope_outputs/acoustic_indices_20242025combined.csv"
  tz: "America/New_York"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Load Packages

```{r, warning=FALSE, include=FALSE}
library(tidyverse)
library(lubridate)
source(here::here("R", "kaleidoscope_functions.R"))
in_repo <- function(p) if (fs::is_absolute_path(p)) p else here::here(p)
```

## Create one CSV

Taking the folder of Kaleidoscope acoustic index outputs (one `acousticindex.csv` per station and year) and creating one master CSV. Station and year come from the folder names, because newer Kaleidoscope file names only contain the recorder unit (e.g. `Unit4_20250603_210000.wav`). Date-times are the local recorder time.

This uses the same reader as `kaleidoscope_data_prep_db.R` (`R/kaleidoscope_functions.R`), so the csv and the database always contain the same rows.

```{r CSV Binding, include=FALSE}
index <- read_kaleidoscope_indices(in_repo(params$kaleidoscope_root), tz = params$tz)
checks <- check_kaleidoscope_indices(index)
```

## Checks

```{r Checks, echo=FALSE, results="asis"}
print(knitr::kable(checks$by_station, caption = "Recordings by station and year"))
print(knitr::kable(checks$problems, caption = "Checks (all should be 0)"))
```

## Write

```{r Write, include=FALSE}
readr::write_csv(index %>% dplyr::mutate(recorded_at = format(recorded_at, "%Y-%m-%d %H:%M:%S")), in_repo(params$combined_out))
```

Wrote `r format(nrow(index), big.mark = ",")` recordings to `r params$combined_out`.

## Previous version

Turned off (`eval=FALSE`). It was written for an older Kaleidoscope export where `TIME` held minutes:seconds; with current exports (`TIME` = "05:00:00.000") gluing `HOUR` onto `TIME` gives strings like "5:05:00:00" and every date-time becomes NA. It also used `setwd()` with hard-coded paths.

```{r CSV Binding old, include=FALSE, eval=FALSE}

setwd('C:/Users/kfaller/Documents/GitHub/Acoustics/Acoustic Index Files')

df<-list.files(path='C:/Users/kfaller/Documents/GitHub/Acoustics/Acoustic Index Files')%>%
    lapply(read_csv) %>%
    bind_rows

setwd('C:/Users/kfaller/Documents/GitHub/Acoustics')

write.csv(df,"ARU_Data_Indicies.csv", row.names = FALSE)

aru_data1<- read.csv("ARU_Data_Indicies.csv", header=TRUE, sep = ",", stringsAsFactors=FALSE)

aru_data2 <- aru_data1 %>%
    rename(MINUTESECOND = TIME)%>%
    unite(TIME, HOUR, MINUTESECOND, sep = ":")

aru_data3 <- aru_data2 %>%
   dplyr::mutate(TIME2 = substr(TIME, 1, nchar(TIME)-3))%>%
   unite(DATETIME, DATE, TIME2, sep = " ")

aru_data3$DATETIME <- mdy_hms(aru_data3$DATETIME)

aru_data4 <- aru_data3 %>%  mutate(DATE = as.Date(DATETIME)) %>%
         mutate(DATE = format(DATE, "%m/%d/%Y"))

aru_data4$DATE <- mdy(aru_data4$DATE)

write.csv(aru_data4,"ARU_Data_Indicies.csv", row.names = FALSE)

```
