---
title: "LCLT Bird Script"
author: "Kelly Faller"
date: "2024-10-16"
output:
  html_document:
    toc: true
    toc_float: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages

```{r, warning=FALSE, include=FALSE}
#graphics
library(ggplot2)
library(tidyverse)
library(lubridate)
library(scales)
library(ggpubr)
library(rstatix)
library(ggpmisc)
library(tibble)
library(timetk)
library(plotly)
library(data.table)
library(plyr)
library(multcomp)
library(purrr)
library(ggsignif)
# density and abundance estimations
library(leaflet)
library(vegan)
library(gt)
library(knitr)
library(kableExtra)
library(envalysis)
library(sads)
library(effects)
library(unmarked)
#reproducibility
library(yaml)
library(here)
#summary
library(vtable)
# also used with pkg:: (install if missing): leafpop, broom
```

## Load Data

This is using the YAML.project in the configs folder. The idea here is to not have hard coded file paths baked into the R script. Adjust file paths and parameters there when necessary.The rest of the code should run without tweaks.
```{r Loading, include=FALSE}
#run YAML
cfg <- yaml::read_yaml(here("configs", "project.yml"))
cfg$paths   <- lapply(cfg$paths, here)
cfg$outputs <- lapply(cfg$outputs, here)
invisible(lapply(cfg$outputs, dir.create, recursive = TRUE, showWarnings = FALSE))

# Reading CSV files into data frames using YAML.
# Every column is read in as text first because the files don't agree on column types (e.g. Distance is a number in one file and text like "150+" in another), which breaks rbind. The columns we need as numbers get converted in the Formatting chunk.
# Each file is also tagged with its survey year. Year comes from the file rather than the SurveyDate column so a typo in a date (e.g. Cattus 2025 visit 1 is entered as 19-06-2026) doesn't move a survey into the wrong year.
read_point_counts <- function(path, year) {
  readr::read_csv(path, col_types = readr::cols(.default = readr::col_character())) %>%
    dplyr::rename(Point_ID = 'Point ID') %>%  #cleaning the name of the column Point ID to take out space (computers hate spaces)
    dplyr::mutate(Year = year)
}

cattus_2025_pc   <- read_point_counts(cfg$paths$cattus_2025_pc, 2025)
lhc_2025_pc      <- read_point_counts(cfg$paths$lhc_2025_pc,    2025)
cattus_2026_pc   <- read_point_counts(cfg$paths$cattus_2026_pc, 2026)
lhc_2026_pc      <- read_point_counts(cfg$paths$lhc_2026_pc,    2026)
site_metadata    <- readr::read_csv(cfg$paths$site_metadata,    show_col_types = FALSE)
habitat_lookup   <- readr::read_csv(cfg$paths$habitat_lookup,   show_col_types = FALSE)

#combining sites
#a;; will need to get adjusted when i get lhc 2023 data
point_counts_2025_raw <- rbind(cattus_2025_pc, lhc_2025_pc)
point_counts_2026_raw <- rbind(cattus_2026_pc, lhc_2026_pc)
point_counts_all_raw  <- rbind(cattus_2025_pc, lhc_2025_pc, cattus_2026_pc, lhc_2026_pc)
```

## Formatting

The same formatting steps get run on point_counts_2025, point_counts_2026, and point_counts_all, so they are written as functions once and then applied to each dataset. That way a fix only has to be made in one place.

```{r Formatting, include=FALSE}
# Species codes that showed up in the 2026 data but aren't in spp_lookup_table.csv yet. These are my best guess at the habitat category - CHECK THEM and then add them to the lookup table csv (anything already in the csv wins over this list, so once they're in the csv this list does nothing). GRFL and VOYE aren't standard 4-letter codes and are left as "Unclassified" until they're checked on the datasheet.
habitat_additions <- tibble::tribble(
  ~AlphaCode, ~Habitat,
  "AMGO", "Upland",        # American Goldfinch
  "BLSK", "Marsh",         # Black Skimmer
  "BTGR", "Marsh",         # Boat-tailed Grackle
  "COHA", "Upland",        # Cooper's Hawk
  "EABL", "Upland",        # Eastern Bluebird
  "EWPE", "Upland",        # Eastern Wood-Pewee
  "GRYE", "Marsh",         # Greater Yellowlegs
  "KIRA", "Marsh",         # King Rail
  "REVI", "Upland",        # Red-eyed Vireo
  "SEPL", "Marsh",         # Semipalmated Plover
  "SESA", "Marsh",         # Semipalmated Sandpiper
  "SOSA", "Marsh",         # Solitary Sandpiper
  "UNNH", "Marsh",         # unidentified night-heron
  "UNSA", "Marsh",         # unidentified sandpiper
  "WTSP", "Upland")        # White-throated Sparrow

habitat_lookup_full <- dplyr::bind_rows(
  habitat_lookup %>% dplyr::select(AlphaCode, Habitat),
  habitat_additions %>% dplyr::filter(!AlphaCode %in% habitat_lookup$AlphaCode))

# TRUE for species that were identified to species. Unidentified birds (UNSA, UNNH, "Sparrow Unidentified") still count toward total birds, but not toward species richness or diversity. The pattern is set in the YAML.
is_identified <- function(code) !grepl(cfg$params$unidentified_pattern, code)

# creating a datetime column and formatting times.
# NOTE: the 2025 datasheets write dates as day-month-year with dashes (30-06-2025) and the 2026 datasheets write them as month/day/year with slashes (6/28/2026). The separator is used to pick the right parser so both years read in correctly.
format_point_counts <- function(df) {
  df %>%
    dplyr::mutate(
      SurveyDate = dplyr::if_else(grepl("/", SurveyDate),
                                  lubridate::mdy(SurveyDate, quiet = TRUE),
                                  lubridate::dmy(SurveyDate, quiet = TRUE)),
      Datetime   = lubridate::parse_date_time(paste(SurveyDate, SurveyTime), orders = c("Ymd IMp", "Ymd HMS", "Ymd HM")),
      dplyr::across(c(TotalCount, VisitNum, TempF, Sky, WindSp, Noise), ~ suppressWarnings(as.numeric(.x))),
      # a missing distance band becomes "Unknown" so those birds aren't silently dropped
      DistBand   = dplyr::if_else(is.na(DistBand) | DistBand == "NA", "Unknown", DistBand),
      DistBand   = factor(DistBand, levels = c("0-50m", "51-100m", ">100m", "Unknown")),
      Identified = is_identified(AlphaCode)) %>%
    #adding in lat and long info for each line of code
    dplyr::left_join(
      site_metadata %>% dplyr::select(Point_ID, Lat, Long),
      by = "Point_ID") %>%
    #including habitat type (really rough categories. mostly for filtering out upland). Codes that aren't in the lookup get "Unclassified" so they're never silently dropped by a habitat filter
    dplyr::left_join(habitat_lookup_full, by = "AlphaCode") %>%
    dplyr::mutate(Habitat = tidyr::replace_na(Habitat, "Unclassified"))
}

point_counts_2025 <- format_point_counts(point_counts_2025_raw)
point_counts_2026 <- format_point_counts(point_counts_2026_raw)
point_counts_all  <- format_point_counts(point_counts_all_raw)

# Shannon Diversity Index (H')
shannon_diversity <- function(x) {
    p <- x / sum(x, na.rm = TRUE)
    p <- p[p > 0]  # Filter out zeroes
    -sum(p * log(p), na.rm = TRUE)}

# Calculates total birds, species richness, and Shannon diversity for whatever grouping is given (e.g. one survey, one distance band of one survey, one site-year).
# - duplicate rows of the same species are added together first (the old pivot_wider code kept only the first one)
# - rows with TotalCount = 0 (birds only noted outside the count period) don't count toward richness
# - unidentified birds count toward Total_Count but not richness or diversity
summarise_metrics <- function(df, ...) {
  df %>%
    dplyr::group_by(..., AlphaCode, Identified) %>%
    dplyr::summarise(Count = sum(TotalCount, na.rm = TRUE), .groups = "drop") %>%
    dplyr::group_by(...) %>%
    dplyr::summarise(
      Total_Count       = sum(Count),
      Species_Richness  = sum(Count > 0 & Identified),
      Shannon_Diversity = shannon_diversity(Count[Identified]),
      .groups = "drop")
}

# One row per survey (Point x Visit x Year) with all distance bands combined. This is the main unit for comparing points and years.
survey_cols <- c("Year", "Site", "Point_ID", "VisitNum", "SurveyDate", "Tide")

make_survey_metrics <- function(df) {
  summarise_metrics(df, Year, Site, Point_ID, VisitNum, SurveyDate, Tide) %>%
    dplyr::mutate(Year = factor(Year), VisitNum = factor(VisitNum))
}

# The "clean" wide datasets: one row per survey x distance band, one column per (identified) species, plus the metrics for that row. These feed the box plots, vegan and unmarked.
id_cols <- c("Point_ID", "Year", "Datetime", "DistBand", "Site", "Tide", "VisitNum")

clean_point_counts <- function(df) {
  wide <- df %>%
    dplyr::filter(Identified) %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(c(id_cols, "AlphaCode")))) %>%
    dplyr::summarise(Count = sum(TotalCount, na.rm = TRUE), .groups = "drop") %>%
    tidyr::pivot_wider(names_from = AlphaCode, values_from = Count, values_fill = 0)

  metrics <- summarise_metrics(df, Point_ID, Year, Datetime, DistBand, Site, Tide, VisitNum)

  dplyr::left_join(wide, metrics, by = id_cols) %>%
    dplyr::mutate(Site = factor(Site), VisitNum = factor(VisitNum), Point_ID = factor(Point_ID), Year = factor(Year))
}

point_counts_2025_clean <- clean_point_counts(point_counts_2025)
point_counts_2026_clean <- clean_point_counts(point_counts_2026)
point_counts_all_clean  <- clean_point_counts(point_counts_all)

# Same thing but only for marsh birds (everything that isn't Upland, same as the old Habitat != "Upland" filter)
marsh_2025_clean <- clean_point_counts(point_counts_2025 %>% filter(Habitat != "Upland"))
marsh_2026_clean <- clean_point_counts(point_counts_2026 %>% filter(Habitat != "Upland"))
marsh_all_clean  <- clean_point_counts(point_counts_all  %>% filter(Habitat != "Upland"))

survey_level_all <- make_survey_metrics(point_counts_all)

# Points surveyed in every year. LHC_07 was added in 2026, so it is left out of anything that compares years point-for-point.
shared_points <- point_counts_all %>%
  dplyr::distinct(Point_ID, Year) %>%
  dplyr::count(Point_ID, name = "n_years") %>%
  dplyr::filter(n_years == dplyr::n_distinct(point_counts_all$Year)) %>%
  dplyr::pull(Point_ID)

year_levels <- as.character(sort(unique(point_counts_all$Year)))
year_a <- year_levels[1]
year_b <- year_levels[length(year_levels)]
```

# Data QC

Checks on the raw datasheets before anything gets summarized. Each table lists the records that need a look; the last column of the summary says how the script handles them so nothing gets dropped silently. Fix what you can in the csv files and these tables will shrink.

```{r QC, echo=TRUE}
qc_all <- point_counts_all  # formatted data, all years

# 1) Survey effort: which points were surveyed on which dates. LHC_07 was added in 2026, so it only has 2026 dates.
effort_table <- qc_all %>%
  dplyr::distinct(Site, Point_ID, Year, VisitNum, SurveyDate) %>%
  dplyr::mutate(Survey = paste0(Year, " visit ", VisitNum)) %>%
  dplyr::select(Site, Point_ID, Survey, SurveyDate) %>%
  dplyr::mutate(SurveyDate = format(SurveyDate, "%d %b %Y")) %>%
  tidyr::pivot_wider(names_from = Survey, values_from = SurveyDate, values_fill = "not surveyed") %>%
  dplyr::arrange(Site, Point_ID)

kable(effort_table, caption = "Survey effort: survey date for each point and visit") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))

qc_all %>%
  dplyr::group_by(Site, Year) %>%
  dplyr::summarise(Points = dplyr::n_distinct(Point_ID),
                   Surveys = dplyr::n_distinct(paste(Point_ID, VisitNum)),
                   Records = dplyr::n(), .groups = "drop") %>%
  kable(caption = "Survey effort by site and year") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))

# 2) Survey dates that don't match the year of the file they came from
qc_date <- qc_all %>%
  dplyr::filter(lubridate::year(SurveyDate) != Year) %>%
  dplyr::distinct(Site, Point_ID, Year, VisitNum, SurveyDate)

# 3) Species codes that aren't in spp_lookup_table.csv
qc_lookup <- qc_all %>%
  dplyr::filter(!AlphaCode %in% habitat_lookup$AlphaCode) %>%
  dplyr::group_by(AlphaCode) %>%
  dplyr::summarise(Years = paste(sort(unique(Year)), collapse = ", "),
                   Records = dplyr::n(), Birds = sum(TotalCount, na.rm = TRUE), .groups = "drop") %>%
  dplyr::left_join(habitat_additions %>% dplyr::rename(`Habitat used (temporary)` = Habitat), by = "AlphaCode") %>%
  dplyr::mutate(`Habitat used (temporary)` = tidyr::replace_na(`Habitat used (temporary)`, "Unclassified"))

# 4) Unidentified birds
qc_unid <- qc_all %>%
  dplyr::filter(!Identified) %>%
  dplyr::select(Year, Point_ID, VisitNum, AlphaCode, TotalCount, Comments)

# 5) The same species entered more than once in the same distance band of the same survey (these get added together)
qc_dup <- qc_all %>%
  dplyr::count(Year, Point_ID, VisitNum, DistBand, AlphaCode, name = "Rows") %>%
  dplyr::filter(Rows > 1)

# 6) Rows with TotalCount = 0 or missing (usually birds only noted outside the count period)
qc_zero <- qc_all %>%
  dplyr::filter(is.na(TotalCount) | TotalCount == 0) %>%
  dplyr::select(Year, Point_ID, VisitNum, AlphaCode, TotalCount, Outside, Comments)

# 7) TotalCount doesn't equal the passive minutes + callback periods + outside columns
period_cols <- c("Min 1", "Min 2", "Min 3", "Min 4", "Min 5", "BLRA", "LEBI", "SORA", "VIRA", "KIRA", "CLRA", "AMBI", "COMO", "Outside")
qc_periods <- qc_all %>%
  dplyr::mutate(Period_Sum = rowSums(dplyr::across(dplyr::all_of(period_cols), ~ suppressWarnings(as.numeric(.x))), na.rm = TRUE)) %>%
  dplyr::filter(Period_Sum != TotalCount) %>%
  dplyr::select(Year, Point_ID, VisitNum, AlphaCode, TotalCount, Period_Sum, dplyr::all_of(period_cols))

# 8) Missing distance band
qc_band <- qc_all %>%
  dplyr::filter(DistBand == "Unknown") %>%
  dplyr::select(Year, Point_ID, VisitNum, AlphaCode, TotalCount)

# 9) Points with no coordinates in pointcount_site_metadata.csv
qc_coords <- qc_all %>%
  dplyr::filter(is.na(Lat) | is.na(Long)) %>%
  dplyr::distinct(Year, Point_ID)

# 10) More than one date/time/tide entered for a single survey
qc_survey <- qc_all %>%
  dplyr::distinct(Year, Point_ID, VisitNum, SurveyDate, SurveyTime, Tide) %>%
  dplyr::count(Year, Point_ID, VisitNum, name = "Versions") %>%
  dplyr::filter(Versions > 1)

qc_summary <- tibble::tribble(
  ~Check, ~Issues, ~`How the script handles it`,
  "Survey date doesn't match file year", nrow(qc_date), "Year is taken from the file, not the date",
  "Species code not in spp_lookup_table.csv", nrow(qc_lookup), "Temporary habitat from habitat_additions (Formatting chunk), otherwise 'Unclassified'",
  "Unidentified bird records", nrow(qc_unid), "Counted in total birds, left out of richness/diversity/accumulation",
  "Duplicate species rows in a survey x distance band", nrow(qc_dup), "Counts are added together",
  "TotalCount is 0 or missing", nrow(qc_zero), "Left out of counts and richness (still on the map popups)",
  "TotalCount differs from sum of count periods", nrow(qc_periods), "TotalCount is used as entered - check the datasheet",
  "Missing distance band", nrow(qc_band), "Kept as 'Unknown' band (in totals, not on ring maps)",
  "Point missing coordinates", nrow(qc_coords), "Point won't show on maps",
  "Survey with more than one date/time/tide", nrow(qc_survey), "Each version becomes its own row - fix the datasheet")

kable(qc_summary, caption = "QC summary") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed")) %>%
  row_spec(which(qc_summary$Issues > 0), bold = TRUE)

show_qc <- function(df, caption) {
  if (nrow(df) == 0) return(invisible(NULL))
  kable(df, caption = caption) %>%
    kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed")) %>%
    scroll_box(height = if (nrow(df) > 12) "350px" else NULL)
}

show_qc(qc_date,    "Survey dates that don't match the file year")
show_qc(qc_lookup,  "Species codes missing from spp_lookup_table.csv (add these to the csv)")
show_qc(qc_unid,    "Unidentified birds")
show_qc(qc_dup,     "Species entered more than once in the same survey x distance band")
show_qc(qc_zero,    "Rows with TotalCount 0 or missing")
show_qc(qc_periods, "TotalCount doesn't match the sum of the count periods")
show_qc(qc_band,    "Missing distance band")
show_qc(qc_coords,  "Points without coordinates")
show_qc(qc_survey,  "Surveys with more than one date/time/tide entered")
```

# Summaries

Broken down by site and year. "Per survey" values are averages over every Point x Visit survey, so they aren't inflated by LHC_07 being added in 2026. "Species detected" is a total for the site-year, which does go up with more surveys, so the shared-points column (points surveyed both years) is the fair year-to-year comparison.

```{r Summaries, echo=TRUE}
mean_se <- function(x) sprintf("%.1f ± %.1f", mean(x), sd(x) / sqrt(length(x)))

site_year_species <- function(df) {
  df %>%
    dplyr::filter(Identified, TotalCount > 0) %>%
    dplyr::group_by(Site, Year) %>%
    dplyr::summarise(spp = dplyr::n_distinct(AlphaCode), .groups = "drop")
}

summary_by_site_year <- survey_level_all %>%
  dplyr::mutate(Year = as.numeric(as.character(Year))) %>%
  dplyr::group_by(Site, Year) %>%
  dplyr::summarise(
    Points  = dplyr::n_distinct(Point_ID),
    Surveys = dplyr::n(),
    `Total birds` = sum(Total_Count),
    `Birds per survey` = mean_se(Total_Count),
    `Species per survey` = mean_se(Species_Richness),
    `Shannon per survey` = sprintf("%.2f ± %.2f", mean(Shannon_Diversity), sd(Shannon_Diversity) / sqrt(dplyr::n())),
    .groups = "drop") %>%
  dplyr::left_join(site_year_species(point_counts_all) %>% dplyr::rename(`Species detected (all points)` = spp), by = c("Site", "Year")) %>%
  dplyr::left_join(site_year_species(point_counts_all %>% dplyr::filter(Point_ID %in% shared_points)) %>%
                     dplyr::rename(`Species detected (shared points)` = spp), by = c("Site", "Year")) %>%
  dplyr::left_join(site_year_species(point_counts_all %>% dplyr::filter(Habitat == "Marsh")) %>%
                     dplyr::rename(`Marsh species detected` = spp), by = c("Site", "Year"))

study_area_species <- point_counts_all %>%
  dplyr::filter(Identified, TotalCount > 0) %>%
  dplyr::group_by(Year) %>%
  dplyr::summarise(`Species detected (both sites)` = dplyr::n_distinct(AlphaCode), .groups = "drop")

kable(summary_by_site_year, caption = "Point count summary by site and year (mean ± SE per survey)") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))

kable(study_area_species, caption = "Species detected across both sites") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))
```

The original vtable summaries, now split by site within each year.

```{r , echo=TRUE}
#sumamry table 2025, by site
st(point_counts_2025_clean %>% dplyr::select(c("Total_Count", "Species_Richness", "Shannon_Diversity", "DistBand", "Site")), group = "Site")

#sumamry table 2026, by site
st(point_counts_2026_clean %>% dplyr::select(c("Total_Count", "Species_Richness", "Shannon_Diversity", "DistBand", "Site")), group = "Site")

#sumamry table all years, by year (survey level so each survey counts once)
st(survey_level_all %>% dplyr::select(c("Total_Count", "Species_Richness", "Shannon_Diversity", "Site", "Year")), group = "Year")
```

### Species checklist

Every species detected, with the number of birds counted at each site in each year (all points). Highlighted rows were only detected in one of the two years at that site.

```{r Species Checklist, echo=TRUE}
checklist <- point_counts_all %>%
  dplyr::filter(TotalCount > 0) %>%
  dplyr::mutate(Column = paste(ifelse(Site == "Cattus Island", "CAT", "LHC"), Year)) %>%
  dplyr::group_by(Habitat, AlphaCode, Column) %>%
  dplyr::summarise(Birds = sum(TotalCount), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = Column, values_from = Birds, values_fill = 0) %>%
  dplyr::select(Habitat, AlphaCode, dplyr::any_of(c(paste("CAT", year_levels), paste("LHC", year_levels)))) %>%
  dplyr::arrange(Habitat, AlphaCode)

one_year_only <- which(
  (checklist[[paste("CAT", year_a)]] > 0) != (checklist[[paste("CAT", year_b)]] > 0) |
  (checklist[[paste("LHC", year_a)]] > 0) != (checklist[[paste("LHC", year_b)]] > 0))

kable(checklist, caption = "Birds counted per species, site and year") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("hover", "condensed")) %>%
  row_spec(one_year_only, background = "#FFF3CD") %>%
  scroll_box(height = "500px")
```

# Graphics

Adjust any graphic parameters in the YAML, aka height, width, DPI.

Each box is the spread of values across the distance bands of a point (one value per survey x distance band), split by visit. The survey-level versions further down are usually easier to read.

```{r Graphic Helpers, include=FALSE}
# shared look for all the box plots so they only need to be styled once
pc_boxplot <- function(data, y, title, ylab, by_year = FALSE) {
  p <- ggplot(data, aes(x = Point_ID, y = .data[[y]], color = VisitNum)) +
    geom_boxplot() +
    labs(title = title, x = "Point ID", y = ylab) +
    theme_publish(base_size = 18) +
    theme(panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text = element_text(face = "bold"),
      legend.position = "right")
  if (by_year) p + facet_grid(Year ~ Site, scales = "free_x") else p + facet_wrap(~ Site, scales = "free_x")
}

# observed species richness (identified species, count > 0) at each point, all visits and bands combined
point_richness <- function(df) {
  df %>%
    dplyr::filter(Identified, TotalCount > 0) %>%
    dplyr::group_by(Year, Site, Point_ID) %>%
    dplyr::summarise(richness = dplyr::n_distinct(AlphaCode), .groups = "drop")
}

save_fig <- function(plot, name, height = cfg$params$height, width = cfg$params$width) {
  ggplot2::ggsave(
    filename = file.path(cfg$outputs$figures_dir, name),
    plot = plot, height = height, width = width, dpi = cfg$params$dpi)
  plot
}
```

### 2025

```{r , echo=TRUE,fig.height=10, fig.width=15}
total_count <- pc_boxplot(point_counts_2025_clean, "Total_Count", "Total Bird Count by Site", "Total bird count")
save_fig(total_count, "total_count_point_counts.png")

marsh_total_count <- pc_boxplot(marsh_2025_clean, "Total_Count", "Total Marsh Bird Count by Site", "Total bird count")
save_fig(marsh_total_count, "marsh_spp_total_count_point_counts.png")

shannon <- pc_boxplot(point_counts_2025_clean, "Shannon_Diversity", "Shannon Diversity by Site", "Shannon Diversity")
save_fig(shannon, "shannon_diversity_point_counts.png")

richness <- pc_boxplot(point_counts_2025_clean, "Species_Richness", "Species Richness by Site", "Species Richness")
save_fig(richness, "species_richness_boxwhisker_point_counts.png")

bar_richness <- point_richness(point_counts_2025) %>%
  ggplot(aes(x = Point_ID, y = richness, fill = Point_ID)) +
  geom_col() +
  labs(y = "Observed Species Richness", x = "Location") +
  theme_publish() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none")
save_fig(bar_richness, "species_richness_bar_point_counts.png")
```

### 2026

```{r , echo=TRUE,fig.height=10, fig.width=15}
total_count_2026 <- pc_boxplot(point_counts_2026_clean, "Total_Count", "Total Bird Count by Site (2026)", "Total bird count")
save_fig(total_count_2026, "total_count_point_counts_2026.png")

marsh_total_count_2026 <- pc_boxplot(marsh_2026_clean, "Total_Count", "Total Marsh Bird Count by Site (2026)", "Total bird count")
save_fig(marsh_total_count_2026, "marsh_spp_total_count_point_counts_2026.png")

shannon_2026 <- pc_boxplot(point_counts_2026_clean, "Shannon_Diversity", "Shannon Diversity by Site (2026)", "Shannon Diversity")
save_fig(shannon_2026, "shannon_diversity_point_counts_2026.png")

richness_2026 <- pc_boxplot(point_counts_2026_clean, "Species_Richness", "Species Richness by Site (2026)", "Species Richness")
save_fig(richness_2026, "species_richness_boxwhisker_point_counts_2026.png")

bar_richness_2026 <- point_richness(point_counts_2026) %>%
  ggplot(aes(x = Point_ID, y = richness, fill = Point_ID)) +
  geom_col() +
  labs(y = "Observed Species Richness", x = "Location", title = "2026") +
  theme_publish() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none")
save_fig(bar_richness_2026, "species_richness_bar_point_counts_2026.png")
```

### All years (faceted by year)

Rows are years and columns are sites, so the same point can be compared straight up and down between years. LHC_07 was added in 2026 so it only shows up in the 2026 row.

```{r , echo=TRUE,fig.height=14, fig.width=15}
total_count_all <- pc_boxplot(point_counts_all_clean, "Total_Count", "Total Bird Count by Site and Year", "Total bird count", by_year = TRUE)
save_fig(total_count_all, "total_count_point_counts_all_years.png")

marsh_total_count_all <- pc_boxplot(marsh_all_clean, "Total_Count", "Total Marsh Bird Count by Site and Year", "Total bird count", by_year = TRUE)
save_fig(marsh_total_count_all, "marsh_spp_total_count_point_counts_all_years.png")

shannon_all <- pc_boxplot(point_counts_all_clean, "Shannon_Diversity", "Shannon Diversity by Site and Year", "Shannon Diversity", by_year = TRUE)
save_fig(shannon_all, "shannon_diversity_point_counts_all_years.png")

richness_all <- pc_boxplot(point_counts_all_clean, "Species_Richness", "Species Richness by Site and Year", "Species Richness", by_year = TRUE)
save_fig(richness_all, "species_richness_boxwhisker_point_counts_all_years.png")

# side-by-side bars so each point's two years sit next to each other
bar_richness_all <- point_richness(point_counts_all) %>%
  dplyr::mutate(Year = factor(Year)) %>%
  ggplot(aes(x = Point_ID, y = richness, fill = Year)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.8) +
  geom_text(aes(label = richness), position = position_dodge(width = 0.8), vjust = -0.4, size = 4) +
  facet_wrap(~ Site, scales = "free_x") +
  labs(y = "Observed Species Richness (both visits)", x = "Location", fill = "Year",
       title = "Species richness at each point by year") +
  theme_publish(base_size = 18) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold"),
    legend.position = "right")
save_fig(bar_richness_all, "species_richness_bar_point_counts_all_years.png")
```

### Species composition

The distance band filter uses `%in%` (the old `==` with two values compared row by row against alternating values and dropped about half the <100 m detections).

```{r Composition Helper, include=FALSE}
# percent of detections (rows) for each species within 100 m, by whatever grouping is given
composition <- function(df, ...) {
  df %>%
    filter(DistBand %in% c("0-50m", "51-100m"), TotalCount > 0) %>%
    dplyr::group_by(..., AlphaCode) %>%
    dplyr::summarise(Count = dplyr::n(), .groups = "drop") %>%
    dplyr::group_by(...) %>%
    dplyr::mutate(TotalCount = sum(Count), Percent = (Count / TotalCount) * 100) %>%
    dplyr::ungroup()
}

plot_composition <- function(df, title, by_year = FALSE) {
  # keeping any species that hit 1% in any panel, ordered by its combined percent so bars line up between years
  keep_spp <- df %>% filter(Percent >= 1) %>% dplyr::pull(AlphaCode) %>% unique()
  p <- df %>%
    filter(AlphaCode %in% keep_spp) %>%
    dplyr::mutate(AlphaCode = reorder(AlphaCode, Percent, FUN = sum)) %>%
    ggplot(aes(x = AlphaCode, y = Percent)) +
    geom_col(width = 0.8, alpha = 0.9) +
    coord_flip() +
    scale_y_continuous(labels = label_percent(scale = 1)) +
    labs(title = title, x = NULL, y = "Percent of total detections") +
    theme_publish(base_size = 18) +
    theme(panel.grid.minor = element_blank(), strip.text = element_text(face = "bold"))
  if (by_year) p + facet_wrap(~ Year) else p
}
```

```{r, echo=TRUE,fig.height=10, fig.width=15}
cattus_data_with_proportions <- composition(point_counts_2025 %>% filter(Site == "Cattus Island"), Site)
lhc_data_with_proportions    <- composition(point_counts_2025 %>% filter(Site == "Lighthouse Center"), Site)

cattus_spp_composition <- plot_composition(cattus_data_with_proportions, "Cattus Island Species Composition (≥ 1% of detections) <100 m")
save_fig(cattus_spp_composition, "cattus_spp_composition.png")

lhc_spp_composition <- plot_composition(lhc_data_with_proportions, "Lighthouse Center Species Composition (≥ 1% of detections) <100 m")
save_fig(lhc_spp_composition, "lhc_spp_composition.png")
```

```{r, echo=TRUE,fig.height=10, fig.width=15}
cattus_data_with_proportions_2026 <- composition(point_counts_2026 %>% filter(Site == "Cattus Island"), Site)
lhc_data_with_proportions_2026    <- composition(point_counts_2026 %>% filter(Site == "Lighthouse Center"), Site)

cattus_spp_composition_2026 <- plot_composition(cattus_data_with_proportions_2026, "Cattus Island Species Composition 2026 (≥ 1% of detections) <100 m")
save_fig(cattus_spp_composition_2026, "cattus_spp_composition_2026.png")

lhc_spp_composition_2026 <- plot_composition(lhc_data_with_proportions_2026, "Lighthouse Center Species Composition 2026 (≥ 1% of detections) <100 m")
save_fig(lhc_spp_composition_2026, "lhc_spp_composition_2026.png")
```

All years, faceted by year.

```{r, echo=TRUE,fig.height=12, fig.width=15}
cattus_spp_composition_all <- plot_composition(
  composition(point_counts_all %>% filter(Site == "Cattus Island"), Year, Site),
  "Cattus Island Species Composition by Year (≥ 1% of detections) <100 m", by_year = TRUE)
save_fig(cattus_spp_composition_all, "cattus_spp_composition_all_years.png")

lhc_spp_composition_all <- plot_composition(
  composition(point_counts_all %>% filter(Site == "Lighthouse Center"), Year, Site),
  "Lighthouse Center Species Composition by Year (≥ 1% of detections) <100 m", by_year = TRUE)
save_fig(lhc_spp_composition_all, "lhc_spp_composition_all_years.png")
```

# Point Count Maps

Adapting code from Serge Morand & Kittipong Chaisiri of HealthDEEP-CNRS, derived from https://rpubs.com/HealthDEEP/1377217

Displaying point count sites with buffers and levels of abundance. Each year is its own layer (toggle them in the layer control), so the map works like a facet by year. LHC_07 only appears in the 2026 layer.

```{r Map Helpers, include=FALSE}
#create a unique species dataframe so each dot represents the number of unique species
species_by_point <- function(df) {
  df %>%
    dplyr::filter(!is.na(AlphaCode)) %>%
    dplyr::group_by(Year, Site, Point_ID, AlphaCode) %>%
    dplyr::summarise(
      Lat  = dplyr::first(Lat),
      Long = dplyr::first(Long),
      .groups = "drop")
}

add_base_layers <- function(map, overlay_groups) {
  map %>%
    addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") %>%
    addProviderTiles(providers$Esri.WorldStreetMap,group ="WSM") %>%
    addTiles(group = "OSM") %>%
    addTiles(urlTemplate = "https://mts1.google.com/vt/lyrs=s&hl=en&src=app&x={x}&y={y}&z={z}&s=G",
             attribution = 'Google',group ="Google") %>%
    addLayersControl(baseGroups = c("Satellite", "WSM","OSM","Google"),
                     overlayGroups = overlay_groups,
                     options = layersControlOptions(collapsed = FALSE)) %>%
    addScaleBar(position = "bottomright",
                scaleBarOptions(maxWidth = 100, metric = TRUE, imperial = FALSE,
                                updateWhenIdle = TRUE))
}

species_cluster_map <- function(df) {
  pts <- species_by_point(df)
  layers <- as.character(sort(unique(pts$Year)))
  colors <- c("blue", "darkorange", "darkgreen", "purple")[seq_along(layers)]
  map <- leaflet()
  for (i in seq_along(layers)) {
    year_data <- pts %>% dplyr::filter(Year == layers[i])
    map <- map %>%
      addCircleMarkers(data = year_data,
                       label = ~as.character(AlphaCode),
                       popup = leafpop::popupTable(year_data),
                       fillColor = colors[i], weight = 0.5,
                       ~ Long,
                       ~ Lat,
                       group = layers[i],
                       clusterOptions = markerClusterOptions(),
                       stroke = FALSE, fillOpacity = 0.6)
  }
  add_base_layers(map, layers)
}
```

```{r , echo=TRUE}
species_cluster_map(point_counts_2025)
```

```{r , echo=TRUE}
species_cluster_map(point_counts_2026)
```

```{r , echo=TRUE}
species_cluster_map(point_counts_all)
```

## Distance band ring maps

Each point is drawn as three rings that match the distance bands on the datasheet: 0–50 m (inner), 50–100 m (middle), and 100–200 m (outer, the ">100m" band). Ring colour is the number of species detected in that band across both visits.

- The **2025** and **2026** layers use the same colour scale so they can be compared directly.
- The **Change** layer shows 2026 minus 2025 (blue = more species in 2026, red = fewer). LHC_07 is grey in the Change layer because it was only surveyed in 2026.
- Click a ring to see the species list for that band, or for the Change layer, which species were gained and lost.

Neighbouring points' outer rings overlap (as in the GIS buffers), so turn layers on one at a time.

```{r Ring Data, include=FALSE}
ring_radii <- setNames(unlist(cfg$params$distance_rings_m), c("0-50m", "51-100m", ">100m"))
ring_labels <- c("0-50m" = "0–50 m", "51-100m" = "50–100 m", ">100m" = "100–200 m")

# species detected in each distance band at each point and year (both visits combined). Bands with no birds get 0.
band_species <- point_counts_all %>%
  dplyr::filter(Identified, TotalCount > 0, DistBand %in% names(ring_radii)) %>%
  dplyr::group_by(Site, Point_ID, Year, DistBand, AlphaCode) %>%
  dplyr::summarise(Birds = sum(TotalCount), .groups = "drop")

point_years <- point_counts_all %>% dplyr::distinct(Site, Point_ID, Year, Lat, Long)

ring_data <- point_years %>%
  tidyr::crossing(DistBand = names(ring_radii)) %>%
  dplyr::mutate(DistBand = factor(DistBand, levels = levels(point_counts_all$DistBand))) %>%
  dplyr::left_join(
    band_species %>%
      dplyr::group_by(Point_ID, Year, DistBand) %>%
      dplyr::summarise(Richness = dplyr::n_distinct(AlphaCode),
                       Birds = sum(Birds),
                       Species = paste(sort(unique(AlphaCode)), collapse = ", "),
                       .groups = "drop"),
    by = c("Point_ID", "Year", "DistBand")) %>%
  dplyr::mutate(Richness = tidyr::replace_na(Richness, 0L),
                Birds = tidyr::replace_na(Birds, 0),
                Species = tidyr::replace_na(Species, "none"),
                Radius = ring_radii[as.character(DistBand)],
                Band = ring_labels[as.character(DistBand)])

# gained / lost species per point and band between the first and last year
band_lists <- band_species %>%
  dplyr::group_by(Point_ID, DistBand, Year) %>%
  dplyr::summarise(spp = list(unique(AlphaCode)), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = Year, values_from = spp, names_prefix = "Y")

ring_change <- ring_data %>%
  dplyr::select(Site, Point_ID, Year, DistBand, Band, Radius, Lat, Long, Richness) %>%
  tidyr::pivot_wider(names_from = Year, values_from = Richness, names_prefix = "R") %>%
  dplyr::mutate(Change = .data[[paste0("R", year_b)]] - .data[[paste0("R", year_a)]]) %>%
  dplyr::left_join(band_lists, by = c("Point_ID", "DistBand")) %>%
  dplyr::mutate(
    Gained = purrr::map2_chr(.data[[paste0("Y", year_b)]], .data[[paste0("Y", year_a)]],
                             ~ paste(sort(setdiff(unlist(.x), unlist(.y))), collapse = ", ")),
    Lost   = purrr::map2_chr(.data[[paste0("Y", year_a)]], .data[[paste0("Y", year_b)]],
                             ~ paste(sort(setdiff(unlist(.x), unlist(.y))), collapse = ", ")))
```

```{r Ring Map, echo=TRUE}
rich_pal   <- colorNumeric("YlGnBu", domain = c(0, max(ring_data$Richness)))
change_max <- max(abs(ring_change$Change), na.rm = TRUE)
change_pal <- colorNumeric("RdBu", domain = c(-change_max, change_max), na.color = "#BDBDBD")

ring_map <- leaflet()

# draw the outer rings first so the smaller ones sit on top
for (band in rev(names(ring_radii))) {
  for (yr in year_levels) {
    rd <- ring_data %>% dplyr::filter(Year == yr, DistBand == band)
    ring_map <- ring_map %>%
      addCircles(data = rd, lng = ~Long, lat = ~Lat, radius = ~Radius,
                 fillColor = ~rich_pal(Richness), fillOpacity = 0.75,
                 color = "white", weight = 1, group = yr,
                 label = ~paste0(Point_ID, " ", Band, ": ", Richness, " species"),
                 popup = ~paste0("<b>", Point_ID, " &ndash; ", Band, " &ndash; ", Year, "</b><br>",
                                 Richness, " species, ", Birds, " birds<br>", Species))
  }
  cd <- ring_change %>% dplyr::filter(DistBand == band)
  ring_map <- ring_map %>%
    addCircles(data = cd, lng = ~Long, lat = ~Lat, radius = ~Radius,
               fillColor = ~change_pal(Change), fillOpacity = 0.8,
               color = "white", weight = 1, group = "Change",
               label = ~paste0(Point_ID, " ", Band, ": ",
                               ifelse(is.na(Change), paste("only surveyed in", year_b), sprintf("%+d species", Change))),
               popup = ~paste0("<b>", Point_ID, " &ndash; ", Band, " &ndash; change ", year_a, " to ", year_b, "</b><br>",
                               ifelse(is.na(Change), paste("Only surveyed in", year_b),
                                      paste0(sprintf("%+d species", Change),
                                             "<br><b>Gained:</b> ", ifelse(Gained == "", "none", Gained),
                                             "<br><b>Lost:</b> ", ifelse(Lost == "", "none", Lost)))))
}

ring_map %>%
  addLabelOnlyMarkers(data = point_years %>% dplyr::distinct(Point_ID, Lat, Long),
                      lng = ~Long, lat = ~Lat, label = ~Point_ID,
                      labelOptions = labelOptions(noHide = TRUE, direction = "center", textOnly = TRUE,
                                                  style = list("font-weight" = "bold", "color" = "black"))) %>%
  addLegend("bottomleft", pal = rich_pal, values = c(0, max(ring_data$Richness)),
            title = "Species in band") %>%
  addLegend("bottomright", pal = change_pal, values = c(-change_max, change_max),
            title = paste("Change in species<br>", year_a, "to", year_b), group = "Change") %>%
  add_base_layers(c(year_levels, "Change")) %>%
  hideGroup(c(year_levels[-1], "Change"))
```

### Ring diagrams (static, for reports)

The same rings drawn as one "target" per point so nothing overlaps. Numbers are species detected in each band. Rings are drawn to scale (0–50, 50–100, 100–200 m), which is why the outer band is so much bigger: it covers 75% of the 200 m circle.

```{r Ring Diagram Helpers, include=FALSE}
# builds a filled ring (annulus) as a polygon with a hole, centred on 0,0, in metres
annulus <- function(r_in, r_out, n = 120) {
  t <- seq(0, 2 * pi, length.out = n)
  outer <- data.frame(x = r_out * cos(t), y = r_out * sin(t), sub = 1)
  inner <- if (r_in > 0) data.frame(x = r_in * cos(rev(t)), y = r_in * sin(rev(t)), sub = 2) else NULL
  rbind(outer, inner)
}

ring_inner <- setNames(c(0, head(unlist(cfg$params$distance_rings_m), -1)), names(ring_radii))

# one ring polygon per row of df (one row = one point x band), with the row's data attached
ring_polys <- function(df) {
  df$.id <- seq_len(nrow(df))
  polys <- purrr::map_dfr(df$.id, function(i) {
    band <- as.character(df$DistBand[i])
    cbind(.id = i, annulus(ring_inner[band], ring_radii[band]))
  })
  dplyr::left_join(polys, df, by = ".id")
}

ring_label_pos <- function(df) {
  df %>% dplyr::mutate(
    y = ifelse(DistBand == "0-50m", 0, (ring_inner[as.character(DistBand)] + ring_radii[as.character(DistBand)]) / 2),
    x = 0)
}

ring_theme <- theme_void(base_size = 16) +
  theme(strip.text = element_text(face = "bold", margin = margin(4, 4, 4, 4)),
        plot.title = element_text(face = "bold"),
        legend.position = "right")
```

```{r Ring Diagrams, echo=TRUE, fig.height=9, fig.width=16}
for (site in sort(unique(ring_data$Site))) {
  site_rings <- ring_data %>% dplyr::filter(Site == site)

  richness_rings <- ggplot(ring_polys(site_rings), aes(x, y)) +
    geom_polygon(aes(group = interaction(.id), subgroup = sub, fill = Richness), color = "white", linewidth = 0.4) +
    geom_label(data = ring_label_pos(site_rings), aes(label = Richness), size = 4, label.size = 0, alpha = 0.7) +
    facet_grid(Year ~ Point_ID, switch = "y") +
    coord_equal() +
    scale_fill_distiller(palette = "YlGnBu", direction = 1, limits = c(0, max(ring_data$Richness))) +
    labs(title = paste(site, "– species detected in each distance band"),
         subtitle = "Inner = 0–50 m, middle = 50–100 m, outer = 100–200 m",
         fill = "Species") +
    ring_theme
  print(richness_rings)
  save_fig(richness_rings, paste0(gsub(" ", "_", tolower(site)), "_distance_band_rings.png"), height = 7, width = 16)

  site_change <- ring_change %>% dplyr::filter(Site == site)
  change_rings <- ggplot(ring_polys(site_change), aes(x, y)) +
    geom_polygon(aes(group = interaction(.id), subgroup = sub, fill = Change), color = "white", linewidth = 0.4) +
    geom_label(data = ring_label_pos(site_change),
               aes(label = ifelse(is.na(Change), "new", sprintf("%+d", Change))), size = 4, label.size = 0, alpha = 0.7) +
    facet_wrap(~ Point_ID, nrow = 1) +
    coord_equal() +
    scale_fill_distiller(palette = "RdBu", direction = 1, limits = c(-change_max, change_max), na.value = "grey80") +
    labs(title = paste(site, "– change in species per distance band,", year_a, "to", year_b),
         subtitle = "Blue = more species in the later year, red = fewer",
         fill = "Change") +
    ring_theme
  print(change_rings)
  save_fig(change_rings, paste0(gsub(" ", "_", tolower(site)), "_distance_band_ring_change.png"), height = 4.5, width = 16)
}
```

Focal species rings: the same idea for a single species (set `focal_species` in the YAML). Numbers are birds counted in each band (both visits combined).

```{r Focal Species Rings, echo=TRUE, fig.height=9, fig.width=16}
focal_rings <- point_years %>%
  tidyr::crossing(DistBand = names(ring_radii)) %>%
  dplyr::mutate(DistBand = factor(DistBand, levels = levels(point_counts_all$DistBand))) %>%
  dplyr::left_join(
    band_species %>% dplyr::filter(AlphaCode == cfg$params$focal_species) %>% dplyr::select(Point_ID, Year, DistBand, Birds),
    by = c("Point_ID", "Year", "DistBand")) %>%
  dplyr::mutate(Birds = tidyr::replace_na(Birds, 0))

focal_ring_plot <- ggplot(ring_polys(focal_rings), aes(x, y)) +
  geom_polygon(aes(group = interaction(.id), subgroup = sub, fill = Birds), color = "white", linewidth = 0.4) +
  geom_label(data = ring_label_pos(focal_rings), aes(label = Birds), size = 3.5, label.size = 0, alpha = 0.7) +
  facet_grid(Year ~ Point_ID, switch = "y") +
  coord_equal() +
  scale_fill_distiller(palette = "Greens", direction = 1) +
  labs(title = paste(cfg$params$focal_species, "– birds counted in each distance band"),
       subtitle = "Inner = 0–50 m, middle = 50–100 m, outer = 100–200 m", fill = "Birds") +
  ring_theme
save_fig(focal_ring_plot, paste0("focal_species_rings_", cfg$params$focal_species, ".png"), height = 7, width = 22)
```

# Species Pool and Sampling Sufficiency

This code is derived from the above author and uses the Vegan r package. It generates a community matrix to estimates the total species pool for the study area, accounting for imperfect detection and undersampling.

The chaos estimator is the true species richness by looking at rare species and assuming how many rare species may have been missed.For my example, my total n value was 71 species, but based on how many species were only detected once or twice, the true species pool is likely around 98 species.This has a standard error of ~17.5 species, which is substantial uncertainty. Based on the result, I likely missed 25-30 species. However, this is a calculation rather than a fact.

The first order jackknife tells us a similar metric from the chaos, using species detected in only one sample to estimate the unseen species using a more conservative approach. In my example, it brought the true total down to 89 with a standard error of around 5. This means I undersampled roughly 18 species.the second order jackknife uses species not just seen in one sample, but also two. So this is less conservative and brought it closer to 100 species.

The boorstrap is a resampling estimator and is the lowest and most conservative, suggesting at least 79 species were present. This means only 8 species were missed.

Across the point count surveys, I observed 71 bird species. However, multiple species pool estimators suggest that true species richness is substantially higher (approximately 80–100 species), indicating incomplete sampling and the presence of undetected or rarely detected species. This pattern is consistent with point count surveys, where detectability varies among species due to behavior, abundance, and habitat use.This leaves room for improvement with acoustic data.

(Those numbers were written for the 2025 data before the QC fixes; unidentified birds and outside-the-count records are now left out, so the tables below will differ slightly.)

```{r Community Matrices, include=FALSE}
# i need to collapse data into a the matrix to create the further calculations. Because of the nature of the datasheet, we are indicating the separate distance bands of birds simultaneously while conducting each point count. The below code will collapse the distance bands because I am not seeing these as separate surveys, but rather one survey accounting for data differently. If I want to consider each distance band a unique survey in the future, I should take that out.
make_comm <- function(clean_df) {
  surveys <- clean_df %>%
    dplyr::select(-dplyr::any_of(c("Total_Count", "Species_Richness", "Shannon_Diversity"))) %>%
    group_by(Point_ID, VisitNum, Year, Site) %>%
    summarise_if(is.numeric, sum, na.rm = TRUE) %>%
    ungroup()
  list(info = surveys %>% dplyr::select(Point_ID, VisitNum, Year, Site),
       matrix = surveys %>% dplyr::select(-Point_ID, -VisitNum, -Year, -Site) %>% mutate_all(~ replace_na(.x, 0)))
}

comm_2025 <- make_comm(point_counts_2025_clean)
comm_2026 <- make_comm(point_counts_2026_clean)
comm_all  <- make_comm(point_counts_all_clean)

comm_matrix      <- comm_2025$matrix
comm_matrix_2026 <- comm_2026$matrix
comm_matrix_all  <- comm_all$matrix
```

```{r, echo=TRUE}
pool <- specpool(comm_matrix)
kable(pool, caption = "Species Pool Estimators (2025)") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))

pool_2026 <- specpool(comm_matrix_2026)
kable(pool_2026, caption = "Species Pool Estimators (2026)") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))

# all years: both years pooled, then split by year and by site x year using the pool argument
pool_all <- specpool(comm_matrix_all)
kable(pool_all, caption = "Species Pool Estimators (all years combined)") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))

pool_by_site_year <- specpool(comm_matrix_all, pool = paste(comm_all$info$Site, comm_all$info$Year))
kable(pool_by_site_year, digits = 1, caption = "Species Pool Estimators by Site and Year") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

This code was also derived from the above author using the Vegan r package. This chunk of code creates a species accumulation curve for the point count data. This curve will tell us at what number of point counts would reach a species plateau based on the matric provided. If the curve has not plateaued, sampling is incomplete. If it is still rising, this means that at this number of surveys, additional surveys would still bring new species. Shallow slopes result in diminishing returns and steep slopes mean that the area is undersampled.

The x axis in my case were actually including 12 sites with 24 total surveys (two surveys at each site within the breeding season). Each visit is being treated as an independent sample.It is normally titled "locations" with one survey at each site being the standard, but I resampled due to the SHARP protocol.

My species area curve is steep at first and then begins to gentle, but doesn't fully plateau. The most common species were detected early in sampling, but additional sampling would continue to add new (rare) species.This can be where ARUs are used to reach that plateau. I am interpreting this to mean that if point counts are done, then when doing species detections with acoustic data, I can just focus on the rare species to determine presence or density through acoustic means.Point count sampling captured most but not all species present in the system. Remaining undetected species are likely rare, cryptic, or transient.

The accumulation curve approaching but not reaching an asymptote reinforces the conclusion that true richness exceeds observed richness, with undetected species remaining with an observed richness at 71 spp, and an estimated pool of 80-100 species.

```{r, echo=TRUE,fig.height=10, fig.width=15}
spec_accum <- specaccum(comm_matrix, method = "random")

plot(spec_accum, ci.type = "poly", col = "blue", lwd = 2,
     ci.lty = 0, ci.col = "lightblue",
     xlab = "Number of Surveys", ylab = "Species Richness", main = "2025")

spec_accum_2026 <- specaccum(comm_matrix_2026, method = "random")

plot(spec_accum_2026, ci.type = "poly", col = "blue", lwd = 2,
     ci.lty = 0, ci.col = "lightblue",
     xlab = "Number of Surveys", ylab = "Species Richness", main = "2026")
```

## Was sampling effort sufficient?

The accumulation curves above only show the surveys we have. To judge whether that was enough, the curves below are broken out by **site and year**, and each one is **extrapolated** (dashed) to twice the number of surveys using the incidence-based method from Chao et al. (2014), the same method the iNEXT package uses. A survey = one visit to one point (all distance bands combined), and only identified species are used.

How to read the table:

- **Chao2**: estimated total number of species present (seen + missed), based on how many species were seen in only one survey (Q1, "uniques") or two surveys (Q2, "duplicates"). Lots of uniques and few duplicates means many species are still being missed.
- **Inventory completeness**: observed species ÷ Chao2. The share of the estimated species pool we actually recorded.
- **Sample coverage**: the chance that a bird detection on the *next* survey would be a species we already have. This is the better measure of whether effort was "enough": coverage above ~90% means the common and moderately common species are well sampled, and what's left is rare, cryptic, or passing-through species. It doesn't depend on the (uncertain) Chao2 estimate.
- **Surveys needed**: how many *additional* point visits it would take to record 90% / 95% of the estimated pool. Divide by the number of points to get the number of additional full rounds of visits.

"Both years" pools all surveys at a site, i.e. the species inventory after two seasons.

```{r Sufficiency Functions, include=FALSE}
# incidence frequencies: for each identified species, the number of surveys it was detected in
incidence <- function(df) {
  surveys <- df %>% dplyr::distinct(Year, Point_ID, VisitNum)
  Y <- df %>%
    dplyr::filter(Identified, TotalCount > 0) %>%
    dplyr::distinct(Year, Point_ID, VisitNum, AlphaCode) %>%
    dplyr::count(AlphaCode, name = "Y") %>%
    dplyr::pull(Y)
  list(Y = Y, T = nrow(surveys), n_points = dplyr::n_distinct(surveys$Point_ID))
}

# Chao2, sample coverage and the extrapolation formula (Chao et al. 2014, Ecological Monographs; Colwell et al. 2012)
sufficiency_stats <- function(inc, targets = unlist(cfg$params$sufficiency_targets)) {
  Y <- inc$Y; T <- inc$T
  Sobs <- length(Y); Q1 <- sum(Y == 1); Q2 <- sum(Y == 2); U <- sum(Y)
  Q0 <- if (Q2 > 0) ((T - 1) / T) * Q1^2 / (2 * Q2) else ((T - 1) / T) * Q1 * (Q1 - 1) / 2
  chao2 <- Sobs + Q0
  coverage <- if (Q2 > 0) 1 - (Q1 / U) * ((T - 1) * Q1 / ((T - 1) * Q1 + 2 * Q2)) else
                          1 - (Q1 / U) * ((T - 1) * (Q1 - 1) / ((T - 1) * (Q1 - 1) + 2))
  surveys_needed <- sapply(targets, function(g) {
    if (Q0 <= 0 || g * chao2 <= Sobs) return(0)
    ceiling(log(1 - (g * chao2 - Sobs) / Q0) / log(1 - Q1 / (T * Q0 + Q1)))
  })
  out <- data.frame(Surveys = T, Points = inc$n_points, Observed = Sobs, Uniques = Q1, Duplicates = Q2,
                    Chao2 = chao2, Completeness = Sobs / chao2, Coverage = coverage)
  for (i in seq_along(targets)) out[[paste0("Surveys_to_", targets[i] * 100, "pct")]] <- surveys_needed[i]
  out
}

# expected species for t surveys: interpolation (rarefaction) up to T, extrapolation beyond T
accumulation_curve <- function(inc, max_t = 2 * inc$T) {
  Y <- inc$Y; T <- inc$T
  Sobs <- length(Y); Q1 <- sum(Y == 1); Q2 <- sum(Y == 2)
  Q0 <- if (Q2 > 0) ((T - 1) / T) * Q1^2 / (2 * Q2) else ((T - 1) / T) * Q1 * (Q1 - 1) / 2
  t <- seq_len(max_t)
  S <- sapply(t, function(tt) {
    if (tt <= T) sum(1 - exp(lchoose(T - Y, tt) - lchoose(T, tt)))
    else if (Q0 == 0) Sobs
    else Sobs + Q0 * (1 - (1 - Q1 / (T * Q0 + Q1))^(tt - T))
  })
  data.frame(Surveys = t, Species = S, Method = ifelse(t <= T, "Observed (rarefied)", "Extrapolated"))
}

# all the groups to run: each site x year, each site with both years pooled, and both sites together
sufficiency_groups <- function(df) {
  sites <- c(sort(unique(df$Site)), "Both sites")
  yrs   <- c(year_levels, "Both years")
  grid <- expand.grid(Site = sites, Year = yrs, stringsAsFactors = FALSE)
  purrr::pmap(grid, function(Site, Year) {
    d <- df
    if (Site != "Both sites") d <- d[d$Site == Site, ]
    if (Year != "Both years") d <- d[d$Year == as.numeric(Year), ]
    list(Site = Site, Year = Year, inc = incidence(d))
  })
}
```

```{r Sufficiency, echo=TRUE, fig.height=8, fig.width=16}
groups <- sufficiency_groups(point_counts_all)

sufficiency_table <- purrr::map_dfr(groups, function(g) cbind(Site = g$Site, Year = g$Year, sufficiency_stats(g$inc)))

curves <- purrr::map_dfr(groups, function(g) {
  st <- sufficiency_stats(g$inc)
  cbind(Site = g$Site, Year = g$Year, Chao2 = st$Chao2, accumulation_curve(g$inc))
}) %>%
  dplyr::mutate(Site = factor(Site, levels = unique(sufficiency_table$Site)),
                Year = factor(Year, levels = c(year_levels, "Both years")))

observed_pts <- curves %>%
  dplyr::group_by(Site, Year) %>%
  dplyr::filter(Method == "Observed (rarefied)") %>%
  dplyr::slice_max(Surveys, n = 1)

accumulation_extrap <- ggplot(curves, aes(x = Surveys, y = Species, color = Year)) +
  geom_line(data = dplyr::filter(curves, Method == "Observed (rarefied)"), linewidth = 1.3) +
  geom_line(data = dplyr::filter(curves, Method == "Extrapolated"), linewidth = 1, linetype = "dashed") +
  geom_point(data = observed_pts, size = 3.5) +
  geom_hline(data = dplyr::distinct(curves, Site, Year, Chao2), aes(yintercept = Chao2, color = Year),
             linetype = "dotted", linewidth = 0.8) +
  facet_wrap(~ Site, scales = "free_x") +
  labs(title = "Species accumulation with extrapolation to twice the effort",
       subtitle = "Solid = surveys done, dot = actual effort, dashed = extrapolated, dotted = Chao2 estimate of total species",
       x = "Number of surveys (point visits)", y = "Species", color = NULL) +
  theme_publish(base_size = 18) +
  theme(panel.grid.minor = element_blank(), strip.text = element_text(face = "bold"), legend.position = "bottom")
save_fig(accumulation_extrap, "species_accumulation_extrapolated.png")

sufficiency_table %>%
  dplyr::mutate(Chao2 = round(Chao2, 1),
                Completeness = scales::percent(Completeness, accuracy = 1),
                Coverage = scales::percent(Coverage, accuracy = 1)) %>%
  kable(caption = "Sampling sufficiency by site and year",
        col.names = c("Site", "Year", "Surveys", "Points", "Species observed", "Uniques (Q1)", "Duplicates (Q2)",
                      "Chao2 estimate", "Inventory completeness", "Sample coverage",
                      paste0("Extra surveys for ", unlist(cfg$params$sufficiency_targets) * 100, "% of pool"))) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed")) %>%
  collapse_rows(columns = 1, valign = "top")
```

```{r Sufficiency Text, echo=FALSE, results="asis"}
# writes a plain-language sentence for each site x year so the interpretation updates with the data
target_cols <- grep("^Surveys_to_", names(sufficiency_table), value = TRUE)
for (i in seq_len(nrow(sufficiency_table))) {
  r <- sufficiency_table[i, ]
  rounds <- r[[target_cols[1]]] / r$Points
  verdict <- if (r$Coverage >= 0.95) "effort was sufficient for all but the rarest species" else
             if (r$Coverage >= 0.90) "effort was sufficient for common and moderately common species, but rare species are still being missed" else
             "effort was not yet sufficient; each new survey is still adding species regularly"
  cat(sprintf("- **%s, %s**: %d surveys recorded %d species out of an estimated %.0f (%.0f%% complete). Sample coverage was %.0f%%, so %s. Reaching %d%% of the estimated pool would take about %d more point visits (≈ %.1f more rounds of visits to all %d points).\n",
              r$Site, r$Year, r$Surveys, r$Observed, r$Chao2, 100 * r$Completeness, 100 * r$Coverage, verdict,
              round(100 * unlist(cfg$params$sufficiency_targets)[1]), r[[target_cols[1]]], rounds, r$Points))
}
```

Things to keep in mind: Chao2 is a *minimum* estimate and gets unstable when there are few duplicates (Q2), and the number of surveys needed goes up quickly for the last few rare species, so 95% targets can look very large. Species that are only passing through (flyovers, migrants late in the window) inflate the uniques; restricting to marsh species or to detections within 100 m would give a tighter estimate for the species the monitoring is really about. The gap between observed and estimated species is where ARU recordings can add the most.

### Sufficiency for marsh species only

Same analysis restricted to marsh-associated species (Habitat = Marsh in the lookup table), since these are the focus of the monitoring.

```{r Sufficiency Marsh, echo=TRUE}
marsh_groups <- sufficiency_groups(point_counts_all %>% dplyr::filter(Habitat == "Marsh"))

purrr::map_dfr(marsh_groups, function(g) cbind(Site = g$Site, Year = g$Year, sufficiency_stats(g$inc))) %>%
  dplyr::mutate(Chao2 = round(Chao2, 1),
                Completeness = scales::percent(Completeness, accuracy = 1),
                Coverage = scales::percent(Coverage, accuracy = 1)) %>%
  kable(caption = "Sampling sufficiency for marsh species",
        col.names = c("Site", "Year", "Surveys", "Points", "Species observed", "Uniques (Q1)", "Duplicates (Q2)",
                      "Chao2 estimate", "Inventory completeness", "Sample coverage",
                      paste0("Extra surveys for ", unlist(cfg$params$sufficiency_targets) * 100, "% of pool"))) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed")) %>%
  collapse_rows(columns = 1, valign = "top")
```

# Year-to-Year Comparison

With only two years of data there is no real "trend" to fit (any two points make a straight line), so this section focuses on what two years can actually tell us, starting with the simplest views.

Year comparisons use **only the points surveyed in both years** (LHC_07 was added in 2026), so a difference isn't just the extra point. LHC_07 is still included everywhere else (summaries, maps, rings, 2026 figures). Caveats: survey dates, tides and weather differed between years, and one year-to-year difference can't separate a real population change from normal year-to-year variation. Treat these as a baseline for future years. Once a third year is added, the paired tests can be swapped for mixed models with Point_ID as a random effect (e.g. `lme4::glmer(Total_Count ~ Year + VisitNum + (1 | Point_ID), family = poisson)`).

```{r Year Comparison Data, include=FALSE}
shared_long   <- point_counts_all %>% dplyr::filter(Point_ID %in% shared_points)
survey_shared <- survey_level_all %>% dplyr::filter(Point_ID %in% shared_points)
marsh_survey_shared <- make_survey_metrics(shared_long %>% dplyr::filter(Habitat == "Marsh"))
focal_survey_shared <- shared_long %>%
  dplyr::distinct(Year, Site, Point_ID, VisitNum) %>%
  dplyr::left_join(shared_long %>%
                     dplyr::filter(AlphaCode == cfg$params$focal_species) %>%
                     dplyr::group_by(Year, Site, Point_ID, VisitNum) %>%
                     dplyr::summarise(Birds = sum(TotalCount, na.rm = TRUE), .groups = "drop"),
                   by = c("Year", "Site", "Point_ID", "VisitNum")) %>%
  dplyr::mutate(Birds = tidyr::replace_na(Birds, 0), Year = factor(Year))

survey_level_long <- survey_shared %>%
  tidyr::pivot_longer(c(Total_Count, Species_Richness, Shannon_Diversity),
                      names_to = "Metric", values_to = "Value") %>%
  dplyr::mutate(Metric = factor(Metric,
                                levels = c("Total_Count", "Species_Richness", "Shannon_Diversity"),
                                labels = c("Birds per survey", "Species per survey", "Shannon diversity")),
                # putting every survey on the same dummy year so the years line up on one calendar axis
                Season_Date = as.Date(paste0("2000-", format(SurveyDate, "%m-%d"))))

# species presence per site and year (shared points), used for turnover
spp_presence <- shared_long %>%
  dplyr::filter(Identified, TotalCount > 0) %>%
  dplyr::distinct(Site, Year, AlphaCode) %>%
  dplyr::bind_rows(dplyr::mutate(., Site = "Both sites") %>% dplyr::distinct()) %>%
  dplyr::mutate(Year = as.character(Year), present = TRUE) %>%
  tidyr::pivot_wider(names_from = Year, values_from = present, values_fill = FALSE) %>%
  dplyr::rename(in_a = dplyr::all_of(year_a), in_b = dplyr::all_of(year_b))
```

## Scorecard

One table with the headline numbers for each site, side by side by year. Arrows show the direction of change. Everything is per survey or on shared points, so the years are directly comparable.

```{r Scorecard, echo=TRUE}
add_both <- function(df) dplyr::bind_rows(df, dplyr::mutate(df, Site = "Both sites"))

score_means <- function(df, value, label) {
  add_both(df) %>%
    dplyr::group_by(Site, Year = as.character(Year)) %>%
    dplyr::summarise(Value = mean({{ value }}), .groups = "drop") %>%
    dplyr::mutate(Metric = label)
}

scorecard <- dplyr::bind_rows(
  add_both(point_counts_all) %>%
    dplyr::group_by(Site, Year = as.character(Year)) %>%
    dplyr::summarise(Value = dplyr::n_distinct(Point_ID), .groups = "drop") %>%
    dplyr::mutate(Metric = "Points surveyed (all)"),
  spp_presence %>%
    tidyr::pivot_longer(c(in_a, in_b), names_to = "Year", values_to = "present") %>%
    dplyr::mutate(Year = ifelse(Year == "in_a", year_a, year_b)) %>%
    dplyr::group_by(Site, Year) %>%
    dplyr::summarise(Value = sum(present), .groups = "drop") %>%
    dplyr::mutate(Metric = "Species detected"),
  score_means(survey_shared, Species_Richness, "Species per survey"),
  score_means(survey_shared, Total_Count, "Birds per survey"),
  score_means(survey_shared, Shannon_Diversity, "Shannon diversity per survey"),
  score_means(marsh_survey_shared, Species_Richness, "Marsh species per survey"),
  score_means(marsh_survey_shared, Total_Count, "Marsh birds per survey"),
  score_means(focal_survey_shared, Birds, paste(cfg$params$focal_species, "per survey"))) %>%
  dplyr::mutate(Site = factor(Site, levels = c(sort(unique(point_counts_all$Site)), "Both sites")),
                Metric = factor(Metric, levels = unique(Metric))) %>%
  tidyr::pivot_wider(names_from = Year, values_from = Value) %>%
  dplyr::mutate(
    a = .data[[year_a]], b = .data[[year_b]],
    Change = b - a,
    `% change` = ifelse(a == 0, NA, 100 * (b - a) / a),
    ` ` = dplyr::case_when(abs(`% change`) < 5 | is.na(`% change`) ~ "≈", Change > 0 ~ "▲", TRUE ~ "▼")) %>%
  dplyr::arrange(Site, Metric)

scorecard %>%
  dplyr::transmute(Site, Metric,
                   !!year_a := round(a, 2), !!year_b := round(b, 2),
                   Change = sprintf("%+.2f", Change),
                   `% change` = ifelse(is.na(`% change`), "", sprintf("%+.0f%%", `% change`)), ` `) %>%
  kable(caption = paste0("Scorecard, ", year_a, " vs ", year_b, " (points surveyed both years; 'Points surveyed' counts all points)"), align = "llrrrrc") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed")) %>%
  column_spec(7, color = ifelse(scorecard$` ` == "▲", "#1B7837", ifelse(scorecard$` ` == "▼", "#B2182B", "grey50")), bold = TRUE) %>%
  collapse_rows(columns = 1, valign = "top")
```

## Year side by side

The simplest comparison: average per survey with standard error bars, for each site and year.

```{r Year Bars, echo=TRUE, fig.height=7, fig.width=15}
year_bars <- survey_level_long %>%
  dplyr::group_by(Site, Year, Metric) %>%
  dplyr::summarise(Mean = mean(Value), SE = sd(Value) / sqrt(dplyr::n()), .groups = "drop") %>%
  ggplot(aes(x = Site, y = Mean, fill = Year)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.75) +
  geom_errorbar(aes(ymin = Mean - SE, ymax = Mean + SE), position = position_dodge(width = 0.8), width = 0.25) +
  geom_text(aes(y = Mean + SE, label = round(Mean, 1)), position = position_dodge(width = 0.8), vjust = -0.5, size = 4.5) +
  facet_wrap(~ Metric, scales = "free_y") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Average per survey by site and year (mean ± SE, points surveyed both years)", x = NULL, y = NULL) +
  theme_publish(base_size = 18) +
  theme(panel.grid.minor = element_blank(), strip.text = element_text(face = "bold"), legend.position = "right")
save_fig(year_bars, "year_comparison_bars.png")
```

## Paired point-level change

Each line is one point (averaged over its two visits in each year). Lines going up mean that point had more birds/species in the later year. The table uses a paired Wilcoxon signed-rank test, which is appropriate for a small number of paired points and doesn't assume normal data. With only 6 points per site, a site-level test can't get below p ≈ 0.03 even if every point changes in the same direction, so focus on the direction and size of the change (mean change, number of points up vs down) more than the p-values.

```{r Paired Change, echo=TRUE, fig.height=10, fig.width=15}
point_year_means <- survey_level_long %>%
  dplyr::group_by(Site, Point_ID, Year, Metric) %>%
  dplyr::summarise(Value = mean(Value), .groups = "drop")

paired_change <- ggplot(point_year_means, aes(x = Year, y = Value, group = Point_ID, color = Site)) +
  geom_line(linewidth = 1, alpha = 0.7) +
  geom_point(size = 3) +
  stat_summary(aes(group = Site), fun = mean, geom = "line", linewidth = 2.5, linetype = "dashed") +
  facet_wrap(~ Metric, scales = "free_y") +
  labs(title = "Change at each point between years (dashed = site mean)",
       x = NULL, y = NULL) +
  theme_publish(base_size = 18) +
  theme(panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold"),
        legend.position = "bottom")
save_fig(paired_change, "paired_point_change_by_year.png")

paired_wide <- point_year_means %>%
  tidyr::pivot_wider(names_from = Year, values_from = Value) %>%
  dplyr::rename(before = dplyr::all_of(year_a), after = dplyr::all_of(year_b))

paired_test <- function(df) {
  df %>%
    dplyr::summarise(
      n_points    = dplyr::n(),
      mean_before = mean(before),
      mean_after  = mean(after),
      mean_change = mean(after - before),
      n_increased = sum(after > before),
      n_decreased = sum(after < before),
      p_value     = suppressWarnings(wilcox.test(after, before, paired = TRUE, exact = FALSE)$p.value),
      .groups = "drop")
}

dplyr::bind_rows(
  paired_wide %>% dplyr::group_by(Metric) %>% paired_test() %>% dplyr::mutate(Site = "Both sites"),
  paired_wide %>% dplyr::group_by(Site, Metric) %>% paired_test()) %>%
  dplyr::select(Site, Metric, dplyr::everything()) %>%
  dplyr::rename(!!paste("mean", year_a) := mean_before, !!paste("mean", year_b) := mean_after) %>%
  kable(digits = c(0, 0, 0, 2, 2, 2, 0, 0, 3),
        caption = paste("Paired change per point,", year_a, "to", year_b, "(paired Wilcoxon signed-rank test)")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed")) %>%
  collapse_rows(columns = 1, valign = "top")
```

## Within-season timing

Each small point is one survey, the large points and lines are the mean for each visit. Because both years are drawn on the same calendar, this shows whether the visit 1 to visit 2 change looked the same each year and whether surveys were done at a similar time of the season (2026 visit 1 was about two weeks earlier than 2025).

```{r Seasonal Timing, echo=TRUE, fig.height=12, fig.width=15}
visit_means <- survey_level_long %>%
  dplyr::group_by(Site, Year, VisitNum, Metric) %>%
  dplyr::summarise(
    Season_Date = mean(Season_Date),
    Mean = mean(Value),
    SE   = sd(Value) / sqrt(dplyr::n()),
    .groups = "drop")

seasonal_timing <- ggplot(survey_level_long, aes(x = Season_Date, y = Value, color = Year)) +
  geom_jitter(width = 0.6, height = 0, alpha = 0.4, size = 2) +
  geom_line(data = visit_means, aes(y = Mean, group = Year), linewidth = 1.2) +
  geom_pointrange(data = visit_means, aes(y = Mean, ymin = Mean - SE, ymax = Mean + SE), size = 0.8) +
  facet_grid(Metric ~ Site, scales = "free_y") +
  scale_x_date(date_labels = "%b %d") +
  labs(title = "Within-season change by year (mean ± SE per visit)",
       x = "Survey date", y = NULL, color = "Year") +
  theme_publish(base_size = 18) +
  theme(panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold"),
        legend.position = "right")
save_fig(seasonal_timing, "seasonal_timing_by_year.png")
```

## Year effect models

GLMs on the survey-level data with Year, Site, their interaction (does the year difference differ between sites?), and VisitNum (early vs late season). Total count uses a quasi-Poisson family because bird counts are usually more variable than a Poisson assumes (flocks). Estimates are exponentiated, so they read as rate ratios: a Year estimate of 0.80 means 20% fewer birds per survey in the later year at the reference site (Cattus Island), holding visit constant.

```{r Year GLM, echo=TRUE}
glm_count_year <- glm(Total_Count ~ Year * Site + VisitNum,
                      data = survey_shared, family = quasipoisson)

glm_richness_year <- glm(Species_Richness ~ Year * Site + VisitNum,
                         data = survey_shared, family = poisson)

kable(broom::tidy(glm_count_year, conf.int = TRUE, exponentiate = TRUE), digits = 3,
      caption = "Total count ~ Year * Site + Visit (quasi-Poisson, rate ratios)") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))

kable(broom::tidy(glm_richness_year, conf.int = TRUE, exponentiate = TRUE), digits = 3,
      caption = "Species richness ~ Year * Site + Visit (Poisson, rate ratios)") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))
```

## Species turnover

Total species richness was similar between years, but that hides a lot of change in *which* species were present. Jaccard similarity is the share of the combined species list that was seen in both years (1 = identical lists, 0 = no species in common). Species seen in only one year are often rare species with one or two detections, so this partly reflects chance detections of rare birds; that ties back to the sampling sufficiency section.

```{r Species Turnover, echo=TRUE, fig.height=6, fig.width=15}
spp_turnover <- spp_presence %>%
  dplyr::group_by(Site) %>%
  dplyr::summarise(
    Shared    = sum(in_a & in_b),
    Lost      = sum(in_a & !in_b),
    Gained    = sum(!in_a & in_b),
    Jaccard   = Shared / (Shared + Lost + Gained),
    !!paste("Only in", year_a) := paste(sort(AlphaCode[in_a & !in_b]), collapse = ", "),
    !!paste("Only in", year_b) := paste(sort(AlphaCode[!in_a & in_b]), collapse = ", "),
    .groups = "drop")

kable(spp_turnover, digits = 2, caption = paste("Species turnover between", year_a, "and", year_b, "(points surveyed both years)")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))

turnover_plot <- spp_turnover %>%
  dplyr::select(Site, Lost, Shared, Gained) %>%
  tidyr::pivot_longer(-Site, names_to = "Status", values_to = "Species") %>%
  dplyr::mutate(Status = factor(Status, levels = c("Lost", "Shared", "Gained"),
                                labels = c(paste("Only in", year_a), "Both years", paste("Only in", year_b)))) %>%
  ggplot(aes(y = Site, x = Species, fill = Status)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = Species), position = position_stack(vjust = 0.5), color = "white", fontface = "bold", size = 5) +
  scale_fill_manual(values = c("#B2182B", "grey55", "#2166AC")) +
  labs(title = "Same richness, different species", x = "Number of species", y = NULL, fill = NULL) +
  theme_publish(base_size = 18) +
  theme(panel.grid.minor = element_blank(), legend.position = "bottom")
save_fig(turnover_plot, "species_turnover.png")
```

## Which species went up or down?

Average birds per survey in each year for every species with at least 10 birds across both years (points surveyed both years). The dot on the left/right shows each year, and the line between them is the change, so the longest lines are the biggest changes. Colour shows the direction. Big swings in flocking species (gulls, terns, swallows) usually reflect one big flock on one survey rather than a real change, so look for consistent changes in resident marsh species.

```{r Species Dumbbell, echo=TRUE, fig.height=14, fig.width=15}
surveys_per_site_year <- survey_shared %>%
  dplyr::count(Site, Year, name = "n_surveys") %>%
  dplyr::mutate(Year = as.character(Year))

spp_change <- shared_long %>%
  dplyr::filter(Identified) %>%
  dplyr::mutate(Year = as.character(Year)) %>%
  dplyr::group_by(Site, Year, Habitat, AlphaCode) %>%
  dplyr::summarise(Total = sum(TotalCount, na.rm = TRUE), .groups = "drop") %>%
  # filling in a 0 for species seen at a site in one year but not the other
  dplyr::group_by(Site) %>%
  tidyr::complete(Year, tidyr::nesting(Habitat, AlphaCode), fill = list(Total = 0)) %>%
  dplyr::ungroup() %>%
  dplyr::left_join(surveys_per_site_year, by = c("Site", "Year")) %>%
  dplyr::mutate(Per_Survey = Total / n_surveys) %>%
  dplyr::group_by(Site, AlphaCode) %>%
  dplyr::filter(sum(Total) >= 10) %>%
  dplyr::mutate(Change = Per_Survey[Year == year_b] - Per_Survey[Year == year_a],
                Direction = ifelse(Change >= 0, paste("Higher in", year_b), paste("Lower in", year_b))) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(Label = reorder(paste(AlphaCode, Site, sep = "___"), Change))

species_dumbbell <- ggplot(spp_change, aes(y = Label, x = Per_Survey)) +
  geom_line(aes(group = Label, color = Direction), linewidth = 2.5, alpha = 0.6) +
  geom_point(aes(shape = Year), size = 3.5, fill = "white") +
  facet_wrap(~ Site, scales = "free_y") +
  scale_y_discrete(labels = function(x) sub("___.*$", "", x)) +
  scale_shape_manual(values = c(21, 19)) +
  scale_color_manual(values = c("#2166AC", "#B2182B")) +
  labs(title = paste("Change in birds per survey,", year_a, "to", year_b),
       subtitle = "Species with ≥ 10 birds across both years, points surveyed both years",
       x = "Average birds per survey", y = NULL, color = NULL, shape = NULL) +
  theme_publish(base_size = 18) +
  theme(panel.grid.minor = element_blank(), strip.text = element_text(face = "bold"), legend.position = "bottom")
save_fig(species_dumbbell, "species_change_dumbbell.png")
```

## Community shift (NMDS)

An ordination places each survey on a map where surveys with similar bird communities sit close together (Bray–Curtis on square-root counts, identified species, shared points). Arrows go from each point's 2025 average position to its 2026 position, so long arrows that all point the same way mean the community shifted between years, while short or random arrows mean it stayed about the same. The PERMANOVA table tests how much of the community difference is explained by site and by year (R² = share of variation; permutations are restricted within points so each point is only compared to itself).

```{r NMDS, echo=TRUE, fig.height=9, fig.width=12}
comm_shared <- make_comm(point_counts_all_clean %>% dplyr::filter(Point_ID %in% shared_points))
comm_shared$matrix <- comm_shared$matrix[, colSums(comm_shared$matrix) > 0]

set.seed(42)
nmds <- vegan::metaMDS(sqrt(comm_shared$matrix), distance = "bray", k = 2, trymax = 100, trace = FALSE)

nmds_scores <- cbind(comm_shared$info, as.data.frame(vegan::scores(nmds, display = "sites")))
nmds_centroids <- nmds_scores %>%
  dplyr::group_by(Site, Point_ID, Year) %>%
  dplyr::summarise(NMDS1 = mean(NMDS1), NMDS2 = mean(NMDS2), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = Year, values_from = c(NMDS1, NMDS2))

nmds_plot <- ggplot(nmds_scores, aes(NMDS1, NMDS2)) +
  stat_ellipse(aes(color = Site, linetype = Year), linewidth = 0.8, level = 0.68) +
  geom_point(aes(color = Site, shape = Year), size = 3, alpha = 0.6) +
  geom_segment(data = nmds_centroids,
               aes(x = .data[[paste0("NMDS1_", year_a)]], y = .data[[paste0("NMDS2_", year_a)]],
                   xend = .data[[paste0("NMDS1_", year_b)]], yend = .data[[paste0("NMDS2_", year_b)]], color = Site),
               arrow = arrow(length = unit(0.25, "cm")), linewidth = 0.9) +
  geom_text(data = nmds_centroids,
            aes(x = .data[[paste0("NMDS1_", year_b)]], y = .data[[paste0("NMDS2_", year_b)]], label = Point_ID),
            size = 3.5, vjust = -0.8, check_overlap = TRUE) +
  coord_equal() +
  labs(title = "Bird community by survey (NMDS)",
       subtitle = sprintf("Arrows: each point's shift from %s to %s. Stress = %.2f (< 0.2 is a usable 2-D picture)", year_a, year_b, nmds$stress)) +
  theme_publish(base_size = 16) +
  theme(panel.grid.minor = element_blank(), legend.position = "right")
save_fig(nmds_plot, "nmds_community_by_year.png")

perm <- permute::how(blocks = comm_shared$info$Point_ID, nperm = 999)
set.seed(42)
permanova_year <- vegan::adonis2(sqrt(comm_shared$matrix) ~ Year + VisitNum,
                                 data = comm_shared$info, method = "bray", permutations = perm)
set.seed(42)
permanova_site <- vegan::adonis2(sqrt(comm_shared$matrix) ~ Site + Year + VisitNum,
                                 data = comm_shared$info, method = "bray", by = "margin", permutations = 999)

kable(as.data.frame(permanova_year), digits = 3,
      caption = "PERMANOVA: year and visit effects (permutations within points)") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))
kable(as.data.frame(permanova_site), digits = 3,
      caption = "PERMANOVA: site, year and visit (marginal effects)") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed"))
```

# More Ways to Look at the Point Counts

## Where each species was found

A grid of every species (rows, grouped by habitat) at every point (columns), coloured by the number of birds counted (both visits combined). This makes it easy to see which species are widespread, which are tied to one site or a few points, and what appeared or disappeared between years.

```{r Species Heatmap, echo=TRUE, fig.height=24, fig.width=16}
heat_data <- point_counts_all %>%
  dplyr::filter(Identified, TotalCount > 0) %>%
  dplyr::group_by(Habitat, AlphaCode, Site, Point_ID, Year) %>%
  dplyr::summarise(Birds = sum(TotalCount), .groups = "drop") %>%
  dplyr::mutate(Year = factor(Year),
                AlphaCode = reorder(AlphaCode, Birds, FUN = sum))

species_heatmap <- ggplot(heat_data, aes(x = Point_ID, y = AlphaCode, fill = Birds)) +
  geom_tile(color = "white", linewidth = 0.3) +
  facet_grid(Habitat ~ Year + Site, scales = "free", space = "free") +
  scale_fill_viridis_c(trans = "log10", option = "mako", direction = -1) +
  labs(title = "Birds counted per species and point", x = NULL, y = NULL, fill = "Birds") +
  theme_publish(base_size = 13) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        strip.text = element_text(face = "bold"),
        strip.text.y = element_text(angle = 0),
        panel.grid = element_blank(),
        legend.position = "right")
save_fig(species_heatmap, "species_by_point_heatmap.png", height = 24, width = 16)
```

## Rank-abundance curves

Species ranked from most to least abundant at each site (log scale). A steep curve means a few species make up most of the birds (low evenness); a flatter, longer curve means a more even and diverse community. The top species are labelled.

```{r Rank Abundance, echo=TRUE, fig.height=8, fig.width=15}
rank_data <- shared_long %>%
  dplyr::filter(Identified, TotalCount > 0) %>%
  dplyr::group_by(Site, Year, AlphaCode) %>%
  dplyr::summarise(Birds = sum(TotalCount), .groups = "drop") %>%
  dplyr::group_by(Site, Year) %>%
  dplyr::arrange(dplyr::desc(Birds), .by_group = TRUE) %>%
  dplyr::mutate(Rank = dplyr::row_number(), Proportion = Birds / sum(Birds)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(Year = factor(Year))

rank_abundance <- ggplot(rank_data, aes(Rank, Proportion, color = Year)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  geom_text(data = dplyr::filter(rank_data, Rank <= 4), aes(label = AlphaCode),
            hjust = -0.25, size = 4, show.legend = FALSE, check_overlap = TRUE) +
  facet_wrap(~ Site) +
  scale_y_log10(labels = scales::label_percent(accuracy = 0.1)) +
  labs(title = "Rank-abundance by site and year (points surveyed both years)",
       x = "Species rank", y = "Share of all birds counted (log scale)") +
  theme_publish(base_size = 18) +
  theme(panel.grid.minor = element_blank(), strip.text = element_text(face = "bold"))
save_fig(rank_abundance, "rank_abundance.png")
```

## Marsh vs upland birds

Share of birds and species that are marsh, upland, or use both, by site and year (all points). A shift toward upland species at a marsh point can be an early sign of habitat change (or just a different survey date).

```{r Habitat Composition, echo=TRUE, fig.height=7, fig.width=15}
habitat_comp <- point_counts_all %>%
  dplyr::filter(TotalCount > 0) %>%
  dplyr::group_by(Site, Year, Habitat) %>%
  dplyr::summarise(Birds = sum(TotalCount),
                   Species = dplyr::n_distinct(AlphaCode[Identified]), .groups = "drop") %>%
  tidyr::pivot_longer(c(Birds, Species), names_to = "Measure", values_to = "Value") %>%
  dplyr::group_by(Site, Year, Measure) %>%
  dplyr::mutate(Share = Value / sum(Value)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(Year = factor(Year),
                Measure = ifelse(Measure == "Birds", "Share of birds", "Share of species"))

habitat_plot <- ggplot(habitat_comp, aes(x = Year, y = Share, fill = Habitat)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = ifelse(Share >= 0.05, scales::percent(Share, accuracy = 1), "")),
            position = position_stack(vjust = 0.5), size = 4.5, color = "white", fontface = "bold") +
  facet_grid(Measure ~ Site) +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = c("Marsh" = "#1B7837", "Upland/Marsh" = "#A6611A", "Upland" = "#DFC27D", "Unclassified" = "grey60")) +
  labs(title = "Marsh vs upland birds by site and year", x = NULL, y = NULL) +
  theme_publish(base_size = 18) +
  theme(panel.grid.minor = element_blank(), strip.text = element_text(face = "bold"))
save_fig(habitat_plot, "habitat_composition_by_year.png")
```

## Detections by distance band

Share of birds detected in each distance band. Most species are harder to detect farther away, so a big share of birds in the outer band usually means flyovers or loud/conspicuous species. If the distance pattern changes a lot between years, differences in counts may partly be differences in detectability (e.g. wind, noise, vegetation height) rather than abundance.

```{r Distance Bands, echo=TRUE, fig.height=7, fig.width=15}
band_comp <- point_counts_all %>%
  dplyr::filter(TotalCount > 0, DistBand != "Unknown") %>%
  dplyr::mutate(Group = ifelse(Habitat == "Marsh", "Marsh species", "Other species")) %>%
  dplyr::group_by(Site, Year, Group, DistBand) %>%
  dplyr::summarise(Birds = sum(TotalCount), .groups = "drop") %>%
  dplyr::group_by(Site, Year, Group) %>%
  dplyr::mutate(Share = Birds / sum(Birds)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(Year = factor(Year), Band = factor(ring_labels[as.character(DistBand)], levels = ring_labels))

band_plot <- ggplot(band_comp, aes(x = Year, y = Share, fill = Band)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = scales::percent(Share, accuracy = 1)), position = position_stack(vjust = 0.5),
            size = 4.5, color = "white", fontface = "bold") +
  facet_grid(Group ~ Site) +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = c("#08519C", "#4292C6", "#9ECAE1")) +
  labs(title = "Share of birds by distance band", x = NULL, y = NULL, fill = "Distance band") +
  theme_publish(base_size = 18) +
  theme(panel.grid.minor = element_blank(), strip.text = element_text(face = "bold"))
save_fig(band_plot, "distance_band_share_by_year.png")
```

# Density and Abundance Models

Density estimation. Also derived from above. Reformatted with ChatGPT to make it easyt to plug and play different spp. 

This code currently models spatial variation in observed abundance using Poisson generalized linear models. Models currently include only geographic coordinates describe spatial gradients in counts but do not represent mechanistic drivers of density. Therefore, abundance estimates derived from these models are interpreted as indices rather than true density estimates. Habitat-based models provide more ecologically meaningful inference.

Other options for the glm estimators should be considered when the data is gathered. Ideas would be vegetation type, habitat, elevation. It would be a good idea to still use spatial smoothing (e.g.TotalCount ~ Habitat + s(Long, Lat) ) or even site effects (TotalCount ~ Habitat + (1 | Point_ID) ).

```{r, echo=TRUE,fig.height=10, fig.width=15}
# Chosen species to be changed in the YAML
species_code <- cfg$params$focal_species

# 1) Filter to one species
data_spp <- point_counts_2025 %>%
  dplyr::filter(AlphaCode == species_code)

# 2) Fit model using predictors. Currently running with lat and long but when I get some veg metrics or other good indicators in there, that would be a much better predictor than location.
m1 <- glm(
  TotalCount ~ Long + Lat,
  data = data_spp,
  family = poisson)

# 3) Tables
tidy_m1 <- broom::tidy(m1, conf.int = TRUE)

stats_m1 <- broom::glance(m1) |>
  dplyr::select(AIC, deviance, df.residual, logLik)

gt::gt(tidy_m1) |>
  gt::tab_header(title = paste0("Poisson GLM Coefficients (", species_code, ")")) |>
  gt::fmt_number(columns = where(is.numeric), decimals = 3)

gt::gt(stats_m1) |>
  gt::tab_header(title = paste0("Model Fit Statistics (", species_code, ")")) |>
  gt::fmt_number(columns = where(is.numeric), decimals = 3)

# 4) Effects + predictions
effects_obj <- effects::allEffects(m1)

plot(
  effects_obj,
  ylab = "Expected count",
  type = "response")

data_spp$predicted_count <- predict(m1, type = "response")

# plotting the predicted density vs the observed density. The bars use one row per point because the prediction is the same for every row at a point (otherwise geom_col stacks them)
ggplot2::ggplot(data_spp, ggplot2::aes(x = Point_ID)) +
  ggplot2::geom_col(ggplot2::aes(y = predicted_count), fill = "grey60",
                    data = data_spp %>% dplyr::distinct(Point_ID, .keep_all = TRUE)) +
  ggplot2::geom_point(
    ggplot2::aes(y = TotalCount, shape = DistBand, color = as.factor(VisitNum)),
    size = 3
  ) +
  ggplot2::labs(
    title = paste0("Observed (points) vs Predicted (bars) Count per Point (", species_code, ")"),
    x = "Point",
    y = "Count"
  ) +
  theme_publish() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))

# 5) Density + scaling summary. Using a fixed point count radius of 200 m2, which the data does include. Decide if I want to restrict to 50 or 100 m for future iterations. Adjust in the YAML. WOuld have to filter by distance bin also (in YAML and then add to code)
data_spp$density_m2 <- data_spp$predicted_count / (pi * cfg$params$point_count_radius^2)

abundance_mean <- mean(data_spp$predicted_count, na.rm = TRUE)
density_mean   <- mean(data_spp$density_m2, na.rm = TRUE)

total_area_m2 <- cfg$params$total_area_m2
total_number  <- density_mean * total_area_m2

data.frame(abundance_mean, density_mean, total_number) |>
  dplyr::rename(
    `abundance (mean among plot)` = abundance_mean,
    `density (per m^2)` = density_mean,
    `total estimated abundance (total sampling area)` = total_number
  ) |>
  gt::gt()
```

2026 density estimation.

```{r, echo=TRUE,fig.height=10, fig.width=15}
species_code <- cfg$params$focal_species

data_spp_2026 <- point_counts_2026 %>%
  dplyr::filter(AlphaCode == species_code)

m1_2026 <- glm(
  TotalCount ~ Long + Lat,
  data = data_spp_2026,
  family = poisson)

gt::gt(broom::tidy(m1_2026, conf.int = TRUE)) |>
  gt::tab_header(title = paste0("Poisson GLM Coefficients 2026 (", species_code, ")")) |>
  gt::fmt_number(columns = where(is.numeric), decimals = 3)

gt::gt(broom::glance(m1_2026) |> dplyr::select(AIC, deviance, df.residual, logLik)) |>
  gt::tab_header(title = paste0("Model Fit Statistics 2026 (", species_code, ")")) |>
  gt::fmt_number(columns = where(is.numeric), decimals = 3)

plot(
  effects::allEffects(m1_2026),
  ylab = "Expected count",
  type = "response")

data_spp_2026$predicted_count <- predict(m1_2026, type = "response")

ggplot2::ggplot(data_spp_2026, ggplot2::aes(x = Point_ID)) +
  ggplot2::geom_col(ggplot2::aes(y = predicted_count), fill = "grey60",
                    data = data_spp_2026 %>% dplyr::distinct(Point_ID, .keep_all = TRUE)) +
  ggplot2::geom_point(
    ggplot2::aes(y = TotalCount, shape = DistBand, color = as.factor(VisitNum)),
    size = 3
  ) +
  ggplot2::labs(
    title = paste0("Observed (points) vs Predicted (bars) Count per Point 2026 (", species_code, ")"),
    x = "Point",
    y = "Count"
  ) +
  theme_publish() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1))

data_spp_2026$density_m2 <- data_spp_2026$predicted_count / (pi * cfg$params$point_count_radius^2)

data.frame(
  abundance_mean = mean(data_spp_2026$predicted_count, na.rm = TRUE),
  density_mean   = mean(data_spp_2026$density_m2, na.rm = TRUE)) |>
  dplyr::mutate(total_number = density_mean * cfg$params$total_area_m2) |>
  dplyr::rename(
    `abundance (mean among plot)` = abundance_mean,
    `density (per m^2)` = density_mean,
    `total estimated abundance (total sampling area)` = total_number
  ) |>
  gt::gt()
```

All years. Same model, with Year added as a factor so the coefficient for Year tells us whether counts of the focal species were higher or lower in 2026 after accounting for location (exponentiate it for a rate ratio, e.g. 1.3 = 30% more birds per detection row). With only two years this is a year-to-year difference, not a trend.

```{r, echo=TRUE,fig.height=10, fig.width=15}
species_code <- cfg$params$focal_species

data_spp_all <- point_counts_all %>%
  dplyr::filter(AlphaCode == species_code) %>%
  dplyr::mutate(Year = factor(Year))

m1_all <- glm(
  TotalCount ~ Long + Lat + Year,
  data = data_spp_all,
  family = poisson)

gt::gt(broom::tidy(m1_all, conf.int = TRUE, exponentiate = TRUE)) |>
  gt::tab_header(title = paste0("Poisson GLM Coefficients, all years (", species_code, ")"),
                 subtitle = "Estimates are exponentiated (rate ratios)") |>
  gt::fmt_number(columns = where(is.numeric), decimals = 3)

gt::gt(broom::glance(m1_all) |> dplyr::select(AIC, deviance, df.residual, logLik)) |>
  gt::tab_header(title = paste0("Model Fit Statistics, all years (", species_code, ")")) |>
  gt::fmt_number(columns = where(is.numeric), decimals = 3)

plot(
  effects::allEffects(m1_all),
  ylab = "Expected count",
  type = "response")

data_spp_all$predicted_count <- predict(m1_all, type = "response")

ggplot2::ggplot(data_spp_all, ggplot2::aes(x = Point_ID)) +
  ggplot2::geom_col(ggplot2::aes(y = predicted_count), fill = "grey60",
                    data = data_spp_all %>% dplyr::distinct(Year, Point_ID, .keep_all = TRUE)) +
  ggplot2::geom_point(
    ggplot2::aes(y = TotalCount, shape = DistBand, color = as.factor(VisitNum)),
    size = 3
  ) +
  ggplot2::facet_wrap(~ Year, ncol = 1) +
  ggplot2::labs(
    title = paste0("Observed (points) vs Predicted (bars) Count per Point by Year (", species_code, ")"),
    x = "Point",
    y = "Count",
    color = "VisitNum"
  ) +
  theme_publish() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
                 strip.text = ggplot2::element_text(face = "bold"))

data_spp_all$density_m2 <- data_spp_all$predicted_count / (pi * cfg$params$point_count_radius^2)

data_spp_all |>
  dplyr::group_by(Year) |>
  dplyr::summarise(
    abundance_mean = mean(predicted_count, na.rm = TRUE),
    density_mean   = mean(density_m2, na.rm = TRUE),
    .groups = "drop") |>
  dplyr::mutate(total_number = density_mean * cfg$params$total_area_m2) |>
  dplyr::rename(
    `abundance (mean among plot)` = abundance_mean,
    `density (per m^2)` = density_mean,
    `total estimated abundance (total sampling area)` = total_number
  ) |>
  gt::gt()
```

Using unmarked. Also derived from above. Reformatted with ChatGPT.

This workflow estimates abundance for a single focal species using an N-mixture model (via the unmarked package). N-mixture models use repeated counts at the same locations to separate two processes:

1. True abundance at a site (how many individuals are actually present), and
2. Detection probability (the chance we detect individuals during a visit, even if they are present).

Because point counts do not detect all birds perfectly, the model treats the observed counts as an imperfect sample of the true abundance.

At present, this code fits a baseline N-mixture model that estimates abundance for a focal species using repeated point counts at each site. The current model assumes constant abundance and constant detectability because covariates are not yet included. In future iterations, the model can be expanded by adding site-level covariates (e.g., LiDAR structure, vegetation composition, hydrology) to explain variation in abundance, and visit-level covariates (e.g., tide, time of day, wind) to explain variation in detection probability.
```{r, echo=TRUE,fig.height=10, fig.width=15}

# 1) Choose focal species column - edited in YAML
species_code <- cfg$params$focal_species

# 2) Build site (Point_ID) x replicate (VisitNum) count table
# This produces ONE row per Point_ID and ONE column per VisitNum
y_df <- point_counts_2025_clean %>%
  dplyr::select(Point_ID, VisitNum, dplyr::all_of(species_code)) %>%
  dplyr::mutate(
    VisitNum = as.character(VisitNum),
    .count = tidyr::replace_na(.data[[species_code]], 0)
  ) %>%
  dplyr::group_by(Point_ID, VisitNum) %>%
  dplyr::summarise(count = sum(.count, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(
    names_from  = VisitNum,
    values_from = count,
    values_fill = 0
  ) %>%
  dplyr::arrange(Point_ID)

# y must be a numeric matrix (no Point_ID column)
y_mat <- as.matrix(dplyr::select(y_df, -Point_ID))

# 3) Build site covariates (currenly only ising point ID, should add in more later like veg, hydro, elevation, etc.) - 
siteCovs <- y_df %>% dplyr::select(Point_ID)

# 4) Fit N-mixture (pcount) model
umf <- unmarked::unmarkedFramePCount(y = y_mat, siteCovs = siteCovs)
fm1 <- unmarked::pcount(~1 ~1, data = umf, K = 50) # assumes no predictors aka detection and abundance are constant and the max number of species is 50. Change in the future to include more covariates like elevation, vegetation cover, etc.

# 5) Predict abundance per site

pred <- predict(fm1, type = "state") #making predictions with the fm1 aka abundance does not vary spatially or by habitat
siteCovs$predicted_abundance <- pred$Predicted

# 6) Density + scaling to total area (200 m radius currently but change in YAML)
plot_area_m2 <- pi * cfg$params$point_count_radius^2

siteCovs$density_m2 <- siteCovs$predicted_abundance / plot_area_m2

predicted_abund <- mean(siteCovs$predicted_abundance, na.rm = TRUE)
density_mean    <- mean(siteCovs$density_m2, na.rm = TRUE)

total_number  <- density_mean * cfg$params$total_area_m2

# 7) Summary table
data.frame(predicted_abund, density_mean, total_number) %>%
  dplyr::rename(
    `abundance (mean among plot)` = predicted_abund,
    `density (per m^2)` = density_mean,
    `total estimated abundance (total sampling area)` = total_number
  ) %>%
  gt::gt()

```

2026 N-mixture model.

```{r, echo=TRUE,fig.height=10, fig.width=15}
species_code <- cfg$params$focal_species

y_df_2026 <- point_counts_2026_clean %>%
  dplyr::select(Point_ID, VisitNum, dplyr::all_of(species_code)) %>%
  dplyr::mutate(
    VisitNum = as.character(VisitNum),
    .count = tidyr::replace_na(.data[[species_code]], 0)
  ) %>%
  dplyr::group_by(Point_ID, VisitNum) %>%
  dplyr::summarise(count = sum(.count, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(
    names_from  = VisitNum,
    values_from = count,
    values_fill = 0
  ) %>%
  dplyr::arrange(Point_ID)

y_mat_2026 <- as.matrix(dplyr::select(y_df_2026, -Point_ID))
siteCovs_2026 <- y_df_2026 %>% dplyr::select(Point_ID)

umf_2026 <- unmarked::unmarkedFramePCount(y = y_mat_2026, siteCovs = siteCovs_2026)
fm1_2026 <- unmarked::pcount(~1 ~1, data = umf_2026, K = 50)

siteCovs_2026$predicted_abundance <- predict(fm1_2026, type = "state")$Predicted
siteCovs_2026$density_m2 <- siteCovs_2026$predicted_abundance / (pi * cfg$params$point_count_radius^2)

data.frame(
  predicted_abund = mean(siteCovs_2026$predicted_abundance, na.rm = TRUE),
  density_mean    = mean(siteCovs_2026$density_m2, na.rm = TRUE)) %>%
  dplyr::mutate(total_number = density_mean * cfg$params$total_area_m2) %>%
  dplyr::rename(
    `abundance (mean among plot)` = predicted_abund,
    `density (per m^2)` = density_mean,
    `total estimated abundance (total sampling area)` = total_number
  ) %>%
  gt::gt()
```

All years, using a "stacked" N-mixture model. Each Point_ID x Year combination is treated as its own site (so 12 points x 2 years + LHC_07 in 2026 = 25 sites), each with its two within-season visits as replicates. Year is added as a site covariate on abundance, so the model estimates abundance separately for each year while sharing one detection probability across years. This is the standard way to handle a short (2–3 year) dataset without a full dynamic (Dail-Madsen) model, which needs more years to estimate. Caveat: the same point in two years is not fully independent, so treat the Year effect as a first look.

```{r, echo=TRUE,fig.height=10, fig.width=15}
species_code <- cfg$params$focal_species

y_df_all <- point_counts_all_clean %>%
  dplyr::select(Point_ID, Year, VisitNum, dplyr::all_of(species_code)) %>%
  dplyr::mutate(
    VisitNum = as.character(VisitNum),
    .count = tidyr::replace_na(.data[[species_code]], 0)
  ) %>%
  dplyr::group_by(Point_ID, Year, VisitNum) %>%
  dplyr::summarise(count = sum(.count, na.rm = TRUE), .groups = "drop") %>%
  tidyr::pivot_wider(
    names_from  = VisitNum,
    values_from = count,
    values_fill = 0
  ) %>%
  dplyr::arrange(Year, Point_ID)

y_mat_all <- as.matrix(dplyr::select(y_df_all, -Point_ID, -Year))
siteCovs_all <- y_df_all %>% dplyr::select(Point_ID, Year)

umf_all <- unmarked::unmarkedFramePCount(y = y_mat_all, siteCovs = as.data.frame(siteCovs_all))
fm_year <- unmarked::pcount(~1 ~Year, data = umf_all, K = 50) # abundance differs by year, detection constant
fm_null <- unmarked::pcount(~1 ~1,    data = umf_all, K = 50) # no year difference

# comparing the two models. If the Year model has the lower AIC (by ~2 or more), there is support for a change in abundance between years
unmarked::modSel(unmarked::fitList(year = fm_year, null = fm_null))

fm_year

siteCovs_all$predicted_abundance <- predict(fm_year, type = "state")$Predicted
siteCovs_all$density_m2 <- siteCovs_all$predicted_abundance / (pi * cfg$params$point_count_radius^2)

siteCovs_all %>%
  dplyr::group_by(Year) %>%
  dplyr::summarise(
    predicted_abund = mean(predicted_abundance, na.rm = TRUE),
    density_mean    = mean(density_m2, na.rm = TRUE),
    .groups = "drop") %>%
  dplyr::mutate(total_number = density_mean * cfg$params$total_area_m2) %>%
  dplyr::rename(
    `abundance (mean among plot)` = predicted_abund,
    `density (per m^2)` = density_mean,
    `total estimated abundance (total sampling area)` = total_number
  ) %>%
  gt::gt()
```

