---
title: "03-BirdNET-Analysis"
author: "Kelly Faller"
date: "2025-06-02"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# BirdNET Analysis Appendix
This appendix contains all analyses and visualizations related to BirdNET—a tool for automated bird sound detection and classification. The BirdNET workflow was separated from the main `02-Acoustics-Analysis.Rmd` file for clarity and reproducibility.

**Why is BirdNET separated?**
- BirdNET is used only occasionally in this project and doesn’t form the core of the daily acoustic analysis.
- Its results and plots are valuable, but they can clutter the main analysis if included directly.
- Having a separate appendix allows for a modular, reproducible workflow—users who don’t need BirdNET analysis can skip this file entirely.

**How to use this appendix:**
1. Open this file (`03-BirdNET-Analysis.Rmd`) in RStudio or an R environment.
2. Adjust any file paths or data-specific parameters as needed.
3. Knit the document to generate a standalone HTML report of BirdNET results. Will be saved in \Outputs\BirdNET

**Important:** This appendix is fully independent of the main `02-Acoustics-Analysis.Rmd` file. If you update your dataset or processing steps, ensure that any BirdNET-specific inputs are updated here as well.

---
# Methods
INCLUDE METHODS HERE!!!!!!!!

## Data Handling

Load Packages

```{r, warning=FALSE, message=FALSE}
# Data manipulation and cleaning
library(tidyverse)
library(dplyr)
library(tidyr)
library(lubridate)
library(tidytext)
library(scales)
#reproducibility
library(yaml)
library(here)
# BirdNET-specific tools
library(birdnetR)
# Visualization
library(ggplot2)
library(ggpubr)
library(viridis)
#sumamry
library(vtable)
library(data.table)
# density and abundance estimations
library(leaflet)
library(vegan)
library(gt)
library(knitr)
library(kableExtra)
library(envalysis)
library(sads)
library(effects)
library(unmarked)
library(glue)
```
Data
```{r YAML data, include=FALSE}
#run YAML 
cfg <- yaml::read_yaml(here("configs", "project.yml"))
cfg$paths   <- lapply(cfg$paths, here)
cfg$outputs <- lapply(cfg$outputs, here)
invisible(lapply(cfg$outputs, dir.create, recursive = TRUE, showWarnings = FALSE))

# Reading CSV files into data frames using YAML 
species_data_birdnet   <- readr::read_csv(cfg$paths$birdnet_results,   show_col_types = FALSE)
species_list_birdnet <- readr::read_csv(cfg$paths$bird_guild,   show_col_types = FALSE)

#connecting species data and species list
species_data_birdnet <- species_data_birdnet %>%
   left_join(dplyr::select(species_list_birdnet, c(Scientific_Name,Lower_Guild, Common_Birds, Guild)), by = 'Scientific_Name')

# formatting
species_data_birdnet$Date <- mdy(species_data_birdnet$Date)
species_data_birdnet$DateTime <- mdy_hm(species_data_birdnet$DateTime)

#make a month and week column
species_data_birdnet <- species_data_birdnet %>%
  mutate(Month = lubridate::month(DateTime, label = TRUE, abbr = TRUE),  Week  = lubridate::week(DateTime))

```

# BirdNet Species Summaries

These data are coming from Cornell's BirdNET Analyzer. It is essentially a GUI version of their popular Merlin app. It requires a lot of decisions on how to analyze your recordings. The user selects confidence levels, sensitiviy, and overlap. The data crunched below is the first deployment (n=10) and is using very lax confidence levels. It will require manual verification (especially for rare species) and the accuracy should be tested. However, we are currently in the stage of seeing how we can best visualize the data. In the near future, we will be testing difference methods to get the most accurate representation of species presence and community composition. 

The summary below counts the amount of unique species identified by BirdNET at each site. These data will be verified to eliminate false detections. The software does a good job of identifying common birds but will occasionally include a few ecologically unreasonable detections. These generally account for very few detections (i.e. <0.01% of all detections), so the numbers appear more inflated in the summaries.
```{r Species Summary, echo=TRUE}

n_distinct(species_data_birdnet$Scientific_Name)
setDT(species_data_birdnet)[, .(count = uniqueN(Scientific_Name)), by = Station]
setDT(species_data_birdnet)[, .(count = uniqueN(Scientific_Name)), by = Site]
guild_count<- setDT(species_data_birdnet)[, .(count = uniqueN(Scientific_Name)), by = Guild]
species_breakdown <- species_data_birdnet %>%
  dplyr::group_by(Station, Scientific_Name) %>%
  dplyr::summarise(Count = dplyr::n(), .groups = "drop") %>%
  dplyr::group_by(Station) %>%
  dplyr::mutate(Percent = round(Count / sum(Count) * 100, 2)) %>%
  dplyr::ungroup()

species_over_1p<- species_breakdown %>% 
  mutate(Percent_1p = case_when(
    Percent > 0.01~ "Above",
    Percent  < 0.01 ~ "Below",
    Percent == 0.01 ~ "At")
    %>% as.factor())

```

# BirdNet Species Outputs

These data are coming from Cornell's BirdNET Analyzer. It is essentially a GUI version of their popular Merlin app. It requires a lot of decisions on how to analyze your recordings. The user selects confidence levels, sensitiviy, and overlap. The data crunched below is the first deployment (n=10) and is using very lax confidence levels. It will require manual verification (especially for rare species) and the accuracy should be tested. However, we are currently in the stage of seeing how we can best visualize the data. In the near future, we will be testing difference methods to get the most accurate representation of species presence and community composition. 

Data is visualized in two ways, common species and guilds. Guilds gives a good overview of how sites vary (i.e. more shorebirds and ducks in a mudflat). Below the graphics will be referencing the same dataset, but either calling out specific common bird species or guild. 

## Species Diversity

*Currently the species graphics are only for the first deployments of 2023. Running BirdNET requires a lot of time and processing power, so we have only completed one round of processing.Will need to update this if it is ever worth the effort*

```{r Species Diversity1, echo=FALSE,fig.height=12, fig.width=18}

#| fig.cap= "Appendix Figure 1: Guild breakdown for all sites."

year_one<- cfg$params$year_one
year_two<- cfg$params$year_two

ggplot(species_data_birdnet %>% filter(Year == year_one,
      !is.na(Month)), aes(x = Month, fill = Lower_Guild)) +
  geom_bar(color = NA) +
  coord_polar(start = 0) +
  theme_bw(base_size = 15) +
  labs( 
    title = paste0("All Station Guild Breakdown (Monthly) ", year_one),
    x = NULL,
    y = "Detections")

ggplot(species_data_birdnet %>% filter(Year == year_two,
      !is.na(Month)), aes(x = Month, fill = Lower_Guild)) +
  geom_bar(color = NA) +
  coord_polar(start = 0) +
  theme_bw(base_size = 15) +
  labs( 
    title = paste0("All Station Guild Breakdown (Monthly) ", year_two),
    x = NULL,
    y = "Detections")

ggplot(species_data_birdnet%>% filter(Year == year_one, !is.na(Week)), aes(x = factor(Week), fill = Lower_Guild)) +
  geom_bar(color = NA) +
  coord_polar(start = 0) +
  theme_bw(base_size = 15) +
  labs(
    title = paste0("All Station Guild Breakdown (Weekly) ", year_one),
    x = NULL,
    y = "Detections")

ggplot(species_data_birdnet%>% filter(Year == year_two, !is.na(Week)), aes(x = factor(Week), fill = Lower_Guild)) +
  geom_bar(color = NA) +
  coord_polar(start = 0) +
  theme_bw(base_size = 15) +
  labs(
    title = paste0("All Station Guild Breakdown (Weekly) ", year_two),
    x = NULL,
    y = "Detections")

ggplot(data= species_data_birdnet%>% filter(Year == year_one), aes(x= DateTime, fill= Lower_Guild), na.rm=T)+ 
         geom_histogram()+ coord_polar()+
         theme_bw(base_size=15)+
         ggtitle("All Station Guild Breakdown ", year_one)+
         labs(x= "DateTime", y= "Guild")

ggplot(data= species_data_birdnet%>% filter(Year == year_two), aes(x= DateTime, fill= Lower_Guild), na.rm=T)+ 
         geom_histogram()+ coord_polar()+
         theme_bw(base_size=15)+
         ggtitle("All Station Guild Breakdown ", year_two)+
         labs(x= "DateTime", y= "Guild")
```

```{r Species Diversity2, echo=FALSE,fig.height=18, fig.width=18}
# remove rows with missing values we care about

x <- species_data_birdnet[!is.na(species_data_birdnet$Station) &
                          !is.na(species_data_birdnet$Common_Name) &
                          !is.na(species_data_birdnet$Year), ]


# keep only years of interest from yaml
x <- x[x$Year %in% c(year_one, year_two), ]

# counts per Station × Common_Name x Year
counts <- as.data.frame(
  table(x$Station, x$Year, x$Common_Name),
  stringsAsFactors = FALSE)

colnames(counts) <- c("Station", "Year", "Common_Name", "Count")

# drop zero counts
counts <- counts[counts$Count > 0, ]

# percent within Station 
counts$Percent <- ave(
  counts$Count,
  counts$Station, counts$Year,
  FUN = function(z) z / sum(z))

# keep only ≥ 1%
species_over_1percent_detections  <- counts[counts$Percent >= 0.01, ]

delta_df <- species_over_1percent_detections %>%
  dplyr::select(Station, Year, Common_Name, Percent) %>%
  tidyr::pivot_wider(
    names_from = Year,
    values_from = Percent,
    values_fill = 0) %>%
  dplyr::mutate(
    Delta = .data[[as.character(year_two)]] - .data[[as.character(year_one)]])

ggplot(delta_df, aes(x = Common_Name, y = Delta)) +
  geom_col() +
  geom_text(
    aes(label = scales::percent(Delta, accuracy = 0.01),
      vjust = ifelse(Delta < 0, 2, -1.5)),
    size = 2.75) +
  facet_grid(~ Station, scales = "free_x") +
  scale_y_continuous(labels = scales::percent) +
  labs(y = paste0("Change in percent of detections (", year_two, " - ", year_one, ")"),
    title = paste0("Change in BirdNET species detections of species with >1% of total detections from ", year_one, " to ", year_two)) +
  theme_bw(base_size = 18) +
  geom_hline(yintercept = 0, color = "black")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
         theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
            annotate("rect", xmin=-Inf, xmax=Inf, ymin=-Inf, ymax=0, alpha=0.15, fill="#F8766D")+
            annotate("rect", xmin=-Inf, xmax=Inf, ymin=0, ymax=Inf, alpha=0.15, fill="#7cae00")


ggplot(species_over_1percent_detections, aes(x = Common_Name, y = Percent, fill = Common_Name)) +
  geom_col() +
  geom_text(
    aes(label = scales::percent(Percent, accuracy = 0.01)),
    vjust = -0.5,
    size = 3) +
  facet_grid(Year ~ Station, scales= "free_x" ) +
  scale_y_continuous(labels = scales::percent) +
  labs(
    y = "Percent",
    fill = "Common Birds",
    title= "Breakdown of BirdNET species detections (>1%)") +
  guides(fill = FALSE) +
  theme_bw(base_size = 18) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1) )

```

```{r Species Diversity3, echo=FALSE,fig.height=12, fig.width=18}

#| #fig.cap= "Appendix Figure 3: Guild breakdown by site during the first deployment (4/26/23 to 5/7/23). Passerines were the most common guild represented in the data due to the high level of SESP detections."

ggplot(species_data_birdnet%>% filter(Year == year_one), aes(x= Lower_Guild,  group= Station)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
    geom_text(aes( label = scales::percent(..prop..,accuracy=0.1),
                   y= ..prop.. ), stat= "count", vjust =- 1.5, size= 2.5) +
    labs(x= "Guild", y = "Percent", fill="Lower_Guild") + 
         ggtitle("Station Guild Breakdown ", year_one)+
  guides(fill= FALSE)+
         theme_bw(base_size=18)+
    facet_grid(~Station) +
    scale_y_continuous(labels = scales::percent)+ 
         theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggplot(species_data_birdnet%>% filter(Year == year_two), aes(x= Lower_Guild,  group= Station)) + 
    geom_bar(aes(y = ..prop.., fill = factor(..x..)), stat="count") +
    geom_text(aes( label = scales::percent(..prop..,accuracy=0.1),
                   y= ..prop.. ), stat= "count", vjust =- 1.5, size= 2.5) +
    labs(x= "Guild", y = "Percent", fill="Lower_Guild") + 
         ggtitle("Station Guild Breakdown ", year_two)+
  guides(fill= FALSE)+
         theme_bw(base_size=18)+
    facet_grid(~Station) +
    scale_y_continuous(labels = scales::percent)+ 
         theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

## Summary Tables
Gives further context to the data, but in a tabular form rather than graphically. 
Broken down by site, station, acoustic indices, site types, and condfidence levels.

```{r Summaries, echo=TRUE}
st(species_data_birdnet %>% dplyr::select(c("Year", "Site", "Station", "Confidence", "Common_Birds")), group= 'Common_Birds', group.long = TRUE)
st(species_breakdown)

```
## Community Ecology Approach
Using vegan package, etc. Add more info here later as this evolves.

```{r Community Matrix, echo=TRUE}

# Build a matrix of species detections to run future code on. This is not filtering any detections. It may be worth including a section above on post-processing for false positives and negatives. Currently, (1/6/26) this code is just the raw results. Change this note when that updates.

#doing one per year
#year one
file_species_year_one <- species_data_birdnet %>%
  filter(!is.na(Scientific_Name), Year == year_one) %>%
  group_by(File_Name, Station, Common_Name) %>%
  dplyr::summarise(
    detections = dplyr::n(),
    .groups = "drop"
  )

comm_abund_year_one <- file_species_year_one %>%
  pivot_wider(names_from = Common_Name,
              values_from = detections,
              values_fill = 0)

comm_matrix_year_one <- comm_abund_year_one %>%
  dplyr::select(-Station, -File_Name)

comm_pa_year_one <- comm_matrix_year_one
comm_pa_year_one[comm_pa_year_one > 0] <- 1 #creates a presence/ absence matrix rather than the number of detections. Avoids calling rate bias.

# Next, as a temporary solution for a lack of post-processing procedure, this code will filter some of the matrix data to get rid of some species that are likely true false positives since they occur in less than 3 files. To be considered "present" in this matrix, you have to be at this site for more than 3 files (which are 15 min recordings).In this case, it got rid of 7 species.

species_site_occ_year_one <- file_species_year_one %>%
  dplyr::distinct(File_Name, Common_Name) %>%     # file-level presence
  dplyr::count(Common_Name, name = "n_files")

valid_species_year_one <- species_site_occ_year_one %>%
  dplyr::filter(n_files >= 3) %>%
  dplyr::pull(Common_Name)

comm_pa_filt_year_one <- comm_pa_year_one[, colnames(comm_pa_year_one) %in% valid_species_year_one, drop = FALSE]

# year two
file_species_year_two <- species_data_birdnet %>%
  filter(!is.na(Scientific_Name), Year == year_two) %>%
  group_by(File_Name, Station, Common_Name) %>%
  dplyr::summarise(
    detections = dplyr::n(),
    .groups = "drop"
  )

comm_abund_year_two <- file_species_year_two %>%
  pivot_wider(names_from = Common_Name,
              values_from = detections,
              values_fill = 0)

comm_matrix_year_two <- comm_abund_year_two %>%
  dplyr::select(-Station, -File_Name)

comm_pa_year_two <- comm_matrix_year_two
comm_pa_year_two[comm_pa_year_two > 0] <- 1 #creates a presence/ absence matrix rather than the number of detections. Avoids calling rate bias.

# Next, as a temporary solution for a lack of post-processing procedure, this code will filter some of the matrix data to get rid of some species that are likely true false positives since they occur in less than 3 files. To be considered "present" in this matrix, you have to be at this site for more than 3 files (which are 15 min recordings).In this case, it got rid of 7 species.


species_site_occ_year_two <- file_species_year_two %>%
  dplyr::distinct(File_Name, Common_Name) %>%     # file-level presence
  dplyr::count(Common_Name, name = "n_files")

valid_species_year_two <- species_site_occ_year_two %>%
  dplyr::filter(n_files >= 3) %>%
  dplyr::pull(Common_Name)

comm_pa_filt_year_two <- comm_pa_year_two[, colnames(comm_pa_year_two) %in% valid_species_year_two, drop = FALSE]

```

Because BirdNET confidence scores are not calibrated probabilities of correct classification and may not reliably reflect species-level accuracy, the results below currently do not have confidence-based filtering. 

This code is derived from the above author and uses the Vegan r package. It generates a community matrix to estimates the total species pool for the study area, accounting for imperfect detection and undersampling.

For more info on it, look at the point count script. 


```{r Species Pool, echo=TRUE}
pool_year_one <- specpool(comm_pa_filt_year_one)

kable(pool_year_one, caption = "Species Pool Estimators") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))

pool_year_two <- specpool(comm_pa_filt_year_two)

kable(pool_year_two, caption = "Species Pool Estimators") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))


```

```{r Species accumulation curve, echo=TRUE,fig.height=10, fig.width=15}
spec_accum_year_one <- specaccum(comm_pa_filt_year_one, method = "random")

plot(spec_accum_year_one, ci.type = "poly", col = "blue", lwd = 2,
     ci.lty = 0, ci.col = "lightblue",
     xlab = "Recordings", ylab = "Species Richness", main= paste("Species Accumulation Curve ", year_one))

spec_accum_year_two <- specaccum(comm_pa_filt_year_two, method = "random")

plot(spec_accum_year_two, ci.type = "poly", col = "blue", lwd = 2,
     ci.lty = 0, ci.col = "lightblue",
     xlab = "Recordings", ylab = "Species Richness", main= paste("Species Accumulation Curve ", year_two))

# now i want to run it on individual species, since this is a community approach. So we get to full richness, but how long should you record for SALS? or others? Also, I want to take into account that for BLRA, a full season of monitoring would be still appropriate
```

Still need to think about and contend with the fact that these are across multiple sites. Should I restrict to one site first? Adapt the code and note when a decision is made.
```{r Species Driven Accumulation Curves, echo=TRUE,fig.height=10, fig.width=15}
# Chosen species to be changed in the YAML
species_code <- cfg$params$focal_species_birdNET

#year one

# Extract presence vector - literally a matrix of just that one species selected above
sp_vec_year_one <- as.integer(comm_pa_filt_year_one[[species_code]])  # 0/1 values per recording
n_samples_year_one <- length(sp_vec_year_one)
stopifnot(n_samples_year_one > 1)  

# Randomize recording order many times
set.seed(123)
n_perm <- 500

accum_mat_year_one <- replicate(n_perm, {
  ord <- sample(sp_vec_year_one, size = n_samples_year_one, replace = FALSE)
  as.integer(cumsum(ord) > 0)   # 0 until first detection, then 1 thereafter
})

#Summarise mean + 95% interval across permutations
acc_df_year_one <- tibble(
  recordings = 1:n_samples_year_one,
  mean = rowMeans(accum_mat_year_one),
  lo = apply(accum_mat_year_one, 1, quantile, probs = 0.025),
  hi = apply(accum_mat_year_one, 1, quantile, probs = 0.975))

ggplot(acc_df_year_one, aes(recordings, mean)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2) +
  geom_step(linewidth = 1) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = paste0(species_code, " accumulation curve ", year_one),
    x = "Number of recordings",
    y = "Cumulative detection probability"
  ) +
  theme_publish(base_size= 15)

# Find first recording where mean detection probability == 1
cutoff_n_year_one <- acc_df_year_one %>%
  dplyr::filter(mean >= 1) %>%
  dplyr::slice(1) %>%
  dplyr::pull(recordings)

if (length(cutoff_n_year_one) == 0) {
  cutoff_n_year_one <- max(acc_df_year_one$recordings)
}

acc_df_trunc_year_one <- acc_df_year_one %>%
  dplyr::filter(recordings <= cutoff_n_year_one)

ggplot(acc_df_trunc_year_one, aes(recordings, mean)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2) +
  geom_step(linewidth = 1) +
  scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0.02))) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.02))) +
  labs(
    title = glue("{species_code} accumulation curve (truncated)  {year_one}"),
    x = "Number of recordings",
    y = "Cumulative detection probability"
  ) +
  theme_publish(base_size = 15)

# year two 

# Extract presence vector - literally a matrix of just that one species selected above
sp_vec_year_two <- as.integer(comm_pa_filt_year_two[[species_code]])  # 0/1 values per recording
n_samples_year_two <- length(sp_vec_year_two)
stopifnot(n_samples_year_two > 1)  

# Randomize recording order many times
set.seed(123)
n_perm <- 500

accum_mat_year_two <- replicate(n_perm, {
  ord <- sample(sp_vec_year_two, size = n_samples_year_two, replace = FALSE)
  as.integer(cumsum(ord) > 0)   # 0 until first detection, then 1 thereafter
})

#Summarise mean + 95% interval across permutations
acc_df_year_two <- tibble(
  recordings = 1:n_samples_year_two,
  mean = rowMeans(accum_mat_year_two),
  lo = apply(accum_mat_year_two, 1, quantile, probs = 0.025),
  hi = apply(accum_mat_year_two, 1, quantile, probs = 0.975))

ggplot(acc_df_year_two, aes(recordings, mean)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2) +
  geom_step(linewidth = 1) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = paste0(species_code, " accumulation curve ", year_two),
    x = "Number of recordings",
    y = "Cumulative detection probability"
  ) +
  theme_publish(base_size= 15)

# Find first recording where mean detection probability == 1
cutoff_n_year_two <- acc_df_year_two %>%
  dplyr::filter(mean >= 1) %>%
  dplyr::slice(1) %>%
  dplyr::pull(recordings)

if (length(cutoff_n_year_two) == 0) {
  cutoff_n_year_two <- max(acc_df_year_two$recordings)
}

acc_df_trunc_year_two <- acc_df_year_two %>%
  dplyr::filter(recordings <= cutoff_n_year_two)

ggplot(acc_df_trunc_year_two, aes(recordings, mean)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2) +
  geom_step(linewidth = 1) +
  scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0, 0.02))) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.02))) +
  labs(
    title = glue("{species_code} accumulation curve (truncated) {year_two}"),
    x = "Number of recordings",
    y = "Cumulative detection probability"
  ) +
  theme_publish(base_size = 15)

```


<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>