---
title: "Arbimon Data Prep Script"
author: "Kelly Faller"
date: "2024-10-11"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages

```{r, warning=FALSE, include=FALSE}

library(tidyverse)
library(lubridate)
library(bit64)
library(magrittr)

```

## Create one CSV

When dowloading the backup off of Arbimon, if you have a large number of recordings, there will be several CSV files that have a record of those recordings. This chunk will takethe folder full of those CSV files and creating one large CSV.

Changing the working directory back to my personal wd that connects with GitHub

```{r CSV Binding, include=FALSE}

setwd('C:/Users/kfaller/Documents/GitHub/Acoustics/Arbimon Data/Recording Files')

# bind all CSV files
df<-list.files(path='C:/Users/kfaller/Documents/GitHub/Acoustics/Arbimon Data/Recording Files')%>% 
    lapply(read_csv) %>% 
    bind_rows 

# Load the metadata dataset (this should be an output of Arbimon backup. Double check that the name of the CSV is the one that it created for you)
metadata <- read.csv("C:/Users/kfaller/Documents/GitHub/Acoustics/Arbimon Data/sites.0001.csv")

# merge the metadata into the recording CSV so that you have the site name, not just site number (which is arbitrarily made up by Arbimon)

recording_data<- merge(df, metadata, by = "site_id")

```

## Lubridate

Working with datetime and creating a new column for Julian day. With Julian day, we can track the time each ARU was recording and create recording schedule graphics.

```{r CSV Binding, include=FALSE}

# Convert the datetime column to POSIXct format (adjust the format to match your data)
recording_data$datetime <- as.POSIXct(recording_data$datetime, format = "%m/%d/%Y %H:%M:%S")

# Add a new column with Julian day using lubridate's yday() function so we can make recording schedule graphics
recording_data$Julian_Day <- yday(recording_data$datetime)

# Add a new column with the year using lubridate's year() function so that we can filter by year
recording_data$Year <- year(recording_data$datetime)

```

## Final Output

Creating a final CSV :)

```{r Final, include=FALSE}

write.csv(recording_data,"C:/Users/kfaller/Documents/GitHub/Acoustics/Recording_Metadata.csv", row.names = FALSE)

```