---
title: "Data Prep Script"
author: "Kelly Faller"
date: '2023-05-20'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages


```{r, warning=FALSE, include=FALSE}

library(tidyverse)
library(lubridate)
library(bit64)
library(magrittr)


```
## Create one CSV

Taking the folder full of individual CSV files for each logger and creating one master CSV

Changing the working directory back to my personal wd that connects with GitHub

```{r CSV Binding, include=FALSE}

setwd('C:/Users/kfaller/Documents/GitHub/Acoustics/Acoustic Index Files')

df<-list.files(path='C:/Users/kfaller/Documents/GitHub/Acoustics/Acoustic Index Files')%>% 
    lapply(read_csv) %>% 
    bind_rows 

setwd('C:/Users/kfaller/Documents/GitHub/Acoustics')

write.csv(df,"ARU_Data_Indicies.csv", row.names = FALSE)  

aru_data1<- read.csv("ARU_Data_Indicies.csv", header=TRUE, sep = ",", stringsAsFactors=FALSE)

```

## Lubridate

Work with dates, times, datetimes.The column title "TIME" is the minute and second.
The column title "Hour" is ...hour. I want this in a format that reads as one time. 


```{r Lubridate, echo=FALSE}

aru_data2 <- aru_data1 %>% 
    rename(MINUTESECOND = TIME)%>% 
    unite(TIME, HOUR, MINUTESECOND, sep = ":") 

aru_data3 <- aru_data2 %>%
   dplyr::mutate(TIME2 = substr(TIME, 1, nchar(TIME)-3))%>% 
   unite(DATETIME, DATE, TIME2, sep = " ") 

aru_data3$DATETIME <- mdy_hms(aru_data3$DATETIME)

aru_data4 <- aru_data3 %>%  mutate(DATE = as.Date(DATETIME)) %>%
         mutate(DATE = format(DATE, "%m/%d/%Y"))

aru_data4$DATE <- mdy(aru_data4$DATE)

write.csv(aru_data4,"ARU_Data_Indicies.csv", row.names = FALSE)     

```


